#========================================================
# GSEA analysis for invaded vs uninvaded nerve DE results
#========================================================
#
# Purpose:
# This script performs gene set enrichment analysis (GSEA) on a ranked
# differential expression table comparing invaded versus uninvaded nerve
# profiles.
#
# Workflow:
#   1. Read the differential expression table.
#   2. Build ranked gene lists using:
#        - ENTREZ IDs for ontology/pathway analyses
#        - gene SYMBOLs for custom GMT-based analyses
#   3. Run multiple GSEA analyses in a reproducible framework.
#   4. Save:
#        - CSV result tables
#        - PDF dotplots
#        - RDS objects
#        - QC outputs and session information
#
# Expected input columns:
#   - Gene
#   - logFC
#   - FDR
#
# Notes:
#   - The ranking formula is:
#         sign(logFC) * -log10(FDR)
#   - A single seed is used throughout for reproducibility.
#   - nPermSimple can be adjusted centrally in the parameter section.
#   - For ENTREZ-based analyses, exported result tables are made more
#     interpretable using setReadable().
#
#========================================================
```{r}
library(renv)
renv::activate() 
```

```{r}
#========================================================
# 1. Load packages
#========================================================
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(tibble)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(ReactomePA)
  library(DOSE)
  library(msigdbr)
  library(enrichplot)
  library(ggplot2)
  library(fgsea)
})
```


```{r}
#========================================================
# 2. User-defined parameters
#========================================================

#---- Reproducibility ----
seed_value <- 1234

#---- GSEA parameters ----
n_perm          <- 100000   # adjust here if needed
min_gs_size     <- 10
max_gs_size     <- 800
p_cutoff        <- 0.05
adjust_method   <- "BH"
show_categories <- 30

#---- Project directories ----
project_dir <- "~/NAS/Aleksandra/Computational/R programming/paper_projects/"
input_dir   <- file.path(project_dir, "input lists")
gmt_dir     <- file.path(project_dir, "single_nuclei_PDAC")
output_dir  <- file.path(project_dir, "output enrichment analysis")

#---- Input files ----
de_file           <- file.path(input_dir, "FullDE_List.csv")
combined_gmt_file <- file.path(gmt_dir, "combined_signatures.gmt")

#---- Output folders ----
csv_dir <- file.path(output_dir, "csv", "GSEA")
pdf_dir <- file.path(output_dir, "pdf", "GSEA")
rds_dir <- file.path(output_dir, "rds", "GSEA")
qc_dir  <- file.path(output_dir, "qc",  "GSEA")
log_dir <- file.path(output_dir, "logs")

dir.create(csv_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(pdf_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(rds_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(qc_dir,  recursive = TRUE, showWarnings = FALSE)
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

#---- Log file ----
log_file <- file.path(log_dir, paste0("GSEA_run_", format(Sys.time(), "%Y%m%d_%H%M%S"), ".log"))

log_message <- function(...) {
  msg <- paste0("[", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "] ", paste(..., collapse = ""))
  cat(msg, "\n")
  write(msg, file = log_file, append = TRUE)
}

set.seed(seed_value)
log_message("Starting GSEA pipeline.")
log_message("Seed set to: ", seed_value)
log_message("nPermSimple set to: ", n_perm)
```

```{r}
#========================================================
# 3. Helper functions
#========================================================

#--------------------------------------------------------
# 3.1 Check that required input columns are present
#--------------------------------------------------------
check_required_columns <- function(df, required_cols = c("Gene", "logFC", "FDR")) {
  missing_cols <- setdiff(required_cols, colnames(df))
  if (length(missing_cols) > 0) {
    stop(
      "Input file is missing required columns: ",
      paste(missing_cols, collapse = ", ")
    )
  }
}

#--------------------------------------------------------
# 3.2 Build ranked gene vector
#--------------------------------------------------------
make_rank_vector <- function(df, gene_col, logfc_col = "logFC", fdr_col = "FDR") {
  df_clean <- df %>%
    dplyr::filter(!is.na(.data[[gene_col]]),
                  !is.na(.data[[logfc_col]]),
                  !is.na(.data[[fdr_col]])) %>%
    dplyr::filter(.data[[fdr_col]] > 0) %>%
    dplyr::mutate(rank_score = sign(.data[[logfc_col]]) * -log10(.data[[fdr_col]])) %>%
    dplyr::distinct(.data[[gene_col]], .keep_all = TRUE)

  ranks <- df_clean$rank_score
  names(ranks) <- df_clean[[gene_col]]
  ranks <- sort(ranks, decreasing = TRUE)

  return(ranks)
}

#--------------------------------------------------------
# 3.3 Save ranked list as CSV
#--------------------------------------------------------
save_rank_table <- function(ranks, gene_id_name, outfile) {
  rank_df <- tibble(
    !!gene_id_name := names(ranks),
    rank_score = as.numeric(ranks)
  )
  readr::write_csv(rank_df, outfile)
}

#--------------------------------------------------------
# 3.4 Save ranked list barplot
#--------------------------------------------------------
save_rank_barplot <- function(ranks, file, main_title) {
  pdf(file, width = 12, height = 6, useDingbats = FALSE)
  barplot(
    ranks,
    border = NA,
    las = 2,
    main = main_title,
    cex.names = 0.4
  )
  dev.off()
}

#--------------------------------------------------------
# 3.5 Export GSEA result table
#--------------------------------------------------------
save_gsea_table <- function(gsea_obj, analysis_name, csv_dir, rds_dir) {
  res <- gsea_obj@result %>% as.data.frame()

  if (nrow(res) == 0) {
    log_message("No enriched terms found for ", analysis_name)
  }

  if (!"Description" %in% colnames(res)) {
    res$Description <- NA_character_
  }

  if (!"geneID" %in% colnames(res)) {
    res$geneID <- NA_character_
  }

  # Make exported table easier to inspect
  preferred_order <- c(
    "ID", "Description", "setSize", "enrichmentScore", "NES",
    "pvalue", "p.adjust", "qvalues", "rank", "leading_edge", "core_enrichment", "geneID"
  )
  existing_order <- preferred_order[preferred_order %in% colnames(res)]
  remaining_cols <- setdiff(colnames(res), existing_order)
  res <- res[, c(existing_order, remaining_cols), drop = FALSE]

  readr::write_csv(res, file.path(csv_dir, paste0(analysis_name, "_results.csv")))
  saveRDS(gsea_obj, file = file.path(rds_dir, paste0(analysis_name, ".rds")))

  return(res)
}

#--------------------------------------------------------
# 3.6 Save GSEA dotplot
#--------------------------------------------------------
save_gsea_dotplot <- function(gsea_obj, analysis_name, pdf_dir, show_categories = 30) {
  res <- gsea_obj@result
  pdf_file <- file.path(pdf_dir, paste0(analysis_name, "_dotplot.pdf"))

  pdf(pdf_file, width = 10, height = 8, useDingbats = FALSE)

  if (nrow(res) > 0) {
    p <- dotplot(gsea_obj, x = "NES", showCategory = show_categories) +
      ggtitle(analysis_name) +
      theme_bw(base_size = 12)
    print(p)
  } else {
    plot.new()
    title(main = paste("No enriched terms for", analysis_name))
    text(0.5, 0.5, "No terms passed the enrichment threshold.")
  }

  dev.off()
}

#--------------------------------------------------------
# 3.7 Add readable gene symbols to exported results
#--------------------------------------------------------
# This improves readability of saved CSV files. For many clusterProfiler
# outputs, geneID contains slash-separated ENTREZ IDs. This function maps
# them back to gene symbols when possible.
add_readable_gene_symbols <- function(result_df) {
  if (!"geneID" %in% colnames(result_df)) {
    result_df$gene_symbols <- NA_character_
    return(result_df)
  }

  unique_ids <- unique(unlist(strsplit(result_df$geneID, "/")))
  unique_ids <- unique_ids[!is.na(unique_ids) & unique_ids != ""]

  if (length(unique_ids) == 0) {
    result_df$gene_symbols <- NA_character_
    return(result_df)
  }

  suppressWarnings({
    id_map <- bitr(
      unique_ids,
      fromType = "ENTREZID",
      toType   = "SYMBOL",
      OrgDb    = org.Hs.eg.db
    )
  })

  if (is.null(id_map) || nrow(id_map) == 0) {
    result_df$gene_symbols <- NA_character_
    return(result_df)
  }

  id_lookup <- stats::setNames(id_map$SYMBOL, id_map$ENTREZID)

  result_df$gene_symbols <- vapply(
    strsplit(result_df$geneID, "/"),
    function(id_vec) {
      syms <- id_lookup[id_vec]
      syms <- syms[!is.na(syms)]
      paste(unique(syms), collapse = "/")
    },
    character(1)
  )

  return(result_df)
}

#--------------------------------------------------------
# 3.8 Save final GSEA outputs consistently
#--------------------------------------------------------
save_all_outputs <- function(gsea_obj, analysis_name, csv_dir, pdf_dir, rds_dir, show_categories = 30) {
  res <- save_gsea_table(gsea_obj, analysis_name, csv_dir, rds_dir)

  if ("geneID" %in% colnames(res)) {
    res <- add_readable_gene_symbols(res)
    readr::write_csv(res, file.path(csv_dir, paste0(analysis_name, "_results.csv")))
  }

  save_gsea_dotplot(gsea_obj, analysis_name, pdf_dir, show_categories = show_categories)

  invisible(res)
}

#--------------------------------------------------------
# 3.9 Run a standard ENTREZ-based GSEA analysis
#--------------------------------------------------------
run_gsea_entrez <- function(fun,
                            analysis_name,
                            gene_list,
                            make_readable = TRUE,
                            ...) {
  log_message("Running analysis: ", analysis_name)

  set.seed(seed_value)

  gsea_obj <- fun(
    geneList      = gene_list,
    nPermSimple   = n_perm,
    minGSSize     = min_gs_size,
    maxGSSize     = max_gs_size,
    pvalueCutoff  = p_cutoff,
    pAdjustMethod = adjust_method,
    verbose       = TRUE,
    eps           = 0,
    seed          = TRUE,
    by            = "fgsea",
    ...
  )

  if (make_readable) {
    gsea_obj <- setReadable(gsea_obj, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
  }

  save_all_outputs(
    gsea_obj = gsea_obj,
    analysis_name = analysis_name,
    csv_dir = csv_dir,
    pdf_dir = pdf_dir,
    rds_dir = rds_dir,
    show_categories = show_categories
  )

  return(gsea_obj)
}

#--------------------------------------------------------
# 3.10 Run generic TERM2GENE GSEA
#--------------------------------------------------------
run_gsea_term2gene <- function(analysis_name,
                               gene_list,
                               term2gene,
                               gene_id_type = c("ENTREZID", "SYMBOL"),
                               make_readable = TRUE) {
  gene_id_type <- match.arg(gene_id_type)

  log_message("Running analysis: ", analysis_name)

  set.seed(seed_value)

  gsea_obj <- GSEA(
    geneList      = gene_list,
    TERM2GENE     = term2gene,
    nPermSimple   = n_perm,
    minGSSize     = min_gs_size,
    maxGSSize     = max_gs_size,
    pvalueCutoff  = p_cutoff,
    pAdjustMethod = adjust_method,
    eps           = 0,
    seed          = TRUE,
    verbose       = TRUE,
    by            = "fgsea"
  )

  if (make_readable && gene_id_type == "ENTREZID") {
    gsea_obj <- setReadable(gsea_obj, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
  }

  save_all_outputs(
    gsea_obj = gsea_obj,
    analysis_name = analysis_name,
    csv_dir = csv_dir,
    pdf_dir = pdf_dir,
    rds_dir = rds_dir,
    show_categories = show_categories
  )

  return(gsea_obj)
}
```


```{r}
#========================================================
# 4. Read input data
#========================================================
log_message("Reading input DE table: ", de_file)

background <- readr::read_csv(de_file, show_col_types = FALSE)
check_required_columns(background)

log_message("Input table loaded successfully.")
log_message("Number of rows in DE table: ", nrow(background))
```

```{r}
#========================================================
# 5. Prepare ranked gene lists
#========================================================

#--------------------------------------------------------
# 5.1 Map gene symbols to ENTREZ IDs
#--------------------------------------------------------
log_message("Converting gene symbols to ENTREZ IDs.")

background_entrez <- bitr(
  background$Gene,
  fromType = "SYMBOL",
  toType   = "ENTREZID",
  OrgDb    = org.Hs.eg.db
)

background_with_id <- background %>%
  dplyr::left_join(background_entrez, by = c("Gene" = "SYMBOL")) %>%
  dplyr::distinct(Gene, .keep_all = TRUE) %>%
  dplyr::filter(!is.na(ENTREZID))

log_message("Genes with valid ENTREZ IDs: ", nrow(background_with_id))

#--------------------------------------------------------
# 5.2 Build ranked lists
#--------------------------------------------------------
ranks_entrez <- make_rank_vector(
  df = background_with_id,
  gene_col = "ENTREZID"
)

ranks_symbol <- make_rank_vector(
  df = background,
  gene_col = "Gene"
)

log_message("ENTREZ ranked list length: ", length(ranks_entrez))
log_message("SYMBOL ranked list length: ", length(ranks_symbol))

#--------------------------------------------------------
# 5.3 Save ranked lists and QC plots
#--------------------------------------------------------
save_rank_table(
  ranks = ranks_entrez,
  gene_id_name = "ENTREZID",
  outfile = file.path(qc_dir, "ranked_genes_entrez.csv")
)

save_rank_table(
  ranks = ranks_symbol,
  gene_id_name = "Gene",
  outfile = file.path(qc_dir, "ranked_genes_symbol.csv")
)

save_rank_barplot(
  ranks = ranks_entrez,
  file = file.path(qc_dir, "ranked_genes_entrez_barplot.pdf"),
  main_title = "Ranked genes (ENTREZ)"
)

save_rank_barplot(
  ranks = ranks_symbol,
  file = file.path(qc_dir, "ranked_genes_symbol_barplot.pdf"),
  main_title = "Ranked genes (SYMBOL)"
)

log_message("Ranked gene lists and QC plots saved.")
```

```{r}
#========================================================
# 6. GO and pathway enrichment analyses
#========================================================

#--------------------------------------------------------
# 6.1 Gene Ontology - Cellular Component
#--------------------------------------------------------
gse_cc <- run_gsea_entrez(
  fun = gseGO,
  analysis_name = "gse_go_cc",
  gene_list = ranks_entrez,
  make_readable = TRUE,
  ont = "CC",
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID"
)

#--------------------------------------------------------
# 6.2 Gene Ontology - Biological Process
#--------------------------------------------------------
gse_bp <- run_gsea_entrez(
  fun = gseGO,
  analysis_name = "gse_go_bp",
  gene_list = ranks_entrez,
  make_readable = TRUE,
  ont = "BP",
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID"
)

#--------------------------------------------------------
# 6.3 Gene Ontology - Molecular Function
#--------------------------------------------------------
gse_mf <- run_gsea_entrez(
  fun = gseGO,
  analysis_name = "gse_go_mf",
  gene_list = ranks_entrez,
  make_readable = TRUE,
  ont = "MF",
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID"
)

#--------------------------------------------------------
# 6.4 Reactome pathway analysis
#--------------------------------------------------------
gse_reactome <- run_gsea_entrez(
  fun = gsePathway,
  analysis_name = "gse_reactome",
  gene_list = ranks_entrez,
  make_readable = TRUE,
  organism = "human"
)

#--------------------------------------------------------
# 6.5 WikiPathways analysis
#--------------------------------------------------------
gse_wikipathways <- run_gsea_entrez(
  fun = gseWP,
  analysis_name = "gse_wikipathways",
  gene_list = ranks_entrez,
  make_readable = TRUE,
  organism = "Homo sapiens"
)

#--------------------------------------------------------
# 6.6 Disease Ontology analysis
#--------------------------------------------------------
gse_do <- run_gsea_entrez(
  fun = gseDO,
  analysis_name = "gse_disease_ontology",
  gene_list = ranks_entrez,
  make_readable = TRUE,
  organism = "hsa"
)

#--------------------------------------------------------
# 6.7 Network of Cancer Genes analysis
#--------------------------------------------------------
gse_ncg <- run_gsea_entrez(
  fun = gseNCG,
  analysis_name = "gse_ncg",
  gene_list = ranks_entrez,
  make_readable = TRUE,
  exponent = 1
)

#--------------------------------------------------------
# 6.8 KEGG pathway analysis
#--------------------------------------------------------
gse_kegg <- run_gsea_entrez(
  fun = gseKEGG,
  analysis_name = "gse_kegg",
  gene_list = ranks_entrez,
  make_readable = TRUE
)
```

```{r}
#========================================================
# 7. Custom GMT-based enrichment analysis
#========================================================

#--------------------------------------------------------
# 7.1 Load custom cell-state GMT signatures
#--------------------------------------------------------
log_message("Reading custom GMT file: ", combined_gmt_file)

m_t2g_cells_combined <- read.gmt(combined_gmt_file)

#--------------------------------------------------------
# 7.2 Run SYMBOL-based custom GSEA
#--------------------------------------------------------
gse_cells_combined <- run_gsea_term2gene(
  analysis_name = "gse_cells_combined",
  gene_list = ranks_symbol,
  term2gene = m_t2g_cells_combined,
  gene_id_type = "SYMBOL",
  make_readable = FALSE
)


#========================================================
# 8. MSigDB collection analyses
#========================================================

run_msigdb_collection <- function(category_name) {
  log_message("Preparing MSigDB collection: ", category_name)

  term2gene <- msigdbr(
    species = "Homo sapiens",
    category = category_name
  ) %>%
    dplyr::select(gs_name, entrez_gene)

  run_gsea_term2gene(
    analysis_name = paste0("gse_msigdb_", category_name),
    gene_list = ranks_entrez,
    term2gene = term2gene,
    gene_id_type = "ENTREZID",
    make_readable = TRUE
  )
}

#--------------------------------------------------------
# 8.1 Hallmark
#--------------------------------------------------------
gse_msigdb_h <- run_msigdb_collection("H")

#--------------------------------------------------------
# 8.2 Positional gene sets
#--------------------------------------------------------
gse_msigdb_c1 <- run_msigdb_collection("C1")

#--------------------------------------------------------
# 8.3 Curated gene sets
#--------------------------------------------------------
gse_msigdb_c2 <- run_msigdb_collection("C2")

#--------------------------------------------------------
# 8.4 Regulatory target gene sets
#--------------------------------------------------------
gse_msigdb_c3 <- run_msigdb_collection("C3")

#--------------------------------------------------------
# 8.5 Computational gene sets
#--------------------------------------------------------
gse_msigdb_c4 <- run_msigdb_collection("C4")

#--------------------------------------------------------
# 8.6 GO gene sets
#--------------------------------------------------------
gse_msigdb_c5 <- run_msigdb_collection("C5")

#--------------------------------------------------------
# 8.7 Oncogenic signatures
#--------------------------------------------------------
gse_msigdb_c6 <- run_msigdb_collection("C6")

#--------------------------------------------------------
# 8.8 Immunologic signatures
#--------------------------------------------------------
gse_msigdb_c7 <- run_msigdb_collection("C7")

#--------------------------------------------------------
# 8.9 Cell type signatures
#--------------------------------------------------------
gse_msigdb_c8 <- run_msigdb_collection("C8")
```

```{r}
#========================================================
# 9. Save reproducibility information
#========================================================

#--------------------------------------------------------
# 9.1 Save session info
#--------------------------------------------------------
writeLines(
  capture.output(sessionInfo()),
  con = file.path(qc_dir, "sessionInfo.txt")
)

#--------------------------------------------------------
# 9.2 Save run parameters
#--------------------------------------------------------
params_df <- tibble(
  parameter = c(
    "seed_value",
    "n_perm",
    "min_gs_size",
    "max_gs_size",
    "p_cutoff",
    "adjust_method",
    "show_categories",
    "de_file",
    "combined_gmt_file"
  ),
  value = c(
    as.character(seed_value),
    as.character(n_perm),
    as.character(min_gs_size),
    as.character(max_gs_size),
    as.character(p_cutoff),
    adjust_method,
    as.character(show_categories),
    de_file,
    combined_gmt_file
  )
)

readr::write_csv(params_df, file.path(qc_dir, "run_parameters.csv"))
```

```{r}
#========================================================
# 10. Completion message
#========================================================
log_message("GSEA pipeline completed successfully.")
log_message("CSV output directory: ", csv_dir)
log_message("PDF output directory: ", pdf_dir)
log_message("RDS output directory: ", rds_dir)
log_message("QC output directory: ", qc_dir)
```
