#========================================
# Pathway activity profiling across nerve categories
#========================================
#
# This script compares pathway-level transcriptional activity across three nerve
# categories: NonInvaded_Adjacent, NonInvaded_Diseased, and Invaded nerves.
#
# Main goal:
# To identify whether invaded nerves show distinct inflammatory or stromal
# pathway activity compared with non-invaded nerves from adjacent or diseased tissue.


```{r}
#========================================
# 2. Install required CRAN and Bioconductor packages
#========================================
# Install the plotting, data manipulation, spatial transcriptomics, and gene-set
# analysis packages needed for the workflow.

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(GSVA)
  library(matrixStats)
  library(msigdbr)
  library(Matrix)
  library(SpatialExperiment)
  library(readr)
  library(purrr)
  library(tibble)
})

c25 <- c(
  "dodgerblue2", "#E31A1C", # red
"green4",
"#6A3D9A", # purple
"#FF7F00", # orange
"black", "gold1",
"skyblue2", "#FB9A99", # lt pink
"palegreen2",
"#CAB2D6", # lt purple
"#FDBF6F", # lt orange
"gray70", "khaki2",
"maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
"darkturquoise", "green1", "yellow4", "yellow3",
"darkorange4", "brown"
)
```

```{r}
spe <- readRDS("/media/Konstantina/12_Olas_project/Publication_files/data/SPE/Integrated_NMF.rds")
colData(spe) <- DataFrame(read.csv("/media/Konstantina/12_Olas_project/Publication_files/data/coldata/DE_SPE_annotaion.csv"))
colnames(spe) <- spe$barcode
```

```{r}

#========================================
# 4. Subset nerve spots and define the three nerve categories
#========================================
# Restrict the dataset to nerve spots and assign each nerve to Adjacent,
# Diseased, or Invaded categories for downstream comparisons.

nerve_spe <- spe[, colData(spe)$is.nerve == TRUE]

nerve_spe$types <- nerve_spe$type
colData(nerve_spe)$types[!colData(nerve_spe)$types == "noPNI"] <- "Diseased"
colData(nerve_spe)$types[colData(nerve_spe)$types == "noPNI"] <- "Adjacent"

colData(nerve_spe)$nerve_types <- nerve_spe$Invaded
colData(nerve_spe)$nerve_types <- ifelse(
  colData(nerve_spe)$Invaded == "NonInvaded",
  paste0(colData(nerve_spe)$Invaded, "_", colData(nerve_spe)$types),
  as.character(colData(nerve_spe)$nerve_types)
)

nerve_spe <- nerve_spe[, !is.na(colData(nerve_spe)$nerve_types)]

# enforce nerve type order
nerve_type_order <- c("NonInvaded_Adjacent", "NonInvaded_Diseased", "Invaded")
```

```{r}
#========================================
# 5. Prepare expression and metadata objects for scoring
#========================================
# Extract the log-normalized expression matrix and organize metadata such as
# nerve type and patient ID for pathway analysis.

meta <- as.data.frame(colData(nerve_spe))
meta$nerve_types <- factor(as.character(meta$nerve_types), levels = nerve_type_order)

# patient order (EDIT)
patient_order <- c("54703", "6249", "70865")
meta$patient_id <- factor(as.character(meta$patient_id), levels = patient_order)

# expression (genes x spots)
assay_name <- "logcounts"
expr <- assay(nerve_spe, assay_name)
expr <- as.matrix(expr)  # ok for nerve subset; keep as matrix for GSVA

# ensure rownames are gene SYMBOLs (required for msigdbr sets)
stopifnot(!is.null(rownames(expr)))
```

```{r} 
#========================================
# 6. Define helper functions for MSigDB gene-set retrieval
#========================================
# Create utility functions to retrieve exact or pattern-matched MSigDB gene sets
# and support flexible pathway selection.

`%||%` <- function(a, b) if (!is.null(a)) a else b

# Pull an EXACT msigdbr gene set by gs_name (FIXED)
msig_exact <- function(gs_name_in, species = "Homo sapiens") {
  tb <- msigdbr(species = species) %>% dplyr::filter(.data$gs_name == gs_name_in)
  if (nrow(tb) == 0) stop("msigdbr: exact gs_name not found: ", gs_name_in)
  unique(tb$gene_symbol)
}

# Pull by PATTERN; if no match, show suggestions
msig_pick <- function(pattern,
                      species = "Homo sapiens",
                      collection = NULL,
                      subcollection = NULL,
                      max_show = 30) {

  tb <- msigdbr(species = species)

  if (!is.null(collection))    tb <- tb %>% dplyr::filter(.data$gs_collection == collection)
  if (!is.null(subcollection)) tb <- tb %>% dplyr::filter(.data$gs_subcollection == subcollection)

  # case-insensitive by lowercasing both sides (no ignore_case needed)
  pat <- tolower(pattern)

  hits <- tb %>%
    dplyr::distinct(.data$gs_name, .data$gs_description) %>%
    dplyr::mutate(desc = as.character(.data$gs_description)) %>%
    dplyr::filter(
      stringr::str_detect(tolower(.data$gs_name), pat) |
      stringr::str_detect(tolower(.data$desc %||% ""), pat)
    )

  if (nrow(hits) == 0) {
    message("No match for: ", pattern, "\nHere are some nearby candidates (try a broader keyword):")

    # also case-insensitive without ignore_case
    print(
      tb %>%
        dplyr::distinct(.data$gs_name, .data$gs_description) %>%
        dplyr::mutate(desc = as.character(.data$gs_description)) %>%
        dplyr::filter(
          stringr::str_detect(tolower(.data$gs_name), "antigen|present|mhc|dendrit|process") |
          stringr::str_detect(tolower(.data$desc %||% ""), "antigen|present|mhc|dendrit|process")
        ) %>%
        dplyr::slice_head(n = max_show)
    )

    stop("No msigdbr match for pattern: '", pattern,
         "' (collection=", collection, " subcollection=", subcollection, ")")
  }

  if (nrow(hits) > 1) {
    message("Multiple matches for pattern '", pattern, "'. Candidates:")
    print(hits %>% dplyr::slice_head(n = max_show))
    stop("Pick ONE exact gs_name from the printed candidates and use msig_exact(gs_name).")
  }

  gs <- hits$gs_name[1]
  unique(tb$gene_symbol[tb$gs_name == gs])
}
```

```{r}
#========================================
# 7. Define helper functions for searching enrichment result files
#========================================
# Provide tools to scan external CSV enrichment outputs for pathway keywords and
# help identify candidate biological programs to include.


`%||%` <- function(a, b) if (!is.null(a)) a else b  # <-- needed for msig_go_id()

# ---- helper: safe CSV read ----
read_csv_safe <- function(f) {
  tryCatch(
    readr::read_csv(f, show_col_types = FALSE, progress = FALSE),
    error = function(e) NULL
  )
}

# ---- main search function ----
search_terms_in_csv_folder <- function(
  folder,
  queries,
  recursive = TRUE,
  ignore_case = TRUE,
  cols_prefer = c("Description","description","term","Term","pathway","Pathway",
                  "gs_name","GS_NAME","ID","id","NAME","Name","set","geneset"),
  max_matches_per_file = 200
) {
  stopifnot(dir.exists(folder))
  if (is.character(queries)) queries <- as.list(queries)

  files <- list.files(folder, pattern = "\\.csv$", full.names = TRUE, recursive = recursive)
  if (length(files) == 0) stop("No .csv files found under: ", folder)

  # normalize queries (literal search; avoids regex surprises)
  q_vec <- unlist(queries)
  q_vec_cmp <- if (ignore_case) tolower(q_vec) else q_vec

  res <- purrr::map_dfr(files, function(f) {
    df <- read_csv_safe(f)
    if (is.null(df) || nrow(df) == 0) return(NULL)

    cn <- colnames(df)

    prefer_present <- intersect(cols_prefer, cn)
    if (length(prefer_present) == 0) {
      search_cols <- cn[vapply(df, function(x) is.character(x) || is.factor(x), logical(1))]
    } else {
      search_cols <- prefer_present
    }
    if (length(search_cols) == 0) return(NULL)

    mat <- df[, search_cols, drop = FALSE]
    mat <- as.data.frame(lapply(mat, as.character), stringsAsFactors = FALSE)
    row_text <- apply(mat, 1, function(x) paste(na.omit(x), collapse = " | "))

    row_text_cmp <- if (ignore_case) tolower(row_text) else row_text

    # literal "contains" match for any query
    hit_idx <- which(vapply(row_text_cmp, function(s) any(stringr::str_detect(s, stringr::fixed(q_vec_cmp))), logical(1)))
    if (length(hit_idx) == 0) return(NULL)

    hit_idx <- head(hit_idx, max_matches_per_file)

    tibble(
      file = f,
      row = hit_idx,
      matched_text = row_text[hit_idx],
      searched_columns = paste(search_cols, collapse = ", ")
    )
  })

  res %>% arrange(file, row)
}

# ---- fetch a C5:BP gene set using GO ID (if MSigDB description contains the GO id) ----
msig_go_id <- function(go_id, species="Homo sapiens", collection="C5", subcollection="BP") {
  tb <- msigdbr(species = species) %>%
    dplyr::filter(.data$gs_collection == collection, .data$gs_subcollection == subcollection)

  hits <- tb %>%
    dplyr::distinct(.data$gs_name, .data$gs_description) %>%
    dplyr::mutate(desc = as.character(.data$gs_description)) %>%
    dplyr::filter(
      stringr::str_detect(.data$gs_name, stringr::fixed(go_id)) |
      stringr::str_detect(.data$desc %||% "", stringr::fixed(go_id))
    )

  if (nrow(hits) == 0) stop("No MSigDB ", collection, ":", subcollection, " set found containing GO id: ", go_id)
  if (nrow(hits) > 1) {
    message("Multiple matches for GO id ", go_id, ". Pick one gs_name below and use msig_exact():")
    print(hits)
    stop("Ambiguous GO id match.")
  }

  msig_exact(hits$gs_name[1], species = species)
}

```

```{r}
#========================================
# 8. Search enrichment result files for biologically relevant pathway terms
#========================================
# Query previous enrichment-analysis tables to identify pathways related to
# antigen presentation, complement, prostaglandin signaling, fibroblasts, and leukocytes.

results_folder <- "/media/Aleksandra/Computational/R programming/paper_projects/GSEA/output enrichment analysis/csv/"

queries <- c(
  "antigen processing", "presentation", "mhc",
  "complement ",
  "prostaglandin",
  "fibroblasts",
  "leukocyte", "rolling",
  "extrafollicular", "follicular", "sars", "cov"
)

hits <- search_terms_in_csv_folder(results_folder, queries, recursive = TRUE)

hits

```


```{r}
#========================================
# 9. Build the final pathway gene-set panel
#========================================
# Define the selected MSigDB gene sets, intersect them with detected genes, and
# retain only pathways with sufficient gene coverage in the dataset.

gene_sets <- list(
  "H: INFLAMMATORY_RESPONSE" = msig_exact("HALLMARK_INFLAMMATORY_RESPONSE"),
  "H: IL6_JAK_STAT3"         = msig_exact("HALLMARK_IL6_JAK_STAT3_SIGNALING"),
  "H: IFNG_RESPONSE"         = msig_exact("HALLMARK_INTERFERON_GAMMA_RESPONSE"),
  "H: IFNA_RESPONSE" =msig_exact("HALLMARK_INTERFERON_ALPHA_RESPONSE"),

  # Gavish (this is a gs_name, so msig_exact is correct)
  "C4: GAVISH_FIBROBLASTS_COMPLEMENT" =
    msig_exact("GAVISH_3CA_METAPROGRAM_FIBROBLASTS_COMPLEMENT"),

  # WikiPathways (IMPORTANT: msigdbr gs_name usually uses underscores, not " | WP####" display text)
  # If your msigdbr truly contains exactly that string, this will work. If not, use the lookup panel to get the exact gs_name.
  "C2: WP_BCELL_SARS_COV2" =
    msig_exact("WP_EXTRAFOLLICULAR_AND_FOLLICULAR_B_CELL_ACTIVATION_BY_SARSCOV2"),

  # GO BP (these look like proper gs_name values)
  "C5: GOBP_REGULATION_OF_ANTIGEN_PROCESSING_AND_PRESENTATION" =
    msig_exact("GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION"),

  "C5: GOBP_RESPONSE_TO_PROSTAGLANDIN_E" =
    msig_exact("GOBP_RESPONSE_TO_PROSTAGLANDIN_E"),

  # Reactome complement
  "C2: REACTOME_INITIAL_TRIGGERING_OF_COMPLEMENT" =
    msig_exact("REACTOME_INITIAL_TRIGGERING_OF_COMPLEMENT"),
  
  "C5: GOBP_LEUKOCYTE_TETHERING_OR_ROLLING" =
    msig_exact("GOBP_LEUKOCYTE_TETHERING_OR_ROLLING")
)

# Intersect gene sets with expression genes and keep sets >= 10 genes
gene_sets_use <- lapply(gene_sets, function(gs) intersect(gs, rownames(expr)))
gene_sets_use <- gene_sets_use[sapply(gene_sets_use, length) >= 10]

cat("Gene set sizes after intersection:\n")
print(sort(sapply(gene_sets_use, length)))

stopifnot(length(gene_sets_use) >= 2)
```

```{r}
#========================================
# 10. Compute ssGSEA pathway scores per spot
#========================================
# Score each nerve spot for the selected pathways using ssGSEA on log-normalized
# expression values.

ssgsea_scores <- gsva(
  expr,
  gene_sets_use,
  method = "ssgsea",
  kcdf = "Gaussian",   # correct for logcounts
  abs.ranking = FALSE
)

# result: pathways x spots

```


dim(ssgsea_scores)
C5: GOBP_REGULATION_OF_ANTIGEN_PROCESSING_AND_PRESENTATION                                           H: IFNG_RESPONSE 
                                                        93                                                        188 
                                  H: INFLAMMATORY_RESPONSE 
                                                       198 
                                                       
```{r}
#========================================
# 11. Aggregate pathway scores from spot level to nerve level
#========================================
# Average spot-level ssGSEA scores within each nerve so that each nerve becomes
# the analysis unit for group comparisons.

score_df <- as.data.frame(t(ssgsea_scores))
score_df$spot_id <- rownames(score_df)

score_long <- score_df %>%
  tidyr::pivot_longer(
    cols = -spot_id,
    names_to = "pathway",
    values_to = "ssgsea_score"
  ) %>%
  left_join(meta %>% tibble::rownames_to_column("spot_id"),
            by = "spot_id")

# ---- aggregate per NerveID ----
nerve_scores <- score_long %>%
  group_by(NerveID, pathway, nerve_types, patient_id) %>%
  summarise(
    mean_score = mean(ssgsea_score),
    .groups = "drop"
  )
```

```{r}
#========================================
# 12. Standardize pathway scores across nerves
#========================================
# Convert aggregated pathway scores to z-scores to allow easier comparison of
# relative pathway activity across nerve categories.

nerve_scores <- nerve_scores %>%
  group_by(pathway) %>%
  mutate(
    z_score = as.numeric(scale(mean_score))
  ) %>%
  ungroup()
```

```{r}
#========================================
# 13. Inspect available pathways before plotting
#========================================
# Check which pathways are present in the scored dataset and confirm that the
# requested pathways are available for visualization.

# what’s available in the plotted data:
sort(unique(nerve_scores$pathway))

# what’s available in your gene_sets list:
names(gene_sets)
```


```{r}
#========================================
# 14. Visualize selected complement and stromal-related pathways
#========================================
# Plot violin, boxplot, and patient-level jitter distributions for selected
# complement, fibroblast, and leukocyte-related pathway scores.


# -----------------------------
# USER CONTROLS
# -----------------------------
gene_sets_to_plot <- c(
  "C5: GOBP_LEUKOCYTE_TETHERING_OR_ROLLING",
  "C2: REACTOME_INITIAL_TRIGGERING_OF_COMPLEMENT",
  "C4: GAVISH_FIBROBLASTS_COMPLEMENT"
)

facet_ncol   <- 1      # <--- set columns here
facet_scales <- "free_y"
base_size    <- 11
panel_aspect <- 0.5   # <--- >1 makes panels taller; <1 flatter

# Export controls (this is what truly controls the final graph size)
out_file <- "nerve_scores_violin_facets.pdf"
out_w    <- 7.0        # inches
out_h    <- 4.5        # inches

# -----------------------------
# DATA
# -----------------------------
nerve_scores_plot <- nerve_scores %>%
  filter(pathway %in% gene_sets_to_plot) %>%
  mutate(
    pathway = factor(pathway, levels = gene_sets_to_plot),
    nerve_types = factor(nerve_types)  # optionally set your order here
  )

# -----------------------------
# PLOT
# -----------------------------
p1 <- ggplot(nerve_scores_plot, aes(x = nerve_types, y = z_score)) +
  geom_violin(fill = "grey85", color = NA, alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.25, outlier.shape = NA,
               fill = "white", color = "grey30", linewidth = 0.4) +
  geom_jitter(aes(color = patient_id),
              width = 0.10, size = 0.9, alpha = 0.95) +
  facet_wrap(~ pathway, scales = facet_scales, ncol = facet_ncol) +
  theme_classic(base_size = base_size) +
  theme(
    aspect.ratio = panel_aspect,
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = base_size * 0.85),
    legend.position = "right",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4),
    panel.spacing = unit(6, "pt")   # spacing between facets
  ) +
  ylab("ssGSEA score (SD units, per nerve)") +
  xlab("Nerve type") +
  labs(color = "Patient ID")

p1

# -----------------------------
# SAVE (recommended)
# -----------------------------
#ggsave(out_file, p1, width = out_w, height = out_h, units = "in", useDingbats = FALSE)

```

```{r}
#========================================
# 15. Save complement and stromal pathway figures
#========================================
# Export the first panel of pathway activity plots for use in figures or reports.

ggsave(
  filename = "complememnt_rolling.pdf",
  plot = p1,
  width = 5,
  height = 10,
  units = "in",
  dpi = 600
)
```

```{r}
#========================================
# 16. Inspect available pathways before plotting
#========================================
# Check which pathways are present in the scored dataset and confirm that the
# requested pathways are available for visualization.

# what’s available in the plotted data:
sort(unique(nerve_scores$pathway))

# what’s available in your gene_sets list:
names(gene_sets)
```

```{r}

#========================================
# 17. Visualize inflammatory and cytokine-related pathways
#========================================
# Plot pathway activity distributions for interferon, IL6-JAK-STAT3, and
# prostaglandin-related signatures across nerve groups.


# -----------------------------
# USER CONTROLS
# -----------------------------
gene_sets_to_plot <- c(
  "H: IFNA_RESPONSE",
  "H: IFNG_RESPONSE",
  "H: IL6_JAK_STAT3",
  "C5: GOBP_RESPONSE_TO_PROSTAGLANDIN_E"
)

facet_ncol   <- 2      # <--- set columns here
facet_scales <- "free_y"
base_size    <- 11
panel_aspect <- 0.5   # <--- >1 makes panels taller; <1 flatter

# Export controls (this is what truly controls the final graph size)
out_file <- "nerve_scores_violin_facets.pdf"
out_w    <- 7.0        # inches
out_h    <- 4.5        # inches

# -----------------------------
# DATA
# -----------------------------
nerve_scores_plot <- nerve_scores %>%
  filter(pathway %in% gene_sets_to_plot) %>%
  mutate(
    pathway = factor(pathway, levels = gene_sets_to_plot),
    nerve_types = factor(nerve_types)  # optionally set your order here
  )

# -----------------------------
# PLOT
# -----------------------------
p2 <- ggplot(nerve_scores_plot, aes(x = nerve_types, y = z_score)) +
  geom_violin(fill = "grey85", color = NA, alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.25, outlier.shape = NA,
               fill = "white", color = "grey30", linewidth = 0.4) +
  geom_jitter(aes(color = patient_id),
              width = 0.10, size = 0.9, alpha = 0.95) +
  facet_wrap(~ pathway, scales = facet_scales, ncol = facet_ncol) +
  theme_classic(base_size = base_size) +
  theme(
    aspect.ratio = panel_aspect,
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = base_size * 0.85),
    legend.position = "right",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4),
    panel.spacing = unit(6, "pt")   # spacing between facets
  ) +
  ylab("ssGSEA score (SD units, per nerve)") +
  xlab("Nerve type") +
  labs(color = "Patient ID")

p2

# -----------------------------
# SAVE (recommended)
# -----------------------------
#ggsave(out_file, p, width = out_w, height = out_h, units = "in", useDingbats = FALSE)

```

```{r}
#========================================
# 18. Save inflammatory pathway figures
#========================================
# Export the second panel of pathway activity plots at publication-ready resolution.

ggsave(
  filename = "pathways_figure2.pdf",
  plot = p2,
  width = 10,
  height = 10,
  units = "in",
  dpi = 600
)
```


```{r}
#========================================
# 20. Test pathway differences across nerve categories
#========================================
# Use repeated-measures ANOVA to evaluate whether pathway activity differs across
# nerve categories while accounting for patient-level pairing.

patient_scores <- nerve_scores %>%
  group_by(patient_id, nerve_types, pathway) %>%
  summarise(mean_z = mean(z_score), .groups="drop")


run_rm_anova <- function(df_one_pathway) {
  df_one_pathway <- droplevels(df_one_pathway)
  fit <- aov(mean_z ~ nerve_types + Error(patient_id/nerve_types), data = df_one_pathway)

  # extract the p-value for nerve_types (within-subject)
  sm <- summary(fit)
  p <- sm[["Error: patient_id:nerve_types"]][[1]][["Pr(>F)"]][1]

  tibble(p_value = p)
}

anova_tbl <- patient_scores %>%
  group_by(pathway) %>%
  group_modify(~ run_rm_anova(.x)) %>%
  ungroup() %>%
  mutate(FDR = p.adjust(p_value, method = "BH"))

anova_tbl
```


                                                       

  