#========================================================
# ORA analysis for invaded-nerve signature gene list
#========================================================
#
# Purpose:
#   Perform over-representation analysis (ORA) on a selected invaded-nerve
#   gene list using a defined DE background universe.
#
# Key features:
#   - Uses a proper background gene universe
#   - Saves ALL tested terms, not only significant ones
#   - Saves significant terms separately
#   - Extracts nerve-related terms by keyword search for manual inspection
#   - Uses robust gene-symbol to ENTREZ mapping with manual corrections
#
# Expected input files:
#   - FullDE_List.csv                : background universe, must contain column "Gene"
#   - All_genes_up_invaded_LR.csv    : selected signature, must contain column "Gene"
#
#========================================================

```{r}
#========================================================
# 1. Load packages
#========================================================
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(tibble)
  library(stringr)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(ReactomePA)
  library(DOSE)
  library(msigdbr)
  library(enrichplot)
  library(ggplot2)
  library(openxlsx)
  library(tidyr)
})
```

```{r}
#========================================================
# 2. Parameters
#========================================================

seed_value <- 1234
set.seed(seed_value)

#----- Paths -----
project_dir <- "~/NAS/Aleksandra/Computational/R programming/paper_projects/GSEA"
input_dir   <- file.path(project_dir, "input lists")
output_dir  <- file.path(project_dir, "figure3_ORA_invaded_nerve")

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(output_dir, "csv"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(output_dir, "pdf"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(output_dir, "rds"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(output_dir, "qc"),  recursive = TRUE, showWarnings = FALSE)

background_file <- file.path(input_dir, "FullDE_List.csv")
signature_file  <- file.path(input_dir, "TableS5_invasion_associated_342_genes.csv")

#----- Enrichment settings -----
min_gs_size   <- 10
max_gs_size   <- 800
adjust_method <- "BH"

# IMPORTANT:
# set cutoffs to 1 so you can inspect all tested terms,
# then filter significant terms afterwards
p_cutoff_full <- 1
q_cutoff_full <- 1

# significance threshold for exported filtered tables
sig_padj_cutoff <- 0.05

#----- Keywords for nerve-related inspection -----
nerve_keywords <- c(
  "nerve", "neural", "neuron", "neuron projection", "axon", "axonal",
  "schwann", "glia", "glial", "myelin", "synap", "neuro",
  "regeneration", "regenerative", "injury", "repair", "guidance",
  "neurotrans", "glutam", "peripheral nervous system", "ensheath"
)
```

```{r}
#========================================================
# 3. Helper functions
#========================================================

#--------------------------------------------------------
# 3.1 Check required columns
#--------------------------------------------------------
check_required_columns <- function(df, required_cols = "Gene") {
  missing_cols <- setdiff(required_cols, colnames(df))
  if (length(missing_cols) > 0) {
    stop("Missing required columns: ", paste(missing_cols, collapse = ", "))
  }
}

#--------------------------------------------------------
# 3.2 Map SYMBOL -> ENTREZ with manual corrections
#--------------------------------------------------------
map_symbols_to_entrez <- function(symbols) {
  symbols <- unique(as.character(symbols))
  symbols <- symbols[!is.na(symbols) & symbols != ""]

  mapped <- AnnotationDbi::mapIds(
    org.Hs.eg.db,
    keys      = symbols,
    column    = "ENTREZID",
    keytype   = "SYMBOL",
    multiVals = "first"
  )

  mapped_df <- tibble(
    Gene = names(mapped),
    ENTREZID = as.character(mapped)
  )

  # Manual corrections for symbols not resolved automatically
  manual_map <- c(
    ST5     = "6764",
    TIAF1   = "399687",
    SPATA5  = "166378",
    C3orf14 = "57415",
    ENDNRB  = "1910"
  )

  idx <- mapped_df$Gene %in% names(manual_map) & is.na(mapped_df$ENTREZID)
  mapped_df$ENTREZID[idx] <- manual_map[mapped_df$Gene[idx]]

  mapped_df %>%
    distinct(Gene, .keep_all = TRUE)
}

#--------------------------------------------------------
# 3.3 Save enrichment tables
#--------------------------------------------------------
save_enrichment_outputs <- function(res_obj, analysis_name, output_dir, show_n = 20) {

  csv_dir <- file.path(output_dir, "csv")
  pdf_dir <- file.path(output_dir, "pdf")
  rds_dir <- file.path(output_dir, "rds")

  saveRDS(res_obj, file = file.path(rds_dir, paste0(analysis_name, ".rds")))

  res_df <- as.data.frame(res_obj@result)

  if (nrow(res_df) == 0) {
    readr::write_csv(res_df, file.path(csv_dir, paste0(analysis_name, "_all_terms.csv")))
    return(invisible(res_df))
  }

  # Save all terms
  readr::write_csv(res_df, file.path(csv_dir, paste0(analysis_name, "_all_terms.csv")))

  # Save significant terms
  sig_df <- res_df %>%
    filter(!is.na(p.adjust) & p.adjust < sig_padj_cutoff)

  readr::write_csv(sig_df, file.path(csv_dir, paste0(analysis_name, "_significant_terms.csv")))
  
  # Save nerve-related keyword matches
  keyword_pattern <- paste(nerve_keywords, collapse = "|")
  nerve_df <- res_df %>%
    filter(str_detect(tolower(Description), keyword_pattern) |
           str_detect(tolower(ID), keyword_pattern))

  readr::write_csv(nerve_df, file.path(csv_dir, paste0(analysis_name, "_nerve_keyword_terms.csv")))

  # Plot only significant terms if they exist; otherwise plot top terms by p.adjust
  pdf(file.path(pdf_dir, paste0(analysis_name, "_dotplot.pdf")), width = 10, height = 8, useDingbats = FALSE)
  if (nrow(sig_df) > 0) {
    print(dotplot(res_obj, showCategory = show_n) + ggtitle(analysis_name))
  } else {
    top_df <- res_df %>%
      arrange(p.adjust, pvalue) %>%
      slice_head(n = min(show_n, nrow(.)))

    if (nrow(top_df) > 0) {
      # rebuild temporary object with top rows for plotting
      tmp_obj <- res_obj
      tmp_obj@result <- top_df
      print(dotplot(tmp_obj, showCategory = min(show_n, nrow(top_df))) + ggtitle(paste0(analysis_name, " (top terms, none significant)")))
    } else {
      plot.new()
      title(main = paste("No terms returned:", analysis_name))
    }
  }
  dev.off()

  invisible(res_df)
}

#--------------------------------------------------------
# 3.4 Generic ORA runner for TERM2GENE
#--------------------------------------------------------
run_ora_term2gene <- function(gene, universe, term2gene, analysis_name, readable = TRUE) {
  ora_obj <- enricher(
    gene          = gene,
    universe      = universe,
    TERM2GENE     = term2gene,
    pvalueCutoff  = p_cutoff_full,
    qvalueCutoff  = q_cutoff_full,
    pAdjustMethod = adjust_method,
    minGSSize     = min_gs_size,
    maxGSSize     = max_gs_size
  )

  if (!is.null(ora_obj) && readable) {
    ora_obj <- setReadable(ora_obj, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
  }

  save_enrichment_outputs(ora_obj, analysis_name, output_dir)
  return(ora_obj)
}

#--------------------------------------------------------
# 3.5 GO/Reactome wrappers
#--------------------------------------------------------
run_ora_go <- function(gene, universe, ont, analysis_name) {
  ora_obj <- enrichGO(
    gene          = gene,
    universe      = universe,
    OrgDb         = org.Hs.eg.db,
    keyType       = "ENTREZID",
    ont           = ont,
    pvalueCutoff  = p_cutoff_full,
    qvalueCutoff  = q_cutoff_full,
    pAdjustMethod = adjust_method,
    minGSSize     = min_gs_size,
    maxGSSize     = max_gs_size,
    readable      = TRUE
  )

  save_enrichment_outputs(ora_obj, analysis_name, output_dir)
  return(ora_obj)
}

run_ora_reactome <- function(gene, universe, analysis_name) {
  ora_obj <- enrichPathway(
    gene          = gene,
    universe      = universe,
    organism      = "human",
    pvalueCutoff  = p_cutoff_full,
    qvalueCutoff  = q_cutoff_full,
    pAdjustMethod = adjust_method,
    minGSSize     = min_gs_size,
    maxGSSize     = max_gs_size,
    readable      = TRUE
  )

  save_enrichment_outputs(ora_obj, analysis_name, output_dir)
  return(ora_obj)
}

run_ora_do <- function(gene, universe, analysis_name) {
  ora_obj <- enrichDO(
    gene          = gene,
    universe      = universe,
    ont           = "DO",
    pvalueCutoff  = p_cutoff_full,
    qvalueCutoff  = q_cutoff_full,
    pAdjustMethod = adjust_method,
    minGSSize     = min_gs_size,
    maxGSSize     = max_gs_size,
    readable      = TRUE
  )

  save_enrichment_outputs(ora_obj, analysis_name, output_dir)
  return(ora_obj)
}

```

```{r}
#========================================================
# 4. Read input data
#========================================================

background <- readr::read_csv(background_file, show_col_types = FALSE)
signature  <- readr::read_csv(signature_file, show_col_types = FALSE)

check_required_columns(background, "Gene")
check_required_columns(signature,  "Gene")

cat("Background genes:", nrow(background), "\n")
cat("Signature genes:", nrow(signature), "\n")
```

```{r}
#========================================================
# 5. Map genes to ENTREZ IDs
#========================================================

background_map <- map_symbols_to_entrez(background$Gene)
signature_map  <- map_symbols_to_entrez(signature$Gene)

# Keep only mapped genes
background_map_filt <- background_map %>% filter(!is.na(ENTREZID))
signature_map_filt  <- signature_map %>% filter(!is.na(ENTREZID))

# Restrict signature to genes present in the background universe
signature_map_filt <- signature_map_filt %>%
  filter(ENTREZID %in% background_map_filt$ENTREZID)

# Character vectors required by clusterProfiler
gene_list <- as.character(signature_map_filt$ENTREZID)
universe  <- as.character(background_map_filt$ENTREZID)

cat("Mapped background genes:", length(universe), "\n")
cat("Mapped signature genes:", length(gene_list), "\n")

# Save mapping tables
readr::write_csv(background_map,      file.path(output_dir, "qc", "background_symbol_to_entrez.csv"))
readr::write_csv(signature_map,       file.path(output_dir, "qc", "signature_symbol_to_entrez.csv"))
readr::write_csv(background_map_filt, file.path(output_dir, "qc", "background_mapped_used.csv"))
readr::write_csv(signature_map_filt,  file.path(output_dir, "qc", "signature_mapped_used.csv"))
```

```{r}
#========================================================
# 6. Run GO and pathway ORA
#========================================================

ora_go_bp <- run_ora_go(gene_list, universe, ont = "BP", analysis_name = "ora_go_bp")
ora_go_cc <- run_ora_go(gene_list, universe, ont = "CC", analysis_name = "ora_go_cc")
ora_go_mf <- run_ora_go(gene_list, universe, ont = "MF", analysis_name = "ora_go_mf")

ora_reactome <- run_ora_reactome(gene_list, universe, analysis_name = "ora_reactome")
ora_do       <- run_ora_do(gene_list, universe, analysis_name = "ora_disease_ontology")
```

```{r}
#========================================================
# 7. Run MSigDB ORA across collections
#========================================================

run_msigdb_ora <- function(collection_name) {
  term2gene <- msigdbr(
    species = "Homo sapiens",
    category = collection_name
  ) %>%
    dplyr::select(gs_name, entrez_gene) %>%
    dplyr::distinct()

  run_ora_term2gene(
    gene = gene_list,
    universe = universe,
    term2gene = term2gene,
    analysis_name = paste0("ora_msigdb_", collection_name),
    readable = TRUE
  )
}

ora_msigdb_h  <- run_msigdb_ora("H")
ora_msigdb_c1 <- run_msigdb_ora("C1")
ora_msigdb_c2 <- run_msigdb_ora("C2")
ora_msigdb_c3 <- run_msigdb_ora("C3")
ora_msigdb_c4 <- run_msigdb_ora("C4")
ora_msigdb_c5 <- run_msigdb_ora("C5")
ora_msigdb_c6 <- run_msigdb_ora("C6")
ora_msigdb_c7 <- run_msigdb_ora("C7")
ora_msigdb_c8 <- run_msigdb_ora("C8")
```

```{r}
#========================================================
# 8. Combine nerve-keyword results across analyses
#========================================================

csv_dir <- file.path(output_dir, "csv")

nerve_files <- list.files(
  csv_dir,
  pattern = "_nerve_keyword_terms\\.csv$",
  full.names = TRUE
)

nerve_summary <- lapply(nerve_files, function(f) {
  df <- readr::read_csv(f, show_col_types = FALSE)
  if (nrow(df) == 0) return(NULL)
  df$analysis <- basename(f) %>% str_remove("_nerve_keyword_terms\\.csv$")
  df
}) %>%
  bind_rows()

readr::write_csv(
  nerve_summary,
  file.path(csv_dir, "combined_nerve_keyword_terms_across_analyses.csv")
)

cat("Combined nerve-keyword table saved.\n")
```

```{r}
#========================================================
# 9. Combine suggestive terms across analyses - top 15 form each
#========================================================

csv_dir <- file.path(output_dir, "csv")

all_term_files <- list.files(
  csv_dir,
  pattern = "_all_terms\\.csv$",
  full.names = TRUE
)

suggestive_summary <- lapply(all_term_files, function(f) {
  df <- readr::read_csv(f, show_col_types = FALSE)

  if (nrow(df) == 0) return(NULL)

  df %>%
    filter(!is.na(pvalue) & pvalue < 0.05) %>%
    filter(is.na(p.adjust) | p.adjust >= sig_padj_cutoff) %>%
    arrange(pvalue) %>%   # ensure ranking within each file
    slice_head(n = 15) %>%  # <-- restrict to top 15 per comparison
    mutate(analysis = basename(f) %>% str_remove("_all_terms\\.csv$"))
}) %>%
  bind_rows() %>%
  arrange(pvalue)

readr::write_csv(
  suggestive_summary,
  file.path(csv_dir, "combined_suggestive_terms_rawp_lt_0.05_top15_per_analysis.csv")
)

cat("Combined suggestive-term table (top 15 per analysis) saved.\n")
```

```{r}
#========================================================
# 10. Combine suggestive terms across analyses (global top 100)
#========================================================

csv_dir <- file.path(output_dir, "csv")

all_term_files <- list.files(
  csv_dir,
  pattern = "_all_terms\\.csv$",
  full.names = TRUE
)

suggestive_summary <- lapply(all_term_files, function(f) {
  df <- readr::read_csv(f, show_col_types = FALSE)

  if (nrow(df) == 0) return(NULL)

  df %>%
    filter(!is.na(pvalue) & pvalue < 0.05) %>%
    filter(is.na(p.adjust) | p.adjust >= sig_padj_cutoff) %>%
    mutate(analysis = basename(f) %>% str_remove("_all_terms\\.csv$"))
}) %>%
  bind_rows() %>%
  arrange(pvalue) %>%        # global ranking across all comparisons
  slice_head(n = 100)        # <-- top 100 overall

readr::write_csv(
  suggestive_summary,
  file.path(csv_dir, "combined_suggestive_terms_top100_global.csv")
)

cat("Combined suggestive-term table (top 100 global) saved.\n")
```


```{r}
#========================================================
# 10. Save session info
#========================================================

writeLines(
  capture.output(sessionInfo()),
  con = file.path(output_dir, "qc", "sessionInfo.txt")
)
```

