# ============================================================
# Cas12a library QC and normalization workflow
# ============================================================
#
# Goal:
# 1) Read the merged guide-count matrix across replicate libraries
# 2) Compute simple library-size normalized CPM values
# 3) Log-transform CPM values for replicate-to-replicate QC
# 4) Assess similarity between replicates by correlation and PCA
# 5) Quantify guide-level variability (CV)
# 6) Summarize per-replicate count distribution metrics
# 7) Compare replicate library QC to the original plasmid/library stock
#
# Notes:
# - CPM + log2(CPM + 1) is used here for exploratory QC plots
# - DESeq2 VST is used for PCA because it stabilizes the variance better
#   than raw counts or plain log-transformed CPM
# - No differential testing is performed in this script
# ============================================================

```{r}
# -----------------------------
# 0) Load required packages
# -----------------------------
suppressPackageStartupMessages({
  library(edgeR)
  library(DESeq2)
  library(pheatmap)
  library(ggplot2)
  library(ggrepel)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(gridExtra)
  library(reldist)
})
```

```{r}
# -----------------------------
# 1) Define input/output files
# -----------------------------
count_file <- "input_files/merged_counts_final_libraryQC.csv"
stock_file <- "input_files/M-AS11_AAJY01_RDA550_G0_CP1906_ScoresSum.csv"
outdir <- "output_files/libraryQC/library_qc_outputs"
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)
```

```{r}
# -----------------------------
# 2) Read merged replicate count matrix
# -----------------------------
# Expected structure:
# - first column = guide/clone identifier
# - remaining columns = replicate count columns
count_df <- read.csv(count_file, check.names = FALSE)

# Preserve guide IDs explicitly
guide_id_col <- colnames(count_df)[1]
colnames(count_df)[1] <- "guide_id"

# Keep only replicate count columns
count_mat <- count_df %>%
  dplyr::select(-guide_id)

# Convert to numeric matrix
count_mat <- as.matrix(count_mat)
mode(count_mat) <- "numeric"
rownames(count_mat) <- count_df$guide_id

# Optional sanity check
stopifnot(all(!is.na(colnames(count_mat))))
stopifnot(ncol(count_mat) >= 2)
```

```{r}
# -----------------------------
# 3) Compute library-size normalization (CPM)
# -----------------------------
# CPM rescales each sample by its total sequencing depth:
# CPM = counts / total_reads * 1e6
library_sizes <- colSums(count_mat, na.rm = TRUE)

cpm_mat <- sweep(count_mat, 2, library_sizes, FUN = "/") * 1e6

# Log-transform CPM for QC visualizations
log2_cpm_mat <- log2(cpm_mat + 1)

# -----------------------------
# 4) Replicate correlation heatmap
# -----------------------------
# Spearman correlation is robust to non-normality and useful for ranking similarity
cor_mat <- cor(log2_cpm_mat, method = "spearman", use = "pairwise.complete.obs")
print(cor_mat)

pheatmap(
  cor_mat,
  display_numbers = TRUE,
  color = colorRampPalette(c("green", "white", "red"))(50),
  main = "Spearman correlation between replicate libraries",
  filename = file.path(outdir, "replicate_spearman_correlation_heatmap.pdf"),
  width = 7,
  height = 6
)
```

```{r}
# -----------------------------
# 5) Guide-level variability after log2(CPM+1)
# -----------------------------
# For each guide:
# - mean abundance across replicates
# - SD across replicates
# - coefficient of variation (CV = SD / mean)
guide_mean_log2cpm <- rowMeans(log2_cpm_mat, na.rm = TRUE)
guide_sd_log2cpm   <- apply(log2_cpm_mat, 1, sd, na.rm = TRUE)
guide_cv_log2cpm   <- ifelse(guide_mean_log2cpm == 0, NA, guide_sd_log2cpm / guide_mean_log2cpm)

cv_df <- data.frame(
  guide_id = rownames(log2_cpm_mat),
  mean_log2CPM = guide_mean_log2cpm,
  sd_log2CPM = guide_sd_log2cpm,
  cv_log2CPM = guide_cv_log2cpm
)

write.csv(cv_df, file.path(outdir, "guide_level_variability_log2CPM.csv"), row.names = FALSE)

# CV histogram
p_cv_hist <- ggplot(cv_df, aes(x = cv_log2CPM)) +
  geom_histogram(bins = 50, fill = "blue", alpha = 0.7) +
  labs(
    title = "Distribution of guide-level CV",
    x = "CV on log2(CPM + 1)",
    y = "Number of guides"
  ) +
  theme_minimal()

ggsave(
  filename = file.path(outdir, "guide_cv_distribution.pdf"),
  plot = p_cv_hist,
  width = 7,
  height = 5
)

# CV vs mean plot
p_cv_mean <- ggplot(cv_df, aes(x = mean_log2CPM, y = cv_log2CPM)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  labs(
    title = "Guide variability as a function of abundance",
    x = "Mean log2(CPM + 1)",
    y = "CV on log2(CPM + 1)"
  ) +
  theme_minimal()

ggsave(
  filename = file.path(outdir, "guide_cv_vs_mean.pdf"),
  plot = p_cv_mean,
  width = 7,
  height = 5
)
```

```{r}
# -----------------------------
# 6) Filter low-count guides before VST / PCA
# -----------------------------
# Keep guides with enough total counts across all replicates
# Adjust threshold if needed depending on library complexity
min_total_count <- 100
filtered_counts <- count_mat[rowSums(count_mat, na.rm = TRUE) > min_total_count, , drop = FALSE]

# DESeq2 expects integer counts
filtered_counts_int <- round(filtered_counts)
mode(filtered_counts_int) <- "integer"

# Dummy metadata because PCA is only for QC here
colData <- data.frame(
  sample = colnames(filtered_counts_int),
  condition = "library_replicate",
  row.names = colnames(filtered_counts_int)
)

dds <- DESeqDataSetFromMatrix(
  countData = filtered_counts_int,
  colData = colData,
  design = ~ 1
)

# VST reduces mean-variance dependence for count data
vst_obj <- vst(dds, blind = TRUE)
vst_mat <- assay(vst_obj)
```

```{r}
# -----------------------------
# 7) PCA on variance-stabilized counts
# -----------------------------
pca_res <- prcomp(t(vst_mat), scale. = TRUE)

pca_df <- data.frame(
  Sample = rownames(pca_res$x),
  PC1 = pca_res$x[, 1],
  PC2 = pca_res$x[, 2]
)

# Variance explained
pca_var <- summary(pca_res)$importance[2, 1:2] * 100

p_pca <- ggplot(pca_df, aes(x = PC1, y = PC2, label = Sample)) +
  geom_point(size = 3, alpha = 0.7, color = "blue") +
  geom_text_repel(size = 4) +
  labs(
    title = "PCA of replicate libraries (DESeq2 VST)",
    x = paste0("PC1 (", round(pca_var[1], 1), "%)"),
    y = paste0("PC2 (", round(pca_var[2], 1), "%)")
  ) +
  theme_minimal()

ggsave(
  filename = file.path(outdir, "replicate_PCA_VST.pdf"),
  plot = p_pca,
  width = 7,
  height = 5
)

print(summary(pca_res))
```

```{r}
# -----------------------------
# 8) Per-replicate QC summary metrics
# -----------------------------
# Convert replicate count matrix to long format so we can summarize each sample
library_qc_long <- as.data.frame(count_mat) %>%
  tibble::rownames_to_column("guide_id") %>%
  pivot_longer(
    cols = -guide_id,
    names_to = "replicate",
    values_to = "reads"
  )

summary_stat <- library_qc_long %>%
  group_by(replicate) %>%
  summarise(
    total_reads = sum(reads, na.rm = TRUE),
    median_reads = median(reads, na.rm = TRUE),
    IQR_reads = IQR(reads, na.rm = TRUE),
    percent_low_reads = 100 * mean(reads < 5, na.rm = TRUE),
    percent_zero_count = 100 * mean(reads == 0, na.rm = TRUE),
    gini_index = reldist::gini(reads),
    .groups = "drop"
  )

print(summary_stat)
write.csv(summary_stat, file.path(outdir, "replicate_qc_summary.csv"), row.names = FALSE)
```

```{r}
# -----------------------------
# 9) Histograms of replicate count distributions
# -----------------------------
replicate_names <- colnames(count_mat)

hist_plots <- map(replicate_names, function(rep_name) {
  ggplot(
    data.frame(reads = count_mat[, rep_name]),
    aes(x = log2(reads + 1))
  ) +
    geom_histogram(fill = "blue", bins = 50) +
    theme_minimal() +
    labs(
      title = paste("Histogram of", rep_name),
      x = "log2(read count + 1)",
      y = "Number of guides"
    )
})

pdf(file.path(outdir, "library_qc_histograms.pdf"), width = 10, height = 8)
grid.arrange(grobs = hist_plots, ncol = 2, nrow = 2)
dev.off()
```

```{r}
# -----------------------------
# 10) QC metrics for original library stock
# -----------------------------
# This section summarizes the plasmid/library stock distribution,
# so it can be compared with the sequenced replicate libraries.
stock_df <- read.csv(stock_file, check.names = FALSE)

# Keep only the guide ID and the count column if extra annotation columns exist
# Adjust this section if your stock file has a different structure
stock_counts <- stock_df[, -c(2, 3), drop = FALSE]

# Rename first column to guide_id for clarity
colnames(stock_counts)[1] <- "guide_id"
stock_value_col <- colnames(stock_counts)[2]

summary_stat_stock <- stock_counts %>%
  summarise(
    total_reads = sum(.data[[stock_value_col]], na.rm = TRUE),
    median_reads = median(.data[[stock_value_col]], na.rm = TRUE),
    IQR_reads = IQR(.data[[stock_value_col]], na.rm = TRUE),
    percent_low_reads = 100 * mean(.data[[stock_value_col]] < 5, na.rm = TRUE),
    percent_zero_count = 100 * mean(.data[[stock_value_col]] == 0, na.rm = TRUE),
    gini_index = reldist::gini(.data[[stock_value_col]])
  )

print(summary_stat_stock)
write.csv(summary_stat_stock, file.path(outdir, "stock_library_qc_summary.csv"), row.names = FALSE)
```

```{r}
# -----------------------------
# 11) Save normalized matrices
# -----------------------------
write.csv(
  cbind(guide_id = rownames(cpm_mat), as.data.frame(cpm_mat)),
  file.path(outdir, "guide_counts_CPM.csv"),
  row.names = FALSE
)

write.csv(
  cbind(guide_id = rownames(log2_cpm_mat), as.data.frame(log2_cpm_mat)),
  file.path(outdir, "guide_counts_log2CPM_plus1.csv"),
  row.names = FALSE
)
```

```{r}
# -----------------------------
# 12) Session summary
# -----------------------------
cat("QC workflow completed.\n")
cat("Input count matrix:", count_file, "\n")
cat("Number of guides:", nrow(count_mat), "\n")
cat("Number of replicate libraries:", ncol(count_mat), "\n")
cat("Filtered guides used for VST/PCA:", nrow(filtered_counts_int), "\n")
cat("Outputs written to:", outdir, "\n")
```