# =========================================================
# MAGeCK-oriented filtering strategies
# =========================================================
#
# Goal:
# Create filtered RAW count matrices for MAGeCK, where filtering is
# decided using voom-normalized data from selected reference groups.
#
# Filtering strategies:
# 1) library_input + T0
# 2) T0 only
# 3) T0 + control   <-- kept as in previous analysis for reproducibility
#
# Important:
# - Filtering is based on mean voom logCPM in selected samples
# - Saved output files are RAW counts (not normalized), suitable for MAGeCK
# =========================================================

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(edgeR)
  library(limma)
  library(stringr)
})
```


```{r}
# =========================================================
# 1. Paths
# =========================================================
input_file <- "input_files/Library_and_Screen.csv"
outdir <- "output_files/mageck_filtering"
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)
```

```{r}
# =========================================================
# 2. Read count matrix
# Expected columns:
#   sgRNA or sgRNAID | Gene | sample1 | sample2 | ...
# =========================================================
count_matrix <- read.csv(input_file, check.names = FALSE)

if ("sgRNAID" %in% colnames(count_matrix) && !"sgRNA" %in% colnames(count_matrix)) {
  colnames(count_matrix)[colnames(count_matrix) == "sgRNAID"] <- "sgRNA"
}

stopifnot(all(c("sgRNA", "Gene") %in% colnames(count_matrix)))

meta_cols <- c("sgRNA", "Gene")
sample_cols <- setdiff(colnames(count_matrix), meta_cols)

meta_info <- count_matrix[, meta_cols, drop = FALSE]
raw_counts <- count_matrix[, sample_cols, drop = FALSE]

raw_counts[] <- lapply(raw_counts, as.numeric)

if (any(is.na(raw_counts))) {
  warning("Some count values are NA. Replacing NA with 0.")
  raw_counts[is.na(raw_counts)] <- 0
}

raw_counts_mat <- as.matrix(raw_counts)
rownames(raw_counts_mat) <- count_matrix$sgRNA
```

```{r}
# =========================================================
# 3. Define sample groups
# Adjust patterns here if your sample names differ
# =========================================================
sample_group_df <- tibble(
  Sample = sample_cols,
  Group = case_when(
    str_detect(sample_cols, "rep\\d*_counts") ~ "library_input",
    str_detect(sample_cols, "^Draxin") ~ "draxin",
    str_detect(sample_cols, "^Ctr") ~ "control",
    str_detect(sample_cols, "^Before") ~ "T0",
    TRUE ~ "unknown"
  )
)

write.csv(
  sample_group_df,
  file.path(outdir, "sample_group_assignment.csv"),
  row.names = FALSE
)
```

```{r}
run_voom_filter <- function(raw_counts_mat,
                            meta_info,
                            sample_group_df,
                            keep_groups,
                            filter_label,
                            outdir,
                            mean_logcpm_cutoff = 2.5) {

 selected_samples <- sample_group_df %>%
  dplyr::filter(Group %in% keep_groups) %>%
  dplyr::pull(Sample)

selected_samples_label <- paste(selected_samples, collapse = "; ")
n_reference_samples <- length(selected_samples)

  if (length(selected_samples) < 1) {
    warning(paste0("Skipping ", filter_label, ": no selected samples found."))
    return(NULL)
  }

  missing_samples <- setdiff(selected_samples, colnames(raw_counts_mat))
  if (length(missing_samples) > 0) {
    stop(
      "These selected samples are missing from raw_counts_mat: ",
      paste(missing_samples, collapse = ", ")
    )
  }

  sample_order <- colnames(raw_counts_mat)
  count_data_sub <- raw_counts_mat[, selected_samples, drop = FALSE]

  group_sub <- sample_group_df$Group[match(selected_samples, sample_group_df$Sample)]
  group_sub <- factor(group_sub)

  if (nlevels(group_sub) < 2) {
    design_sub <- matrix(1, nrow = length(selected_samples), ncol = 1)
    colnames(design_sub) <- "Intercept"
    rownames(design_sub) <- selected_samples
  } else {
    design_sub <- model.matrix(~ group_sub)
    rownames(design_sub) <- selected_samples
  }

  dge_sub <- edgeR::DGEList(counts = count_data_sub)
  dge_sub <- edgeR::calcNormFactors(dge_sub)

  v_sub <- limma::voom(dge_sub, design_sub, plot = FALSE)

  mean_logcpm_sub <- rowMeans(v_sub$E, na.rm = TRUE)
  keep_idx <- mean_logcpm_sub >= mean_logcpm_cutoff

  filtered_raw_counts <- raw_counts_mat[keep_idx, , drop = FALSE]

  filtered_raw_df <- filtered_raw_counts %>%
    as.data.frame() %>%
    tibble::rownames_to_column(var = "sgRNA") %>%
    dplyr::left_join(meta_info, by = "sgRNA") %>%
    dplyr::select(sgRNA, Gene, dplyr::all_of(sample_order))

  stats_df <- tibble::tibble(
  filter_name = filter_label,
  reference_groups = paste(keep_groups, collapse = " + "),
  selected_samples = selected_samples_label,
  n_reference_samples = n_reference_samples,
  n_reference_groups = nlevels(group_sub),
  mean_logcpm_cutoff = mean_logcpm_cutoff,
  n_sgRNAs_before = nrow(raw_counts_mat),
  n_sgRNAs_after = sum(keep_idx),
  n_sgRNAs_removed = sum(!keep_idx)
)
  
  keep_table <- tibble::tibble(
    sgRNA = rownames(raw_counts_mat),
    Gene = meta_info$Gene[match(rownames(raw_counts_mat), meta_info$sgRNA)],
    mean_logCPM_reference = mean_logcpm_sub,
    keep = keep_idx
  )

  write.csv(
    filtered_raw_df,
    file.path(outdir, paste0("MAGECK_input_", filter_label, "_RAW_COUNTS.csv")),
    row.names = FALSE,
    quote = FALSE
  )

  write.csv(
    keep_table,
    file.path(outdir, paste0("MAGECK_filter_decisions_", filter_label, ".csv")),
    row.names = FALSE
  )

  write.csv(
    stats_df,
    file.path(outdir, paste0("MAGECK_filter_summary_", filter_label, ".csv")),
    row.names = FALSE
  )

  return(list(
    filter_name = filter_label,
    selected_samples = selected_samples,
    group_sub = group_sub,
    design_sub = design_sub,
    keep_idx = keep_idx,
    stats = stats_df,
    keep_table = keep_table,
    filtered_raw_df = filtered_raw_df
  ))
}
```

```{r}
# =========================================================
# 5. Run filtering strategy 1:
#    library_input + T0
# =========================================================
res_library_T0 <- run_voom_filter(
  raw_counts_mat = raw_counts_mat,
  meta_info = meta_info,
  sample_group_df = sample_group_df,
  keep_groups = c("library_input", "T0"),
  filter_label = "voom_libraryInput_plus_T0",
  outdir = outdir,
  mean_logcpm_cutoff = 2.5
)
```

```{r}
# =========================================================
# 6. Run filtering strategy 2:
#    T0 only
# =========================================================
res_T0_only <- run_voom_filter(
  raw_counts_mat = raw_counts_mat,
  meta_info = meta_info,
  sample_group_df = sample_group_df,
  keep_groups = c("T0"),
  filter_label = "voom_T0_only",
  outdir = outdir,
  mean_logcpm_cutoff = 2.5
)
```

```{r}
# =========================================================
# 7. Run filtering strategy 3:
#    T0 + control
#    kept as before for reproducibility
# =========================================================
res_T0_control <- run_voom_filter(
  raw_counts_mat = raw_counts_mat,
  meta_info = meta_info,
  sample_group_df = sample_group_df,
  keep_groups = c("T0", "control"),
  filter_label = "voom_T0_plus_control",
  outdir = outdir,
  mean_logcpm_cutoff = 2.5
)
```

```{r}
# =========================================================
# 8. Collect all filter summaries
# =========================================================
all_stats <- list(
  res_library_T0,
  res_T0_only,
  res_T0_control
) %>%
  purrr::compact() %>%
  purrr::map("stats") %>%
  bind_rows()

if (nrow(all_stats) > 0) {
  write.csv(
    all_stats,
    file.path(outdir, "MAGECK_filtering_overview.csv"),
    row.names = FALSE
  )
}
```

```{r}
# =========================================================
# 9. Console summary
# =========================================================
cat("MAGeCK-oriented filtering complete.\n")
cat("Input file:", input_file, "\n")
cat("Total sgRNAs before filtering:", nrow(raw_counts_mat), "\n")
cat("Outputs written to:", outdir, "\n\n")

if (!is.null(res_library_T0)) {
  cat("Strategy 1: library_input + T0\n")
  print(res_library_T0$stats)
  cat("\n")
}

if (!is.null(res_T0_only)) {
  cat("Strategy 2: T0 only\n")
  print(res_T0_only$stats)
  cat("\n")
}

if (!is.null(res_T0_control)) {
  cat("Strategy 3: T0 + control\n")
  print(res_T0_control$stats)
  cat("\n")
}
```
