
```{r}
library(readr)
library(dplyr)
library(stringr)
library(pracma)
library(lme4)
library(lmerTest)
library(emmeans)
```

```{r}
# =======================
# INPUT / OUTPUT
# =======================
file <- "input_files/growth_tecan_BXPC3.csv"
out_dir <- "output/growth_assays/"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r}
# =======================
# READ DATA
# Required columns:
#   Experiment
#   Replicate
#   Draxin
#   Hr
#   Count
# =======================
df <- read_csv(file, show_col_types = FALSE) %>%
  mutate(
    Experiment = factor(Experiment),
    Replicate  = factor(Replicate),
    Draxin     = str_trim(as.character(Draxin)),
    Hr         = as.numeric(Hr),
    Count      = as.numeric(Count)
  ) %>%
  filter(!is.na(Experiment), !is.na(Replicate), !is.na(Draxin), !is.na(Hr), !is.na(Count))

# Optional manual order
draxin_order <- c("0", "0.55", "8", "17", "34", "68", "140")
draxin_order_present <- draxin_order[draxin_order %in% unique(df$Draxin)]

if (length(draxin_order_present) == 0) {
  draxin_order_present <- sort(unique(df$Draxin))
}

df <- df %>%
  mutate(Draxin = factor(Draxin, levels = draxin_order_present))

# =======================
# CHECK BASELINE
# =======================
baseline_check <- df %>%
  group_by(Experiment, Replicate, Draxin) %>%
  summarise(
    n_timepoints = n(),
    baseline_hr = min(Hr, na.rm = TRUE),
    baseline_count = Count[which.min(Hr)],
    .groups = "drop"
  )

write.csv(
  baseline_check,
  file.path(out_dir, "baseline_check.csv"),
  row.names = FALSE
)
```

```{r}
# =======================
# NORMALIZE TO BASELINE
# within each Experiment × Replicate × Draxin
# =======================
df_norm <- df %>%
  arrange(Experiment, Replicate, Draxin, Hr) %>%
  group_by(Experiment, Replicate, Draxin) %>%
  mutate(
    baseline_hr = min(Hr, na.rm = TRUE),
    baseline_count = Count[which.min(Hr)],
    Count_norm = Count / baseline_count
  ) %>%
  ungroup()

write.csv(
  df_norm,
  file.path(out_dir, "normalized_timecourse.csv"),
  row.names = FALSE
)
```

```{r}
# =======================
# CALCULATE AUC
# One AUC per replicate curve
# =======================
df_auc <- df_norm %>%
  arrange(Experiment, Replicate, Draxin, Hr) %>%
  group_by(Experiment, Replicate, Draxin) %>%
  summarise(
    baseline_hr = first(baseline_hr),
    AUC_norm = trapz(Hr, Count_norm),
    .groups = "drop"
  )

write.csv(
  df_auc,
  file.path(out_dir, "auc_normalized_by_replicate.csv"),
  row.names = FALSE
)

# =======================
# MIXED MODEL
# =======================
m_auc <- lmer(AUC_norm ~ Draxin + (1 | Experiment), data = df_auc)

anova_auc <- anova(m_auc)

write.csv(
  as.data.frame(anova_auc),
  file.path(out_dir, "auc_mixed_model_anova.csv"),
  row.names = TRUE
)

# =======================
# POST HOC: each Draxin vs control
# =======================
emm_auc <- emmeans(m_auc, ~ Draxin)

posthoc_auc <- contrast(
  emm_auc,
  method = "trt.vs.ctrl",
  ref = 1,
  adjust = "dunnett"
)

write.csv(
  as.data.frame(emm_auc),
  file.path(out_dir, "auc_emmeans.csv"),
  row.names = FALSE
)

write.csv(
  as.data.frame(posthoc_auc),
  file.path(out_dir, "auc_draxin_vs_ctrl_dunnett.csv"),
  row.names = FALSE
)

print("Done.")