# General info about the code #
This script contains the NMF visualization workflow adapted from Konstantina’s original code for figure generation in the manuscript. The modifications mainly concern plot layout, ordering of components and sample groups, color handling, and text sizing to improve readability and consistency in publication figures.
Figures generated by this script
This script generates the following outputs:
NMF factor representation across tissue blocks for all components and for a selected subset of components. (Figure 1E vs 2026_March)
Mean NMF activity per spot across tissue blocks, shown separately for selected tumour-associated and nerve-associated factors.(Figure S2B vs 2026_March)
Gene loading plots for all NMF factors and for a selected subset of factors used in the main figure (Figure S1 vs 2026_March)
Export tables of top-ranked genes for each NMF factor, for downstream interpretation and supplementary material (TableS2)


#This section loads all required R packages, imports shared helper functions, defines the color palettes used throughout the plotting workflow, and reads the processed spatial object (spe). It sets up the plotting environment and ensures that barcode identifiers are used as column names, which is required for matching NMF exposures back to sample metadata in later steps.

```{r}
# ============================================================
# 1. Setup: packages, colors, and input data
# ============================================================

library(patchwork)
library(SpatialExperiment)
library(scuttle)
library(batchelor)
library(scran)
library(scater)
library(bluster)
library(ggspavis)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RcppML)
library(stringr)
library(ggrepel)
theme_set(theme_minimal(base_size=14))
library(irlba)
library(coop)
source("../scripts/general_functions.R")

c25 <- c(
  "dodgerblue2", "#E31A1C", # red
"green4",
"#6A3D9A", # purple
"#FF7F00", # orange
"black", "gold1",
"skyblue2", "#FB9A99", # lt pink
"palegreen2",
"#CAB2D6", # lt purple
"#FDBF6F", # lt orange
"gray70", "khaki2",
"maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
"darkturquoise", "green1", "yellow4", "yellow3",
"darkorange4", "brown"
)

color_maping <- c( "Acinar" = "gold1",
                   "B_cells" = "maroon",
                   "Lymphocytes" = "green4",
                   "Endocrine" = "yellow4",
                   "Myeloid(Macrophages)" = "#6A3D9A",
                   "Endothelial" = "#FF7F00",
                   "CAF" = "skyblue2",
                   "Adipocytes" = "steelblue4",
                   "Ductal_normal" = "black",
                   "Myeloid(C1q_macrophages)" = "#CAB2D6",
                   "Nerve" = "#FDBF6F",
                   "Muscle" = "orchid1",
                   "Ductal_malignant" = "gray70")

spe <- readRDS("../data/SPE/PNI_SPE_QC.rds")
colnames(spe) <- spe$barcode
```

# This section defines the helper functions used throughout the script for data wrangling and figure generation. These functions prepare NMF output tables, standardize group labels and colors, and generate reusable plotting panels for NMF gene loadings and spot composition summaries.

```{r}
# ============================================================
# 2. Helper functions for NMF table building and plotting
# ============================================================

lighten_hex <- function(col, amount = 0.55) {
  rgb <- grDevices::col2rgb(col)
  rgb2 <- rgb + (255 - rgb) * amount
  grDevices::rgb(rgb2[1,]/255, rgb2[2,]/255, rgb2[3,]/255)
}

build_nmf_tables <- function(out, spe, rename_map,
                             patient_cols = c("P1"="#E31A1C","P2"="green4","P3"="dodgerblue2")) {

  # --- W + H (scale-aware) ---
  w <- out@w
  colnames(w) <- sub("^nmf", "NMF_", colnames(w))

  H <- out@h
  H <- sweep(H, 1, out@d, `*`)
  rn <- rownames(H)
  rn <- sub("^nmf(\\d+)$", "NMF_\\1", rn, perl = TRUE)
  rownames(H) <- rn

  # --- map each spot/cell -> sample_id ---
  # IMPORTANT: assumes rownames(t(H)) match spe$barcode
  df <- as.data.frame(t(H))
  df$barcode <- rownames(df)

  map_barcode_to_sample <- tibble(
    barcode   = spe$barcode,
    sample_id = spe$sample_id
  )

  df <- df %>%
    left_join(map_barcode_to_sample, by = "barcode")

  # --- rename sample_id to your nice group labels ---
  df <- df %>%
    mutate(group = unname(rename_map[as.character(sample_id)]))

  # --- long exposures summed per (component, group) ---
  long_sum <- df %>%
    select(-barcode, -sample_id) %>%
    pivot_longer(-group, names_to = "component", values_to = "value") %>%
    group_by(component, group) %>%
    summarise(value = sum(value, na.rm = TRUE), .groups = "drop")

  # --- proportions per component ---
  long_prop <- long_sum %>%
    group_by(component) %>%
    mutate(prop = value / sum(value)) %>%
    ungroup()

  # --- order components by numeric id (descending for coord_flip) ---
  long_prop <- long_prop %>%
    mutate(nmf_num = as.numeric(sub("^NMF_", "", component))) %>%
    arrange(nmf_num) %>%
    mutate(component = factor(component, levels = rev(unique(component))))

  # --- annotate group -> patient/status ---
  group_info <- long_prop %>%
    distinct(group) %>%
    mutate(
      patient = str_extract(group, "Patient\\s*\\d+"),
      patient = str_replace(patient, "Patient\\s*", "P"),
      status  = case_when(
        str_detect(group, regex("adjacent", ignore_case = TRUE)) ~ "Adjacent",
        str_detect(group, regex("diseased", ignore_case = TRUE)) ~ "Diseased",
        TRUE ~ "Other"
      ),
      patient = factor(patient, levels = c("P1","P2","P3")),
      status  = factor(status, levels = c("Diseased","Adjacent","Other"))
    )

  # --- group fill colors: strong for diseased, light for adjacent ---
  group_fill_vals <- setNames(
    vapply(seq_len(nrow(group_info)), function(i) {
      base <- patient_cols[as.character(group_info$patient[i])]
      if (is.na(base)) base <- "grey60"
      if (as.character(group_info$status[i]) == "Adjacent") lighten_hex(base, 0.55) else base
    }, FUN.VALUE = character(1)),
    group_info$group
  )

  # --- spot counts per renamed group (for mean-per-spot plots) ---
  spot_counts <- as.data.frame(table(spe$sample_id))
  colnames(spot_counts) <- c("sample_id", "n_spots")

  map_df <- tibble(
    sample_id = names(rename_map),
    group     = unname(rename_map)
  )

  spot_counts_group <- map_df %>%
    left_join(spot_counts, by = "sample_id") %>%
    select(group, n_spots)

  # --- long_abs: add n_spots and per-spot activity ---
  long_abs <- long_sum %>%
    left_join(spot_counts_group, by = "group") %>%
    mutate(value_per_spot = value / n_spots) %>%
    left_join(group_info, by = "group")

  list(
    w = w,
    H = H,
    long_prop = long_prop,
    long_abs  = long_abs,
    group_info = group_info,
    group_fill_vals = group_fill_vals
  )
}

# Robust loadings
plotLoadings <- function(amp, n = 15,
                         gene_size = 7,
                         title_size = 8,
                         ncol = 5) {

  # Ensure genes are rows, factors are columns
  if (nrow(amp) < ncol(amp)) amp <- t(amp)

  if (is.null(colnames(amp))) colnames(amp) <- paste0("Factor ", seq_len(ncol(amp)))
  if (is.null(rownames(amp))) stop("Supply gene names as rownames(amp).")

  K <- ncol(amp)
  if (is.null(ncol)) ncol <- min(5, K)
  nrow_grid <- ceiling(K / ncol)

  plist <- vector("list", K)

  for (k in seq_len(K)) {
    nm  <- colnames(amp)[k]
    vmin <- min(amp[, k], na.rm = TRUE)
vmax <- max(amp[, k], na.rm = TRUE)
if (isTRUE(all.equal(vmin, vmax))) {
  sc <- rep(0, nrow(amp))
  names(sc) <- rownames(amp)
} else {
  sc <- (amp[, k] - vmin) / (vmax - vmin)
}
    ord <- sort(sc, decreasing = TRUE)
    ord <- head(ord, n)

    fp <- data.frame(
      Gene  = factor(names(ord), levels = rev(names(ord))),
      Value = as.numeric(ord)
    )

    plist[[k]] <- ggplot(fp, aes(x = Gene, y = Value)) +
      geom_segment(aes(xend = Gene, y = 0, yend = Value), linewidth = 0.4) +
      geom_point(size = 3.6, pch = 21, fill = "steelblue", color = "grey20") +
      coord_flip() +
      scale_y_continuous(limits = c(0, 1.05), breaks = c(0, 0.5, 1)) +
      labs(title = nm, x = NULL, y = NULL) +
      theme_minimal(base_size = gene_size) +
      theme(
        legend.position = "none",

        # font sizes as you asked
        plot.title  = element_text(size = title_size, face = "bold", hjust = 0.5),
        axis.text.y = element_text(size = gene_size),
        axis.text.x = element_text(size = gene_size),

        # grid: show value grid only (after coord_flip, this is major.x)
        panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank(),
        panel.grid.major.x = element_line(color = "grey85", linewidth = 0.3),

        plot.margin = margin(6, 6, 6, 6)
      )
  }

  cowplot::plot_grid(plotlist = plist, ncol = ncol, nrow = nrow_grid)
}

plotLoadings_A4 <- function(amp, n = 15,
                            gene_size = 7,
                            title_size = 8,
                            ncol = 5,
                            panel_spacing = 1.2) {

  if (nrow(amp) < ncol(amp)) amp <- t(amp)

  if (is.null(colnames(amp))) colnames(amp) <- paste0("Factor ", seq_len(ncol(amp)))
  if (is.null(rownames(amp))) stop("Supply gene names as rownames(amp).")

  K <- ncol(amp)
  nrow_grid <- ceiling(K / ncol)

  plist <- vector("list", K)

  for (k in seq_len(K)) {
    nm <- colnames(amp)[k]

    # robust 0..1 scaling (avoids NaN if constant)
    vmin <- min(amp[, k], na.rm = TRUE)
    vmax <- max(amp[, k], na.rm = TRUE)
    if (isTRUE(all.equal(vmin, vmax))) {
      sc <- rep(0, nrow(amp))
      names(sc) <- rownames(amp)
    } else {
      sc <- (amp[, k] - vmin) / (vmax - vmin)
    }

    ord <- sort(sc, decreasing = TRUE)
    ord <- head(ord, n)

    df <- data.frame(
      Gene  = factor(names(ord), levels = rev(names(ord))),
      Value = as.numeric(ord)
    )

    plist[[k]] <- ggplot(df, aes(x = Gene, y = Value)) +
      geom_segment(aes(xend = Gene, y = 0, yend = Value), linewidth = 0.35) +
      geom_point(size = 3.6, shape = 21, fill = "steelblue", color = "black") +
      coord_flip() +
      scale_y_continuous(
        limits = c(0, 1.05),
        breaks = c(0, 0.5, 1),
        expand = expansion(mult = c(0, 0.02))
      ) +
      labs(title = nm) +
      theme_minimal(base_size = gene_size) +
      theme(
        plot.title = element_text(size = title_size, face = "bold", hjust = 0.5),
        axis.text.y = element_text(size = gene_size),
        axis.text.x = element_text(size = gene_size),
        axis.title  = element_blank(),

        # ---- GRID CONTROL ----
        panel.grid.major.y = element_blank(),  # no grid on gene axis
        panel.grid.minor   = element_blank(),
        panel.grid.major.x = element_line(color = "grey85", linewidth = 0.3),

        plot.margin = margin(5, 5, 5, 5)
      )
  }  

  cowplot::plot_grid(
    plotlist = plist,
    ncol = ncol,
    nrow = nrow_grid,
    rel_heights = rep(panel_spacing, nrow_grid)
  )
}

plot_spot_composition_2x2 <- function(spe,
                                      panel_var,        # e.g. "NerveClass4" (invaded/uninvaded/proximal/distal)
                                      tissue_var,       # e.g. "Tissue" (Diseased/Adjacent)
                                      patient_var,      # e.g. "patient_id" or "Patient"
                                      nerve_id_var,     # e.g. "NerveID"
                                      panels = c("Invaded", "Uninvaded", "Tumor proximal", "Tumor distal"),
                                      tissue_levels = c("Diseased","Adjacent"),
                                      patient_cols = c("P1"="#E31A1C","P2"="green4","P3"="dodgerblue2"),
                                      base_size = 13) {

  df <- as.data.frame(colData(spe))

  # keep only rows with required columns non-missing
  need <- c(panel_var, tissue_var, patient_var, nerve_id_var)
  stopifnot(all(need %in% colnames(df)))

  df <- df %>%
    dplyr::filter(
      !is.na(.data[[panel_var]]),
      !is.na(.data[[tissue_var]]),
      !is.na(.data[[patient_var]])
    )

  # Standardize patient labels to P1/P2/P3 if needed (safe no-op if already)
  df[[patient_var]] <- as.character(df[[patient_var]])
  df[[patient_var]] <- stringr::str_replace(df[[patient_var]], "Patient\\s*", "P")

  # factor order
  df[[panel_var]]  <- factor(df[[panel_var]], levels = panels)
  df[[tissue_var]] <- factor(df[[tissue_var]], levels = tissue_levels)
  df[[patient_var]] <- factor(df[[patient_var]], levels = names(patient_cols))

  # spot counts per (panel, tissue, patient)
  spot_tab <- df %>%
    dplyr::count(
      panel = .data[[panel_var]],
      tissue = .data[[tissue_var]],
      patient = .data[[patient_var]],
      name = "n_spots"
    ) %>%
    dplyr::filter(!is.na(panel), !is.na(tissue), !is.na(patient))

  # nerve counts per panel (unique NerveID), regardless of tissue/patient
  nerve_tab <- df %>%
    dplyr::filter(!is.na(.data[[nerve_id_var]])) %>%
    dplyr::mutate(panel = .data[[panel_var]]) %>%
    dplyr::group_by(panel) %>%
    dplyr::summarise(n_nerves = dplyr::n_distinct(.data[[nerve_id_var]]), .groups = "drop")

  # label position: top of each facet, slightly above max
  y_max <- spot_tab %>%
    dplyr::group_by(panel) %>%
    dplyr::summarise(y = max(n_spots, na.rm = TRUE), .groups = "drop") %>%
    dplyr::left_join(nerve_tab, by = "panel") %>%
    dplyr::mutate(
      y = ifelse(is.finite(y), y * 1.08, 0),
      label = paste0("Nerves: ", n_nerves)
    )

  ggplot(spot_tab, aes(x = tissue, y = n_spots, fill = patient)) +
    geom_col(position = position_stack(), width = 0.78, color = NA) +
    facet_wrap(~ panel, ncol = 2, scales = "free_y") +
    scale_fill_manual(values = patient_cols, drop = FALSE) +
    geom_text(
      data = y_max,
      aes(x = tissue_levels[2], y = y, label = label),  # place on right bar (Adjacent) by default
      inherit.aes = FALSE,
      hjust = 1,
      vjust = 0,
      size = 3.8
    ) +
    labs(x = NULL, y = "Number of spots", fill = NULL) +
    theme_minimal(base_size = base_size) +
    theme(
      panel.grid.major.x = element_blank(),
      strip.text = element_text(face = "bold"),
      legend.position = "right"
    )
}

```

#This section loads the fitted NMF model, extracts the component loading and exposure matrices, and reformats component names into a manuscript-friendly NMF_x naming scheme. It then maps each spot to its sample of origin, applies the human-readable block labels used in the paper, and constructs the summarized data tables needed for all subsequent NMF representation and activity plots.

```{r}

# ============================================================
# 3. Load NMF output and build summary tables
# ============================================================

out <- readRDS("../data/coldata/PNI_NMF_L105-k30.rds")
w30 <- out@w
colnames(w30) <- paste0(colnames(w30),"_","K30")

H <- out@h
H <- sweep(H, 1, out@d, `*`)   # scale-aware exposures

comp_names <- rownames(H)
comp_names <- sub("^nmf(\\d+)$", "NMF_\\1", comp_names, perl = TRUE)
rownames(H) <- comp_names

df <- as.data.frame(t(H))
df$group <- plyr::mapvalues(rownames(df), from = spe$barcode, to=spe$sample_id)

# tidy and normalize to proportions within each component
long <- df |>
  pivot_longer(-group, names_to = "component", values_to = "value") |>
  group_by(component, group) |>
  summarise(value = sum(value), .groups = "drop")

long <- long |>
  group_by(component) |>
  mutate(prop = value / sum(value)) |>
  ungroup()

rename_map <- c("70865bl15_highPNI" = "Patient 3 diseased",
               "6249bl18_noPNIandhighPNI" = "Patient 2 diseased",
               "70865bl5_noPNI" = "Patient 3 adjacent",
               "54703bl38_noPNI" = "Patient 1 adjacent",
               "54703bl28_highPNI" = "Patient 1 diseased",
               "6249bl23_noPNI" = "Patient 2 adjacent")

long$group <- rename_map[ long$group ]
long$nmf_num <- as.numeric(sub("^NMF_", "", long$component))
long <- long[order(long$nmf_num), ]
ordered_levels <- unique(long$component)
ordered_levels <- rev(ordered_levels)
long$component <- factor(long$component, levels = ordered_levels)

res <- build_nmf_tables(out, spe, rename_map)
# res$long_prop, res$long_abs, res$group_fill_vals, res$w ...
```

#This section defines and generates a stacked bar plot showing the proportional representation of every NMF factor across tissue blocks. The plot is intended as a global overview of how each factor is distributed across samples and disease contexts, using patient- and tissue-status–aware color coding.

```{r}
# ============================================================
# 4. Plot NMF factor representation across tissue blocks
# ============================================================

plot_representation_all <- function(long_prop, group_fill_vals, base_size = 14) {
  ggplot(long_prop, aes(x = component, y = prop, fill = group)) +
    geom_col(width = 0.8) +
    coord_flip() +
    scale_y_continuous(limits = c(0, 1.01), breaks = seq(0, 1, 0.25)) +
    scale_fill_manual(values = group_fill_vals) +
    labs(title = "Representation in tissue block", x = NULL, y = NULL) +
    theme_minimal(base_size = base_size) +
    theme(
      axis.ticks.y = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.title = element_blank(),
      plot.title = element_text(hjust = 0.5)
    )
}

p1 <- plot_representation_all(res$long_prop, res$group_fill_vals, base_size = 14)
p1
```

#This section focuses on a selected subset of NMF factors that are most relevant for the manuscript. It generates a stacked representation plot with a controlled ordering of blocks in the stack, so that diseased and adjacent samples, as well as patient identities, appear in a consistent and interpretable order across factors.

```{r}
# ============================================================
# 4B. Plot selected NMF factor representation across tissue blocks
# ============================================================


plot_representation_keep_stack <- function(long_prop, keep,
                                          group_fill_vals,
                                          base_size = 22,
                                          order_stack = c("Diseased","Adjacent","Other"),
                                          order_patient = c("P1","P2","P3")) {

  dat <- long_prop %>%
    filter(component %in% keep) %>%
    mutate(component = factor(component, levels = keep)) %>%
    left_join(
      long_prop %>% distinct(group) %>%
        mutate(
          patient = str_replace(str_extract(group, "Patient\\s*\\d+"), "Patient\\s*", "P"),
          status  = case_when(
            str_detect(group, regex("adjacent", ignore_case = TRUE)) ~ "Adjacent",
            str_detect(group, regex("diseased", ignore_case = TRUE)) ~ "Diseased",
            TRUE ~ "Other"
          ),
          patient = factor(patient, levels = order_patient),
          status  = factor(status, levels = order_stack)
        ),
      by = "group"
    )

  # compute stack order
  group_levels_stack <- dat %>%
    distinct(group, patient, status) %>%
    mutate(
      status_rank  = as.integer(status),
      patient_rank = as.integer(patient),
      stack_rank = status_rank * 10L + patient_rank
    ) %>%
    arrange(stack_rank, group) %>%
    pull(group)

  dat$group <- factor(dat$group, levels = group_levels_stack)

  ggplot(dat, aes(x = component, y = prop, fill = group)) +
    geom_col(width = 0.8, position = position_stack(reverse = TRUE)) +
    coord_flip() +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
    scale_fill_manual(values = group_fill_vals) +
    labs(title = "Representation in tissue block", x = NULL, y = NULL) +
    theme_minimal(base_size = base_size) +
    theme(
      axis.ticks.y = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.title = element_blank(),
      plot.title = element_text(hjust = 0.5)
    )
}

keep5 <- c("NMF_27","NMF_19","NMF_14","NMF_13","NMF_4")
p2_vs2 <- plot_representation_keep_stack(res$long_prop, keep5, res$group_fill_vals, base_size = 22)
p2_vs2
```
#This section plots absolute NMF activity normalized per spot for the selected factors. Tumour-associated and nerve-associated factors are displayed in separate panels to improve interpretability and to avoid compressing the scale of the smaller nerve-related component by the larger tumour-related signals.

```{r}

# ============================================================
# 5. Plot mean NMF activity per spot for selected factors
# ============================================================

plot_abs_split <- function(long_abs, keep, nerve_keep,
                           comp_cols,
                           base_size = 14) {

  dat <- long_abs %>%
    filter(component %in% keep) %>%
    mutate(component = factor(component, levels = keep))

  tumor_keep <- setdiff(keep, nerve_keep)

  p_tumor <- dat %>%
    filter(component %in% tumor_keep) %>%
    ggplot(aes(x = group, y = value_per_spot, fill = component)) +
    geom_col(width = 0.85) +
    facet_grid(~ status, scales = "free_x", space = "free_x") +
    scale_fill_manual(values = comp_cols) +
    labs(title = "NMF component activity per block (mean per spot)",
         x = NULL, y = "Mean NMF activity per spot") +
    theme_minimal(base_size = base_size) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      panel.grid.major.x = element_blank(),
      strip.text = element_text(size = base_size),
      legend.title = element_blank()
    )

  p_nerve <- dat %>%
    filter(component %in% nerve_keep) %>%
    ggplot(aes(x = group, y = value_per_spot, fill = component)) +
    geom_col(width = 0.85) +
    facet_grid(~ status, scales = "free_x", space = "free_x") +
    scale_fill_manual(values = comp_cols) +
    labs(title = NULL, x = NULL, y = "Mean NMF activity per spot") +
    theme_minimal(base_size = base_size) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      panel.grid.major.x = element_blank(),
      strip.text = element_text(size = base_size),
      legend.title = element_blank()
    )

  p_tumor / p_nerve + patchwork::plot_layout(heights = c(2, 1))
}

comp_cols <- c(
  "NMF_4"  = "#1b9e77",
  "NMF_13" = "#d95f02",
  "NMF_14" = "#7570b3",
  "NMF_19" = "#e7298a",
  "NMF_27" = "#66a61e"
)

p_abs <- plot_abs_split(res$long_abs, keep5, nerve_keep = c("NMF_27"), comp_cols = comp_cols, base_size = 14)
p_abs
```
#This section generates publication-ready plots of the top gene loadings for each NMF factor. It first uses the standardized W matrix returned by the helper function, then produces a full A4 supplementary panel showing all factors and a second plot restricted to the subset of factors highlighted in the main figure. The same section can also be used to export the full loading matrix if needed.

```{r}
# ============================================================
# 6. Plot top gene loadings for NMF factors
# ============================================================

# after:
# res <- build_nmf_tables(out, spe, rename_map)

w <- res$w  # <- always take W from the builder

# 1) Export full W
#write.csv(w, "../data/coldata/NMF_genes.csv")

# 2) A4 panel of ALL factors
pA4 <- plotLoadings_A4(
  w,
  n = 15,
  gene_size = 7,
  title_size = 8,
  ncol = 5
)

print(pA4)
ggsave("../figures/S1_A_A4.pdf",
       pA4, width = 8.27, height = 11.69, units = "in")

# 3) A4 landscape panel for a subset (keep5)
keep5 <- c("NMF_4", "NMF_13", "NMF_14", "NMF_19", "NMF_27")

p4 <- plotLoadings(
  w[, keep5, drop = FALSE],
  n = 15,
  gene_size = 7,
  title_size = 8,
  ncol = length(keep5)
)

print(p4)
ggsave("../figures/F1_C.pdf",
       p4, width = 11.69, height = 8.27, units = "in")  # A4 landscape

# 4) Optional single factor
# p5 <- plotLoadings(w[, "NMF_27", drop = FALSE], n = 15, gene_size = 10, title_size = 11, ncol = 1)
# ggsave("../figures/F1_D.pdf", p5, width = 4, height = 4)
# print(p5)
```
# This section defines a helper function to export the top-ranked genes for each NMF factor in a long-format table suitable for downstream interpretation, supplementary tables, or manual annotation. For each factor, genes are ranked by loading value, and optionally rescaled to a 0–1 range to facilitate comparison of relative within-factor importance.

```{r}
# ============================================================
# 7. Export top-ranked genes per NMF factor
# ============================================================

export_top_genes_long <- function(w,
                                  top_n = 200,
                                  file = "../tables/NMF_top200_genes_long.csv",
                                  scale01 = TRUE) {
  stopifnot(!is.null(rownames(w)), !is.null(colnames(w)))

  # Make sure columns look like NMF_1, NMF_2, ...
  # (skip if you already have this)
  colnames(w) <- sub("^nmf", "NMF_", colnames(w))

  nmfs <- colnames(w)

  out_list <- lapply(nmfs, function(k) {
    v <- w[, k]
    v <- v[is.finite(v)]
    v <- v[order(v, decreasing = TRUE)]
    v <- head(v, top_n)

    df <- tibble::tibble(
      NMF = k,
      Rank = seq_along(v),
      Gene = names(v),
      Loading = as.numeric(v)
    )

    if (scale01) {
      lo <- min(df$Loading, na.rm = TRUE)
      hi <- max(df$Loading, na.rm = TRUE)
      df <- df %>%
        dplyr::mutate(Loading_scaled01 = ifelse(isTRUE(all.equal(lo, hi)), 0, (Loading - lo) / (hi - lo)))
    }
    df
  })

  tab <- dplyr::bind_rows(out_list)

  # Ensure output dir exists
  dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)

  readr::write_csv(tab, file)
  message("Wrote: ", normalizePath(file, mustWork = FALSE))
  tab
}

# ============================================================
# 1) Export top-200 genes per NMF (publication-style long table)
# ============================================================
w <- res$w

top200_long <- export_top_genes_long(
  w,
  top_n = 200,
  file = "../tables/NMF_top200_genes_long.csv",
  scale01 = TRUE
)

# optional quick peek
dplyr::count(top200_long, NMF) %>% print(n = 50)

```
