# Identification of Nerves
==================================================
With the NMF factors we can now call nerves and group them into invaded and non-invaded.
This code generates figures that explain spatial identificaiton of four tissue regions i.e. invaded, non-invaded and nerve:distal and nerve:proximal tumor spots. Figure S3 as for 2026_March 
This code also contains elements used for Figure 1A as for 2026_March

It also generates stacked barplots to depict NMF composition of invaded and non-invaded nerves, individual nerve level 

It also generates a plot that depicts NMF tumor contribution in the invaded and non-invaded nerve borders Each point represents an individual spot in direct contact with a nerve, grouped by the nerve it borders (BorderID). The + symbol marks the mean tumor signal across all border spots associated with each nerve. Facets separate spots bordering invaded versus non-invaded nerves.

## Load libraries, helper functions, color palette, and input objects
This chunk loads required packages, defines helper functions and colors, and imports the integrated spatial object together with the NMF component table.


```{r}
library(patchwork)
library(SpatialExperiment)
library(scuttle)
library(batchelor)
library(scran)
library(scater)
library(bluster)
library(ggspavis)
library(dplyr)
library(RcppML)
library(ggrepel)
library(BayesSpace)
library(RANN)
library(plyr)
library(dplyr)
theme_set(theme_minimal(base_size=14))
library(irlba)
source("../scripts/general_functions.R")
# --------------

minmaxscale <- function(x, ...) {
    return((x - min(x, ...)) /(max(x, ...) - min(x, ...)))
}
c25 <- c(
  "dodgerblue2", "#E31A1C", # red
"green4",
"#6A3D9A", # purple
"#FF7F00", # orange
"black", "gold1",
"skyblue2", "#FB9A99", # lt pink
"palegreen2",
"#CAB2D6", # lt purple
"#FDBF6F", # lt orange
"gray70", "khaki2",
"maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
"darkturquoise", "green1", "yellow4", "yellow3",
"darkorange4", "brown"
)

spe <- readRDS("../data/SPE/Integrated_NMF.rds")
colnames(spe) <- spe$barcode

compk30 <- DataFrame(read.csv("../data/coldata/NMFcomp.csv"))
names(compk30)[names(compk30) == "Myeloid.Macrophages."] <- "Myeloid(Macrophages)"
names(compk30)[names(compk30) == "Myeloid.C1q_macrophages."] <- "Myeloid(C1q_macrophages)"
rownames(compk30) <- compk30$barcode
```

**Nerve Signal**
#This chunk extracts the nerve, tumor, and C1q macrophage NMF factors and visualizes the nerve signal across samples. Used for Figure 1A workflow and Figure S3.
```{r}
# ========================
# Nerve NMF signal plotting
# ========================

comp <- as.matrix(t(reducedDims(spe)[["compk30"]]))
spe$nerve <- comp['Nerve',]
spe$tumor <- comp['Ductal_malignant',]
spe$C1qM <- comp['Myeloid(C1q_macrophages)',]
nerve_sginal <- ggcells(spe, aes(x=array_col, y=-array_row, color=nerve)) +
  geom_point(size=0.4) +
  facet_wrap(~sample_id,
             labeller = as_labeller(c(
               "70865bl15_highPNI" = "Patient 3 - Diseased",
               "6249bl18_noPNIandhighPNI" = "Patient 2 - Diseased",
               "70865bl5_noPNI" = "Patient 3 - Adjacent",
               "54703bl38_noPNI" = "Patient 1 - Adjacent",
               "54703bl28_highPNI" = "Patient 1 - Diseased",
               "6249bl23_noPNI" = "Patient 2 - Adjacent"
             ))) +
  theme_void()+
  scale_color_viridis_b()

nerve_sginal
#ggsave("nerve_signal_figure1A.pdf", nerve_sginal, height= 10, width=14, dpi=600)
```

```{r}
library(ggplot2)
library(ggspavis)
library(grid)

# ========================
# Nerve NMF signal plotting
# with legend title + approximate scale bar
# ========================

comp <- as.matrix(t(reducedDims(spe)[["compk30"]]))
spe$nerve <- comp["Nerve", ]

# Approximate conversion on the array grid:
# one Visium spot diameter ~55 µm
# here we represent that visually as ~1 grid unit
scale_bar_len <- 1   # one spot-width on the array grid
scale_bar_lab <- "550 µm"

# choose a manual location for the scale bar
x_start <- min(spe$array_col, na.rm = TRUE) + 5
x_end   <- x_start + scale_bar_len
y_bar   <- -max(spe$array_row, na.rm = TRUE) + 5

nerve_signal <- ggcells(
  spe,
  aes(x = array_col, y = -array_row, color = nerve)
) +
  geom_point(size = 0.4) +
  facet_wrap(
    ~sample_id,
    labeller = as_labeller(c(
      "70865bl15_highPNI" = "Patient 3 - Diseased",
      "6249bl18_noPNIandhighPNI" = "Patient 2 - Diseased",
      "70865bl5_noPNI" = "Patient 3 - Adjacent",
      "54703bl38_noPNI" = "Patient 1 - Adjacent",
      "54703bl28_highPNI" = "Patient 1 - Diseased",
      "6249bl23_noPNI" = "Patient 2 - Adjacent"
    ))
  ) +
  coord_fixed() +
  theme_void() +
  theme(
    strip.text = element_text(size = 12),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  ) +
  scale_color_viridis_c(
    name = "Nerve factor activity (%)",
    limits = c(0, 100),
    breaks = c(0, 20, 40, 60, 80, 100)
  ) +
  guides(
    color = guide_colorbar(
      barheight = unit(50, "mm"),
      frame.colour = "black",
      ticks.colour = "black"
    )
  ) +
  annotate(
    "segment",
    x = x_start, xend = x_end,
    y = y_bar, yend = y_bar,
    linewidth = 0.8,
    colour = "black"
  ) +
  annotate(
    "text",
    x = (x_start + x_end) / 2,
    y = y_bar + 2,
    label = scale_bar_lab,
    size = 3
  )

nerve_signal

# ========================
# Add 1 mm approximate scale bar
# ========================

scale_bar_len <- 10   # ~10 spots ≈ 1 mm
scale_bar_lab <- "1 mm (approx.)"

# position
x_start <- min(spe$array_col, na.rm = TRUE) + 5
x_end   <- x_start + scale_bar_len
y_bar   <- -max(spe$array_row, na.rm = TRUE) + 5

p <- ggcells(
  spe,
  aes(x = array_col, y = -array_row, color = nerve)
) +
  geom_point(size = 0.4) +
  facet_wrap(~sample_id) +
  coord_fixed() +
  theme_void() +
  scale_color_viridis_c(
    name = "Nerve factor activity (%)",
    limits = c(0, 100)
  ) +

  # scale bar
  annotate(
    "segment",
    x = x_start, xend = x_end,
    y = y_bar, yend = y_bar,
    linewidth = 1,
    colour = "black"
  ) +
  annotate(
    "text",
    x = (x_start + x_end) / 2,
    y = y_bar + 2,
    label = scale_bar_lab,
    size = 3
  )

p
```

**C1q macrophages signal**
#This chunk visualizes the spatial distribution of the C1q macrophage NMF factor across samples.

```{r}
# ========================
# C1q macrophage NMF signal plotting
# ========================
p<- ggcells(spe, aes(x=array_col, y=-array_row, color=C1qM)) +
  geom_point(size=0.8) +
  facet_wrap(~sample_id,
             labeller = as_labeller(c(
               "70865bl15_highPNI" = "Patient 3 - Diseased",
               "6249bl18_noPNIandhighPNI" = "Patient 2 - Diseased",
               "70865bl5_noPNI" = "Patient 3 - Adjacent",
               "54703bl38_noPNI" = "Patient 1 - Adjacent",
               "54703bl28_highPNI" = "Patient 1 - Diseased",
               "6249bl23_noPNI" = "Patient 2 - Adjacent"
             ))) +
  theme_void()+ 
  scale_color_viridis_b() + labs(color = "C1q macrophages signal")
#ggsave("../figures/C1q_macrophages_spatial.pdf", p, width = 16, height = 12)
p
```

#Used for figure 1A workflow to depict tumor NMF factor in space 
#This chunk visualizes the spatial distribution of the tumor NMF factor across samples. Used for Figure 1A workflow.

```{r}
# ========================
# Tumor NMF signal plotting
# ========================
p2<- ggcells(spe, aes(x=array_col, y=-array_row, color=tumor)) +
  geom_point(size=0.4) +
  facet_wrap(~sample_id,
             labeller = as_labeller(c(
               "70865bl15_highPNI" = "Patient 3 - Diseased",
               "6249bl18_noPNIandhighPNI" = "Patient 2 - Diseased",
               "70865bl5_noPNI" = "Patient 3 - Adjacent",
               "54703bl38_noPNI" = "Patient 1 - Adjacent",
               "54703bl28_highPNI" = "Patient 1 - Diseased",
               "6249bl23_noPNI" = "Patient 2 - Adjacent"
             ))) +
  theme_void()+ 
  scale_color_viridis_b() + labs(color = "Mean tumor signal")
#ggsave("../figures/tumor_signa_spatial.pdf", p2, width = 16, height = 12)
p2
```

#Thresholding for calling nerves
Nerves are defined as spots with more than 20% nerve signal from the NMF characterization.
Call nerve-positive spots
This chunk applies the nerve threshold and visualizes all spots classified as nerve-positive. Used for Figure S3 workflow.

```{r}
# ========================
# Tresholding for calling nerves (anything that has 20% nerve signal)
# ========================

spe$is.nerve <- spe$nerve > 20
ggcells(spe, aes(x=array_col, y=-array_row, color=is.nerve)) +
  geom_point(size=0.8) +
  facet_wrap(~sample_id,
             labeller = as_labeller(c(
               "70865bl15_highPNI" = "Patient 3 - Diseased",
               "6249bl18_noPNIandhighPNI" = "Patient 2 - Diseased",
               "70865bl5_noPNI" = "Patient 3 - Adjacent",
               "54703bl38_noPNI" = "Patient 1 - Adjacent",
               "54703bl28_highPNI" = "Patient 1 - Diseased",
               "6249bl23_noPNI" = "Patient 2 - Adjacent"
             ))) +
  scale_color_manual(values=c("black","grey50"))
```
#Group into individual nerves
Connecting spots that are directly connected into units of nerves, called NerveID here. We only consider nerves covered by at least 2 spots.
Assign connected nerve spots into nerve units
This chunk shifts coordinates across samples, identifies connected nerve-positive spots, and assigns them into individual NerveID groups.

```{r}
# ======================================
# Group into individual nerves
# ======================================

# Shift Coordinates
spe$col <- spe$array_col
spe$row <- spe$array_row

offset_row <- 150
offset_col <- 250

for (i in seq_along(unique(spe$sample_id))) {
  off_row <- offset_row * i-1
  off_col <- offset_col * i-1
  smp <- unique(spe$sample_id)[i]
  spe$col[spe$sample_id==smp] <- spe$col[spe$sample_id==smp] + off_col
  spe$row[spe$sample_id==smp] <- spe$row[spe$sample_id==smp] + off_row
}

spe_nerve <- spe[,spe$is.nerve]
pos <- cbind(spe_nerve$row,spe_nerve$col)
nn <- BayesSpace:::find_neighbors(pos,radius=2,method="manhattan") # 
dd <- dist(pos,method="manhattan")
tree <- hclust(dd,method="single")
spe_nerve$NerveID <- cutree(tree,h=2)
add <- data.frame(colData(spe_nerve)[,c("barcode","NerveID")])
colData(spe) <- DataFrame(dplyr::left_join(data.frame(colData(spe)),add))

setna <- as.integer(names(which(table(spe$NerveID)<2)))
spe$NerveID[spe$NerveID %in% setna] <- NA
```

This gives us the following IDs.
# This chunk shows the spatial distribution of the identified nerve units. Used for Figure S3 workflow.

```{r}
# ======================================
# Plotting individual nerves
# ======================================

ggcells(spe, aes(x=array_col, y=-array_row, color=factor(NerveID))) +
  geom_point(size=0.8) +
  facet_wrap(~sample_id,
             labeller = as_labeller(c(
               "70865bl15_highPNI" = "Patient 3 - Diseased",
               "6249bl18_noPNIandhighPNI" = "Patient 2 - Diseased",
               "70865bl5_noPNI" = "Patient 3 - Adjacent",
               "54703bl38_noPNI" = "Patient 1 - Adjacent",
               "54703bl28_highPNI" = "Patient 1 - Diseased",
               "6249bl23_noPNI" = "Patient 2 - Adjacent"
             ))) +
  theme_void() +
  theme(legend.position ='none')
```

#We can now threshold on the mean tumor contribution of individual nerves (not spots) to group them into invaded and non-invaded. We set the threshold at 1%. Plot mean tumor contribution per nerve This chunk calculates the mean tumor signal for each nerve and visualizes its distribution to support the invasion threshold. Used for Figure S3B.

```{r}
# ======================================
# Plotting mean tumor contribution per individual nerve
# ======================================

pd <- data.frame(colData(spe))
pd <- pd[!is.na(pd$NerveID),]

pd <- dplyr::group_by(pd, NerveID)
pd <- dplyr::mutate(pd, mnTumor = mean(tumor, na.rm = TRUE))
pd <- dplyr::ungroup(pd)

pd <- pd[!duplicated(pd$NerveID),]

b <- ggplot(pd, aes(x = mnTumor)) +
  geom_histogram(bins = 30,
                 fill = "grey70",
                 color = "black",
                 linewidth = 0.2) +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             linewidth = 0.6) +
  labs(
    x = "Mean Tumor Contribution per Nerve",
    y = "Number of Nerves"
  ) +
  theme_minimal(base_size = 14) +   # controls global scaling
  theme(
    axis.title = element_text(size = 14),
    axis.text  = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.minor = element_blank()
  )

b
```

```{r}
ggsave("../figures/Mean_tumor_contribution.pdf", b, width = 10, height = 10, dpi=300)
```

# Annotate nerves as invaded or non-invaded. This chunk assigns invasion status to each nerve based on the 1% mean tumor threshold and adds the. annotation back to the spatial object.

```{r}
# ======================================
# Adding invasion status to spe
# ======================================

pd$Invaded <- pd$mnTumor > 1
add <- pd[,c("NerveID","mnTumor","Invaded")]
colData(spe) <- DataFrame(dplyr::left_join(data.frame(colData(spe)),add))
spe$Invaded <- ifelse(spe$Invaded,"Invaded","NonInvaded")
```

**Summary plot**
# Summary plot of nerve classes. This chunk overlays invaded and non-invaded nerves on top of the tumor signal background to summarize the region classification workflow. Also used for Figure S3.

```{r}
# ======================================
# Plotting classified nerves + tumor signal
# ======================================

spe_nerve <- spe[,!is.na(spe$NerveID)]
bg <- data.frame(colData(spe)[,c("array_col","array_row","tumor","sample_id")])
ggcells(spe_nerve, aes(x=array_col, y=-array_row)) +
  geom_point(data=bg,size=1,pch=21,aes(fill=tumor,color=NULL),stroke=0.05) +
  geom_point(size=0.5,pch=19,aes(color=Invaded)) +
  facet_wrap(~sample_id,
             labeller = as_labeller(c(
               "70865bl15_highPNI" = "Patient 3 - Diseased",
               "6249bl18_noPNIandhighPNI" = "Patient 2 - Diseased",
               "70865bl5_noPNI" = "Patient 3 - Adjacent",
               "54703bl38_noPNI" = "Patient 1 - Adjacent",
               "54703bl28_highPNI" = "Patient 1 - Diseased",
               "6249bl23_noPNI" = "Patient 2 - Adjacent"
             ))) +
  scale_fill_viridis_c(option="A") +
  scale_color_brewer(palette="Accent") +
  labs(color="")
```

#Identification of border regions
==================================================
#We define the border region as follows:
#Spots with direct contact to a nerve
#Spots with a minimum tumor content of 50%
#Identify spots that are in direct contact with nerves
#This chunk defines a helper function to detect spatial contacts and then identifies all spots directly touching a nerve.

```{r, warning=FALSE, message=FALSE}

getContacts <- function(input,rad,knn,query=rownames(input),graph=FALSE,x="pxl_col_in_fullres",y="pxl_row_in_fullres") {
    require(RANN)
    require(Matrix)
    require(igraph)
    query.data <- input[query,c(x,y)]
    nn <- nn2(data=input[,c(x,y)],query=query.data,k=knn,radius=rad,searchtype="radius")
    if (!graph) {
    nn.ct <- apply(nn[["nn.idx"]], 1, function(ROW) {
        tmp <- sapply(ROW,function(ELEMENT) {
               out <- NA
               if(ELEMENT!=0) {
               out <- input$Cluster[ELEMENT]
               }
               return(out)
                   })
        return(unlist(tmp))
    })

    # Second link back the indicies to the 'celltype ID'
    expected <- unique(input$Cluster)
    out.list <- lapply(1:ncol(nn.ct),function(COL) {
        contacts <- nn.ct[,COL]
        contacts <- contacts[-1] # This removes the cell itself as it is always returned as nearest neighbor
        tbbl <- table("Contact_Celltype"=contacts)
        freq <- as.data.frame(tbbl)
        mssng <- setdiff(expected,freq$Contact_Celltype)
        if (length(mssng)>0) {
        add  <- data.frame("Contact_Celltype"=mssng,
                   "Freq"=rep(0,length(mssng)))
        freq <- rbind(freq,add)
        }
        freq$barcode <- input[query,"barcode"][COL]
        freq$Cluster <- input[query,"Cluster"][COL]
        return(freq)
    })
    out.contacts <- do.call(rbind,out.list)
    return(as.data.frame(out.contacts))
    }else {
    indx.knn <- nn[["nn.idx"]]

    m.adj <- Matrix(0, nrow=nrow(indx.knn), ncol=nrow(indx.knn), sparse=TRUE) 
    rownames(m.adj) <- colnames(m.adj) <- rownames(indx.knn) <- rownames(input) <- input$barcode

    for (i in seq_len(nrow(m.adj))) {
        m.adj[i,rownames(indx.knn)[indx.knn[i,]]] <- 1
    }
    diag(m.adj) <- 0 # Remove the self connection
    igr <- graph_from_adjacency_matrix(m.adj)
    igr <- set_vertex_attr(igr,name="Cluster",value = unlist(input[names(V(igr)),"Cluster"]))
    out <- data.frame()
    for (ct in unique(input$Cluster)) {
        vs <- V(igr)$name[V(igr)$Cluster==ct]
        egds <- unlist(lapply(adjacent_vertices(igr,v=vs),names))
        if (length(egds)==0) { # In case a cell type as 0 edges
        tmp <- data.frame("Edges"=0,
                  "From"=ct,
                  "To"=unique(input$Cluster))
        } else {
        tbbl <- table(input[egds,"Cluster"])
        tmp <- data.frame("Edges"=as.numeric(unname(tbbl)),
                  "From"=ct,
                  "To"=names(tbbl))
        }
        mssng <- setdiff(unique(input$Cluster),tmp$To)
        if (length(mssng)>0) {
        add  <- data.frame("From"=ct,
                   "To"=mssng,
                   "Edges"=0)
        tmp <- rbind(tmp,add)
        }
        out <- rbind(out,tmp)
    }
    return(out)
    }
}

out <- list()
nneighbors <- 2
rd <- nneighbors * 23
kn <- (nneighbors * 6) + 1
# With pixels
input <- data.frame(spatialCoords(spe))
input$barcode <- spe$barcode
input$Cluster <- spe$NerveID
input$sample_id <- spe$sample_id

for (smp in unique(input$sample_id)) {
# K is 7 as it includes itslef as nn
  inp <- input[input$sample_id == smp,]
  cons <- getContacts(input=inp,rad=rd,knn=kn)
  cons <- cons[is.na(cons$Cluster) & !is.na(cons$Contact_Celltype),]
  cons <- cons[cons$Freq>0,]
# Rare cases where a spot touches two different nerves, will be randomly assigned to one of the two as border
  cons <- cons[!duplicated(cons$barcode),]
  out[[smp]] <- data.frame("barcode"=cons$barcode,
                           "BorderID"=cons$Contact_Celltype,
                           "sample_id"=smp
  )
}

add <- Reduce(rbind,out)
add$BorderInvaded <- plyr::mapvalues(add$BorderID,spe$NerveID,spe$Invaded,warn_missing =FALSE)
```

# Add all spots touching a nerve back to the object
This chunk stores the border-region annotations in the main spatial object.

```{r,message=FALSE}
add <- select(add,-sample_id)
colData(spe) <- DataFrame(dplyr::left_join(data.frame(colData(spe)),add),by="barcode")
colnames(spe) <- spe$barcode
```

#Visualize spots in direct nerve proximity
# This chunk highlights all spots in direct contact with identified nerves. Used for Figure S3C.

```{r}
exdata <- data.frame(colData(spe)[spe$barcode %in% add$barcode,])
ggcells(spe, aes(x=array_col, y=-array_row, color=factor(spe$NerveID))) +
  geom_point(size=0.8) +
  geom_point(data=exdata,size=0.8,color="black") +
  facet_wrap(~sample_id,
             labeller = as_labeller(c(
               "70865bl15_highPNI" = "Patient 3 - Diseased",
               "6249bl18_noPNIandhighPNI" = "Patient 2 - Diseased",
               "70865bl5_noPNI" = "Patient 3 - Adjacent",
               "54703bl38_noPNI" = "Patient 1 - Adjacent",
               "54703bl28_highPNI" = "Patient 1 - Diseased",
               "6249bl23_noPNI" = "Patient 2 - Adjacent"
             ))) +
  theme_void() +
  theme(legend.position ="none")
```

#Zone of no contact
==================================================
We further define an area of 8 spots from the nerve to identify tumor cells that are not nerve proximal.
Identify nerve-proximal and nerve-distal regions
This chunk expands the contact radius to define the broader nerve-proximal zone and annotates spots accordingly.

```{r,warning=FALSE, message=FALSE}
out <- list()
nneighbors <- 100
rd <- nneighbors * 23
kn <- (nneighbors * 6) + 1
# With pixels
input <- data.frame(spatialCoords(spe))
input$barcode <- spe$barcode
input$Cluster <- spe$NerveID
input$sample_id <- spe$sample_id

for (smp in unique(input$sample_id)) {
# K is 7 as it includes itslef as nn
  inp <- input[input$sample_id == smp,]
  cons <- getContacts(input=inp,rad=rd,knn=kn)
  cons <- cons[is.na(cons$Cluster) & !is.na(cons$Contact_Celltype),]
  cons <- cons[cons$Freq>0,]
# Rare cases where a spot touches two different nerves, will be randomly assigned to one of the two as border
  cons <- cons[!duplicated(cons$barcode),]
  out[[smp]] <- data.frame("barcode"=cons$barcode,
                           "BorderID"=cons$Contact_Celltype,
                           "sample_id"=smp
  )
}

add2 <- Reduce(rbind,out)
spe$NerveProximal <- spe$barcode %in% add2$barcode

ggcells(spe, aes(x=array_col, y=-array_row, color=NerveProximal)) +
  geom_point(size=0.7) +
  facet_wrap(~sample_id,
             labeller = as_labeller(c(
               "70865bl15_highPNI" = "Patient 3 - Diseased",
               "6249bl18_noPNIandhighPNI" = "Patient 2 - Diseased",
               "70865bl5_noPNI" = "Patient 3 - Adjacent",
               "54703bl38_noPNI" = "Patient 1 - Adjacent",
               "54703bl28_highPNI" = "Patient 1 - Diseased",
               "6249bl23_noPNI" = "Patient 2 - Adjacent"
             ))) +
  theme(legend.position ="none") +
  ggtitle("Nerve Proximal") +
  scale_color_manual(values=c("grey90","grey40"))
```

#Verify: We can now again check that the tumor content is higher in the vicinity of the invaded nerves.
#Each point represents an individual spot in direct contact with a nerve, grouped by the nerve it borders (BorderID). The + symbol marks the mean tumor signal across all border spots associated with each nerve. Facets separate spots bordering invaded versus non-invaded nerves.

```{r}
spe_nerve <- spe[,!is.na(spe$BorderInvaded)]
ggcells(spe_nerve, aes(x=BorderID,y=tumor)) +
  geom_jitter(color="grey80") +
  stat_summary(geom = "point",fun.y = "mean",
    size = 6, shape = "+",show.legend=FALSE) +
  facet_wrap(~BorderInvaded) +
  theme(axis.text.x =element_blank())
```

#Figure S3A_as_for_2026_March
```{r}
p0 <- ggcells(spe[, spe$patient_id %in% "54703"], aes(x = array_col, y = -array_row)) +
  geom_point(
    data = bg[bg$sample_id %in% c("54703bl38_noPNI", "54703bl28_highPNI"), ],
    size = 1,
    pch = 21,
    aes(fill = tumor, color = NULL),
    stroke = 0.05
  ) +
  geom_point(size = 0.5, pch = 19, aes(color = Invaded)) +
  facet_wrap(
    ~sample_id,
    labeller = as_labeller(c(
      "54703bl38_noPNI" = "Patient 1 - Adjacent",
      "54703bl28_highPNI" = "Patient 1 - Diseased"
    ))
  ) +
  scale_fill_viridis_c(option = "A") +
  scale_color_brewer(palette = "Accent") +
  theme_void() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x  = element_blank(),
    axis.text.y  = element_blank()
  ) +
  labs(color = "")

p1 <- ggcells(spe[, spe$patient_id %in% "54703"], aes(x=array_col, y=-array_row, color=nerve)) +
  geom_point(size=0.8) +
  facet_wrap(~sample_id,
             labeller = as_labeller(c(
              "54703bl38_noPNI" = "Patient 1 - Adjacent",
               "54703bl28_highPNI" = "Patient 1 - Diseased"
             ))) +
  scale_color_viridis_b()+
  theme_void() +
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  axis.text.x  = element_blank(),
  axis.text.y  = element_blank()
) + labs(color = "Nerve signal")

p2 <- ggcells(spe[, spe$patient_id %in% "54703"], aes(x=array_col, y=-array_row, color=is.nerve)) +
  geom_point(size=0.8)+ 
  facet_wrap(~sample_id) +
  scale_color_manual(values=c("black","grey50"))+
  theme_void() +
  theme(legend.position ='right', strip.text = element_blank(),
  strip.background = element_blank())+
  labs(color = "Spots with > 20% nerve signal")

p3 <- ggcells(spe[, spe$patient_id %in% "54703"], aes(x=array_col, y=-array_row, color=factor(NerveID))) +
  geom_point(size=0.8) +
  facet_wrap(~sample_id) +
  theme_void() +
  theme(legend.position ='none', strip.text = element_blank(),
  strip.background = element_blank())

p4 <- p0 + theme(legend.position ='right', strip.text = element_blank(),
  strip.background = element_blank())

combined_p <- p1/p2/p3/p4
#ggsave("../figures/S3_A.pdf", combined_p, width = 8, height = 12, dpi = 300)
combined_p
```
#Here we plot Figure S3C as for 2026_March
```{r}
exdata_b <- exdata[exdata$patient_id == "54703", ]
p1b <- ggcells(spe[, spe$patient_id %in% "54703"], aes(x=array_col, y=-array_row, color=factor(spe[, spe$patient_id %in% "54703"]$NerveID))) +
  geom_point(size=0.8) +
  geom_point(data=exdata_b,size=0.8,color="black") +
  facet_wrap(~sample_id,
             labeller = as_labeller(c(
              "54703bl38_noPNI" = "Patient 1 - Adjacent",
               "54703bl28_highPNI" = "Patient 1 - Diseased"
             ))) +
  theme_void() +
  theme(legend.position ="none",
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  axis.text.x  = element_blank(),
  axis.text.y  = element_blank()
)

p2b <- ggcells(spe[, spe$patient_id %in% "54703"], aes(x=array_col, y=-array_row, color=NerveProximal)) +
  geom_point(size=0.7) +
  facet_wrap(~sample_id) +
  scale_color_manual(
    name   = "Spots distal to nerves ( >8 spots away)", 
    values = c(`TRUE` = "grey40", `FALSE` = "grey90"),
    labels = c(`TRUE` = "No", `FALSE` = "Yes"))+
  theme_void() +
  theme(legend.position ='right', strip.text = element_blank(),
  strip.background = element_blank())

combined_pb <- p1b/p2b
#ggsave("../figures/S2_B.pdf", combined_pb, width = 10, height = 8, dpi = 300)
combined_pb
```

# Per nerve average composition
Since we have so many good nerves, it is important to see the per nerve NMF composition to see possible differences. The aim to see how similar the nerves are. 

```{r}
# Pseudobulking
nerve_ids <- DataFrame(
  barcode = spe$barcode,
  id=spe$NerveID,
  invasion = spe$Invaded,
  patient = spe$patient_id
)

Comp_nerve <- as.data.frame(compk30) 
names(Comp_nerve) <- names(compk30)
Comp_nerve$ID <- plyr::mapvalues(Comp_nerve$barcode, from = nerve_ids$barcode, to=nerve_ids$id)
Comp_nerve <- tidyr::drop_na(Comp_nerve)
rownames(Comp_nerve) <- Comp_nerve$barcode
Comp_nerve$barcode <- NULL

cell_types <- setdiff(names(Comp_nerve), "ID")
Comp_nerve <- as_tibble(Comp_nerve)

nerve_comp <- Comp_nerve %>%
  dplyr::group_by(ID) %>%
  dplyr::summarise(
    dplyr::across(dplyr::all_of(cell_types), ~ mean(as.numeric(.x), na.rm = TRUE)),
    .groups = "drop"
  )

nerve_comp <- nerve_comp %>%
  tidyr::pivot_longer(cols = dplyr::all_of(cell_types), names_to = "CellType", values_to = "mean_score") %>%
  dplyr::group_by(ID) %>%
  dplyr::mutate(pct = if (sum(mean_score, na.rm = TRUE) == 0) 0 else 100 * mean_score / sum(mean_score, na.rm = TRUE)) %>%
  dplyr::ungroup()

nerve_comp$invasion <- plyr::mapvalues(nerve_comp$ID, from = nerve_ids$id, to=nerve_ids$invasion, warn_missing = FALSE)
nerve_comp$patient_id <- plyr::mapvalues(nerve_comp$ID, from = nerve_ids$id, to=nerve_ids$patient, warn_missing = FALSE)
```

```{r}
color_maping <- c( "Acinar" = "gold1",
                   "B_cells" = "maroon",
                   "Lymphocyte(CD4)" = "green4",
                   "Lymphocytes" = "green4",
                   "Myeloid(Macrophages)" = "#6A3D9A",
                   "Endothelial" = "#FF7F00",
                   "Endocrine" = "yellow4",
                   "Stellate" = "khaki2",
                   "CAF" = "skyblue2",
                   "Adipocytes" = "steelblue4",
                   "Lymphocyte(NK)" = "green2",
                   "Lymphocyte(CD8)" = "palegreen2",
                   "Myeloid" = "#CAB2D6",
                   "Schwann" = "#FDBF6F",
                   "Myeloid(C1q_macrophages)" = "#CAB2D6",
                   "Nerve" = "#FDBF6F",
                   "Muscle" = "orchid1",
                   "Ductal_malignant" = "gray70",
                   "Ductal_normal" = "black",
                   "B_cells_(Plasma cells)"  = "brown1")

```


# Here we plot NMF composition per individual invaded nerve
```{r}
p5 <- ggplot(nerve_comp[nerve_comp$invasion == 'Invaded',], aes(x = ID, y = pct, fill = CellType)) +
  geom_bar(width = 0.5, stat="identity",color = NA) +
  scale_fill_manual(values = color_maping) +
  ggtitle("Invaded nerves NMF composition")+
  labs(x = "Nerve ID",
       y = "Composition") +
  facet_grid(~ patient_id, scales = "free_x", space = "free_x", labeller = as_labeller(c(
               "70865" = "Patient 3",
               "6249" = "Patient 2",
               "54703" = "Patient 1"))) +
  theme_minimal()+
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
#ggsave("../figures/Invaded_nerves_NMF_composition.pdf", p5, width = 14, height = 10)
p5
```
# Here we plot NMF composition per individual non-invaded nerve
```{r}
p6 <- ggplot(nerve_comp[nerve_comp$invasion == 'NonInvaded',], aes(x = ID, y = pct, fill = CellType)) +
  geom_bar(width = 0.5, stat="identity",color = NA) +
  scale_fill_manual(values = color_maping) +
  ggtitle("Non-invaded nerves NMF composition")+
  labs(x = "Nerve ID",
       y = "Composition") +
  facet_grid(~ patient_id, scales = "free_x", space = "free_x", labeller = as_labeller(c(
               "70865" = "Patient 3",
               "6249" = "Patient 2",
               "54703" = "Patient 1"))) +
  theme_minimal()+
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
#ggsave("../figures/Noninvaded_nerves_NMF_composition.pdf", p6, width = 14, height = 10)
p6
```

```{r}
write.csv(data.frame(colData(spe)),"../data/coldata/nerveID.csv")
```







