# scRNAseq and snRNAseq pre-pocessing 
This is the pre-processing and annotation of the sc/snRNA-seq data from: 
"Chen, M. M., Gao, Q., Ning, H., Chen, K., Gao, Y., Yu, M., ... & Zhang, Z. (2025). Integrated single-cell and spatial transcriptomics uncover distinct cellular subtypes involved in neural invasion in pancreatic cancer. Cancer Cell."

The annotated data will be used to deconvolve the ST data that we have.

```{r}
library(Seurat)
library(ggplot2)
library(patchwork)
library(readxl)


scRNA.data <- Read10X(data.dir = "../data/Deconvolution/GEOdata/scRNAseq/")
# Initialize the Seurat object with the raw (non-normalized data).
scRNA <- CreateSeuratObject(counts = scRNA.data, project = "scRNA", min.cells = 3, min.features = 200)

snRNA.data <- Read10X(data.dir = "../data/Deconvolution/GEOdata/snRNAseq/")
# Initialize the Seurat object with the raw (non-normalized data).
snRNA <- CreateSeuratObject(counts = snRNA.data, project = "snRNA", min.cells = 3, min.features = 200)

```

```{r}
scRNA[["percent.mt"]] <- PercentageFeatureSet(scRNA, pattern = "^MT-")
VlnPlot(scRNA, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```
```{r}
snRNA[["percent.mt"]] <- PercentageFeatureSet(snRNA, pattern = "^MT-")
VlnPlot(snRNA, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```
```{r}
scRNA <- subset(scRNA, subset = nFeature_RNA > 200 & nFeature_RNA < 6000 & percent.mt < 25 & nCount_RNA < 25000)

snRNA <- subset(snRNA, subset = nFeature_RNA > 200 & nFeature_RNA < 6000 & percent.mt < 25 & nCount_RNA < 25000)
```

Now we want to map the metadata provided by the authors in our Seurat object 
```{r}
scRNA_metadata <- read.csv("../data/Deconvolution/GEOdata/scRNAseq/metadata.csv")
scRNA$barcode <-names(scRNA@active.ident)
scRNA$cell_type <- plyr::mapvalues(scRNA$barcode, from = scRNA_metadata$X, to=scRNA_metadata$all_celltype)
scRNA <- subset(scRNA, subset = !is.na(cell_type))
scRNA$tissue <- plyr::mapvalues(scRNA$barcode, from = scRNA_metadata$X, to=scRNA_metadata$tissue)
scRNA$patients <- plyr::mapvalues(scRNA$barcode, from = scRNA_metadata$X, to=scRNA_metadata$patients)
```

```{r}
snRNA_metadata <- read.csv("../data/Deconvolution/GEOdata/snRNAseq/metadata.csv")
snRNA$barcode <-names(snRNA@active.ident)
snRNA$cell_type <- plyr::mapvalues(snRNA$barcode, from = snRNA_metadata$X, to=snRNA_metadata$all_celltype)
snRNA <- subset(snRNA, subset = !is.na(cell_type))
snRNA$tissue <- plyr::mapvalues(snRNA$barcode, from = snRNA_metadata$X, to=snRNA_metadata$tissue)
snRNA$patients <- plyr::mapvalues(snRNA$barcode, from = snRNA_metadata$X, to=snRNA_metadata$patients)
```

```{r}
merged_obj <- merge(x = scRNA, y = snRNA, add.cell.ids = c("scRNA", "snRNA"),)
# For our analysis we don't need the PBMC data so we remove them 
merged_obj <- subset(merged_obj, subset = !(merged_obj$tissue == "PBMC"))
merged_obj <- NormalizeData(merged_obj)
merged_obj <- FindVariableFeatures(merged_obj)
merged_obj <- ScaleData(merged_obj)
merged_obj <- RunPCA(merged_obj)
```
Then, according to the authors, we integrated the two datasets with the following parameters and function: 
```{r}
#devtools::install_github("ycli1995/bbknnR")
library(bbknnR)
merged_obj <- RunBBKNN(merged_obj, batch_key = 'orig.ident', n_pcs = 30,run_UMAP = TRUE)
```
```{r}
merged_obj <- FindNeighbors(merged_obj, dims = 1:10)
merged_obj <- FindClusters(merged_obj, resolution = 1)
```

```{r}
merged_obj <- JoinLayers(merged_obj)
DimPlot(merged_obj, reduction = "umap")
```
We can use their metadata to characterize the ductal malignant cells as follows: 
```{r}
# make a character copy so new labels are allowed
merged_obj$cell_type_refined <- as.character(merged_obj$cell_type)

merged_obj$cell_type_refined[merged_obj$cell_type == "Ductal" & merged_obj$tissue == "Tumor"]           <- "Ductal_malignant"
merged_obj$cell_type_refined[merged_obj$cell_type == "Ductal" & merged_obj$tissue == "Adjacent_normal"] <- "Ductal_normal"
merged_obj <- subset(merged_obj, subset = !(merged_obj$cell_type_refined == "Platelet")) # Because after removing PBMC samples only 2 cells left and that leads to errors when running RCTD
```
Now we can use this for the reference.
```{r}
saveRDS(merged_obj,"../data/Deconvolution/sc_sn_merged_obj.Rds")
```



