# Plotting of Deconvolution data 
This section quantifies immune cell composition and macrophage-associated transcriptional programs in spatial transcriptomic spots corresponding to pancreatic nerves. Spatial spots annotated as nerves were classified into three biologically meaningful categories based on pathological annotation:
Invaded, Non-Invaded Diseased, Non-invaded Adjacent.

Cell type proportions estimated using RCTD deconvolution, while transcriptional programs were derived from NMF decomposition of spatial transcriptomic data.

#Figures generated in this section
#Figure2F_violin_Adaptive.pdf
#Distribution of adaptive immune and plasma cell proportions across nerve categories.
#Figure2F_violin_Myeloid_NMF.pdf
#Distribution of macrophage-associated NMF transcriptional programs across nerve categories.
#Individual_nerve_deconvolution_3cat.pdf
#Cell-type composition of individual nerves based on RCTD deconvolution.
#Nerves_mean_composition_by_cat3.pdf
#Mean cell-type composition across nerve categories.
#heatmap_deconvolution_cat3_nerves.pdf
#Enrichment heatmap of deconvolved cell types across nerve categories.
#Nerves_mean_composition_NMF_by_cat3.pdf
#Mean transcriptional program composition across nerve categories


```{r}
#=================================
#1. Load Libraries and input data 
#================================
library(spacexr)
library(readxl)
library(stringr)
library(Seurat)
library(ggplot2)
library(patchwork)
library(S4Vectors)
library(SpatialExperiment)
library(SummarizedExperiment)
library(scuttle)
library(scran)
library(scater)
library(ggspavis)
library(dplyr)
library(ggrepel)
library(batchelor)
library(reshape2)
library(Matrix)
library(scatterpie)
library(tidyr)
library(pheatmap)
library(tidyverse)
library(purrr)
library(grid)
library(zellkonverter)
library(lme4)
library(msigdbr)

c25 <- c(
  "dodgerblue2", "#E31A1C", # red
"green4",
"#6A3D9A", # purple
"#FF7F00", # orange
"black", "gold1",
"skyblue2", "#FB9A99", # lt pink
"palegreen2",
"#CAB2D6", # lt purple
"#FDBF6F", # lt orange
"gray70", "khaki2",
"maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
"darkturquoise", "green1", "yellow4", "yellow3",
"darkorange4", "brown"
)


merged_obj <-readRDS("../data/Deconvolution/sc_sn_merged_obj.Rds")
spe <- readRDS("../data/SPE/Integrated_NMF.rds")
colData(spe) <- DataFrame(read.csv("../data/coldata/DE_SPE_annotaion.csv"))
colnames(spe) <- spe$barcode

W_df <- readRDS("../data/Deconvolution/RCTD_weights.Rds")
weights_df_copy <- W_df

```

```{r}
#==========================================
# 2. Identify nerve spots and define groups
#==========================================

# Identify nerve spots
nerve_mask <- colData(spe)$is.nerve == TRUE

nerve_meta_df <- data.frame(
  barcode     = as.character(spe$barcode[nerve_mask]),
  patient_id  = as.factor(colData(spe)$patient_id[nerve_mask]),
  NerveID     = as.factor(colData(spe)$NerveID[nerve_mask]),
  Invaded     = as.character(colData(spe)$Invaded[nerve_mask]),
  type_raw    = as.character(colData(spe)$type[nerve_mask])
)

# Create category (new column in new dataframe only)
nerve_meta_df$types_grouped <- ifelse(
  nerve_meta_df$type_raw == "noPNI",
  "Adjacent",
  "Diseased"
)

nerve_meta_df$nerve_category <- ifelse(
  nerve_meta_df$Invaded == "NonInvaded",
  paste0("NonInvaded_", nerve_meta_df$types_grouped),
  nerve_meta_df$Invaded
)

nerve_meta_df <- nerve_meta_df[!is.na(nerve_meta_df$nerve_category), ]

# Join with RCTD weights (copy only)
analysis_df <- nerve_meta_df %>%
  left_join(weights_df_copy, by = "barcode")
```

```{r}
#=====================================================
# RCTD-only plotting: 4 selected immune cell types
#=====================================================

#=====================================================
# Set plotting order and labels
#=====================================================

nerve_order <- c("NonInvaded_Adjacent", "NonInvaded_Diseased", "Invaded")
patient_order <- c("54703", "6249", "70865")

rctd_selected_cols <- c(
  "Myeloid",
  "Lymphocyte(NK)",
  "Lymphocyte(CD8)",
  "B_cells_(Plasma_cells)"
)

rctd_labels <- c(
  "Myeloid"                = "Myeloid",
  "Lymphocyte(NK)"         = "NK lymphocytes",
  "Lymphocyte(CD8)"        = "CD8 T cells",
  "B_cells_(Plasma_cells)" = "B/plasma cells"
)
```

```{r}
#=====================================================
# Prepare spot-level RCTD table
#=====================================================

rctd_spot_long <- analysis_df %>%
  dplyr::select(
    barcode,
    patient_id,
    NerveID,
    nerve_category,
    dplyr::all_of(rctd_selected_cols)
  ) %>%
  dplyr::mutate(
    barcode = as.character(barcode),
    NerveID = as.character(NerveID),
    patient_id = factor(as.character(patient_id), levels = patient_order),
    nerve_category = factor(as.character(nerve_category), levels = nerve_order)
  ) %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(rctd_selected_cols),
    names_to = "CellType",
    values_to = "Proportion"
  ) %>%
  dplyr::mutate(
    CellType = factor(CellType, levels = rctd_selected_cols),
    CellTypeLabel = factor(
      rctd_labels[as.character(CellType)],
      levels = unname(rctd_labels[rctd_selected_cols])
    )
  ) %>%
  dplyr::filter(
    !is.na(nerve_category),
    !is.na(NerveID),
    !is.na(Proportion)
  )
```

```{r}
#=====================================================
# Aggregate from spot level to nerve level
#=====================================================

rctd_nerve_long <- rctd_spot_long %>%
  dplyr::group_by(
    NerveID,
    patient_id,
    nerve_category,
    CellType,
    CellTypeLabel
  ) %>%
  dplyr::summarise(
    MeanProportion = mean(Proportion, na.rm = TRUE),
    n_spots = dplyr::n(),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    nerve_category = factor(nerve_category, levels = nerve_order),
    patient_id = factor(patient_id, levels = patient_order)
  )

```

```{r}
#=====================================================
# Plot spot-level RCTD proportions
#=====================================================

facet_ncol   <- 2
facet_scales <- "free_y"
base_size    <- 11
panel_aspect <- 0.5

p_rctd_spot <- ggplot(
  rctd_spot_long,
  aes(x = nerve_category, y = Proportion)
) +
  geom_violin(
    fill = "grey85",
    color = NA,
    alpha = 0.6,
    trim = FALSE
  ) +
  geom_boxplot(
    width = 0.25,
    outlier.shape = NA,
    fill = "white",
    color = "grey30",
    linewidth = 0.4
  ) +
  geom_jitter(
    aes(color = patient_id),
    width = 0.10,
    size = 0.9,
    alpha = 0.95
  ) +
  facet_wrap(
    ~ CellTypeLabel,
    scales = "free_y",
    ncol = 2
  ) +
  theme_classic(base_size = base_size) +
  theme(
    aspect.ratio = panel_aspect,
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = base_size * 0.85),
    legend.position = "right",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4),
    panel.spacing = unit(6, "pt")
  ) +
  ylab("RCTD proportion per spot") +
  xlab("Nerve category") +
  labs(color = "Patient ID")

p_rctd_spot

```

```{r}
#=====================================================
# Plot nerve-level RCTD proportions
#=====================================================

p_rctd_nerve <- ggplot(
  rctd_nerve_long,
  aes(x = nerve_category, y = MeanProportion)
) +
  geom_violin(
    fill = "grey85",
    color = NA,
    alpha = 0.6,
    trim = FALSE
  ) +
  geom_boxplot(
    width = 0.25,
    outlier.shape = NA,
    fill = "white",
    color = "grey30",
    linewidth = 0.4
  ) +
  geom_jitter(
    aes(color = patient_id),
    width = 0.10,
    size = 0.9,
    alpha = 0.95
  ) +
  facet_wrap(
    ~ CellTypeLabel,
    scales = "free_y",
    ncol = 2
  ) +
  theme_classic(base_size = base_size) +
  theme(
    aspect.ratio = panel_aspect,
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = base_size * 0.85),
    legend.position = "right",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4),
    panel.spacing = unit(6, "pt")
  ) +
  ylab("Mean RCTD proportion per nerve") +
  xlab("Nerve category") +
  labs(color = "Patient ID")

p_rctd_nerve

```

```{r}
#=====================================================
# Save RCTD-only 4-panel plots
#=====================================================

ggsave(
  filename = "../figures/RCTD_selected_immune_spot_level.pdf",
  plot = p_rctd_spot,
  width = 7,
  height = 7,
  units = "in",
  dpi = 600,
  useDingbats = FALSE
)

ggsave(
  filename = "../figures/RCTD_selected_immune_nerve_level.pdf",
  plot = p_rctd_nerve,
  width = 7,
  height = 7,
  units = "in",
  dpi = 600,
  useDingbats = FALSE
)
```

