# Robust Cell Type Decomposition (RCTD) 
For this part we want to deconvolute the spatial data using a scRNA + snRNA seq reference dataset.

```{r}
library(spacexr)
library(readxl)
library(stringr)
library(Seurat)
library(ggplot2)
library(patchwork)
library(S4Vectors)
library(SpatialExperiment)
library(SummarizedExperiment)
library(scuttle)
library(scran)
library(scater)
library(ggspavis)
library(dplyr)
library(ggrepel)
library(batchelor)
library(reshape2)
library(Matrix)
library(scatterpie)
library(tidyr)
library(pheatmap)
library(tidyverse)
library(purrr)
library(grid)
library(zellkonverter)

c25 <- c(
  "dodgerblue2", "#E31A1C", # red
"green4",
"#6A3D9A", # purple
"#FF7F00", # orange
"black", "gold1",
"skyblue2", "#FB9A99", # lt pink
"palegreen2",
"#CAB2D6", # lt purple
"#FDBF6F", # lt orange
"gray70", "khaki2",
"maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
"darkturquoise", "green1", "yellow4", "yellow3",
"darkorange4", "brown"
)


merged_obj <-readRDS("../data/Deconvolution/sc_sn_merged_obj.Rds")
spe <- readRDS("../data/SPE/Integrated_NMF.rds")
colData(spe) <- DataFrame(read.csv("../data/coldata/DE_SPE_annotaion.csv"))
colnames(spe) <- spe$barcode

```

First, we need to set the reference data (scRNA &snRNA seq)
```{r}
# extract information to pass to the RCTD Reference function
counts <- merged_obj[["RNA"]]$counts
cluster <- as.factor(merged_obj$cell_type_refined) # to characterize tumor 
names(cluster) <- colnames(merged_obj)
nUMI <- merged_obj$nCount_RNA
names(nUMI) <- colnames(merged_obj)
reference <- Reference(counts, cluster, nUMI)
saveRDS(reference,"../data/Deconvolution/reference.Rds")
```

```{r}
counts <- spe@assays@data@listData[["counts"]]

image_names <- as.list(unique(spe$sample_id))
sample_ids <- colData(spe)$sample_id
coordinates_list <- lapply(image_names, function(image_name) {
  pos <- spatialCoords(spe)[sample_ids == image_name, ]
  colnames(pos) <- c('x','y')
  return(pos)
})

coords <- do.call(rbind, coordinates_list)
coords[is.na(colnames(coords))] <- NULL
coords_df<-as.data.frame(coords)

query <- SpatialRNA(coords_df, counts, colSums(counts))
saveRDS(query,"../data/Deconvolution/query.Rds")
```

**This part was run in Euler (high-performance computing (HPC) clusters)**
```{r}
RCTD <- create.RCTD(query, reference, max_cores = 8)
RCTD <- run.RCTD(RCTD, doublet_mode = "doublet")
```

```{r}
RCTD <- readRDS("../data/Deconvolution/RCTD.Rds")

#After the run we need to normalize the weights so then each pixel represents cell type proportions
weights <- RCTD@results$weights
normalized_weights <- normalize_weights(weights)
```

Now we want to see the spot composition, from the RCTD results
```{r}

W <- Diagonal(x = 1 / rowSums(weights)) %*% weights   # row-normalize without densifying

barcodes <- weights@Dimnames[[1]]

W_df <- as.data.frame(as.matrix(W))  
W_df$barcode <- barcodes
W_df <- dplyr::relocate(W_df, barcode)

W_df$barcode <- as.character(W_df$barcode)

names(W_df)[names(W_df) == "B"] <- "B_cells"
names(W_df)[names(W_df) == "CD4T"] <- "Lymphocyte(CD4)"
names(W_df)[names(W_df) == "CD8T"] <- "Lymphocyte(CD8)"
names(W_df)[names(W_df) == "NK"] <- "Lymphocyte(NK)"
names(W_df)[names(W_df) == "Plasma"] <- "B_cells_(Plasma_cells)"

saveRDS(W_df, "../data/Deconvolution/RCTD_weights.Rds")
```
We map the colors for deconvolution and NMF to match the similar cell types.
```{r}

color_maping <- c( "Acinar" = "gold1",
                   "B_cells" = "maroon",
                   "Lymphocyte(CD4)" = "green4",
                   "Lymphocytes" = "green4",
                   "Myeloid(Macrophages)" = "#6A3D9A",
                   "Endothelial" = "#FF7F00",
                   "Endocrine" = "yellow4",
                   "Stellate" = "khaki2",
                   "CAF" = "skyblue2",
                   "Adipocytes" = "steelblue4",
                   "Lymphocyte(NK)" = "green2",
                   "Lymphocyte(CD8)" = "palegreen2",
                   "Myeloid" = "#CAB2D6",
                   "Schwann" = "#FDBF6F",
                   "Myeloid(C1q_macrophages)" = "#CAB2D6",
                   "Nerve" = "#FDBF6F",
                   "Muscle" = "orchid1",
                   "Ductal_malignant" = "gray70",
                   "Ductal_normal" = "black",
                   "B_cells_(Plasma cells)"  = "brown1")

```
# Per nerve average composition
Since we have so many good nerves, it is important to see the per nerve composition to see possible differences. The aim to see how similar the nerves are. 

```{r}
# Pseudobulking
nerve_ids <- DataFrame(
  barcode = spe$barcode,
  id=spe$NerveID,
  invasion = spe$Invaded,
  patient = spe$patient_id
)

W_nerve <-W_df
W_nerve$ID <- plyr::mapvalues(W_nerve$barcode, from = nerve_ids$barcode, to=nerve_ids$id)
W_nerve<- W_nerve %>% drop_na()
rownames(W_nerve) <- W_nerve$barcode
W_nerve$barcode <- NULL

cell_cols <- setdiff(names(W_nerve), "ID")

nerve_decon_w <- W_nerve %>%
  group_by(ID) %>%
  summarise(
    across(all_of(cell_cols), ~ mean(as.numeric(.x), na.rm = TRUE)),
    .groups = "drop"
  )

nerve_decon_w_long <- nerve_decon_w %>%
  pivot_longer(cols = all_of(cell_cols), names_to = "CellType", values_to = "mean_score") %>%
  group_by(ID) %>%
  mutate(pct = if (sum(mean_score, na.rm = TRUE) == 0) 0 else 100 * mean_score / sum(mean_score, na.rm = TRUE)) %>%
  ungroup()


nerve_decon_w_long$invasion <- plyr::mapvalues(nerve_decon_w_long$ID, from = nerve_ids$id, to=nerve_ids$invasion, warn_missing = FALSE)
nerve_decon_w_long$patient_id <- plyr::mapvalues(nerve_decon_w_long$ID, from = nerve_ids$id, to=nerve_ids$patient, warn_missing = FALSE)
```

```{r}
p1 <- ggplot(nerve_decon_w_long[nerve_decon_w_long$invasion == 'Invaded',], aes(x = ID, y = pct, fill = CellType)) +
  geom_bar(width = 0.5, stat="identity",color = NA) +
  scale_fill_manual(values = color_maping) +
  ggtitle("Invaded nerves deconvolution composition")+
  labs(x = "Nerve ID",
       y = "Composition") +
  facet_grid(~ patient_id, scales = "free_x", space = "free_x", labeller = as_labeller(c(
               "70865" = "Patient 3",
               "6249" = "Patient 2",
               "54703" = "Patient 1"))) +
  theme_minimal()+
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
ggsave("../figures/Invaded_nerves_composition.pdf", p1, width = 14, height = 10)
p1
```
```{r}
p2 <- ggplot(nerve_decon_w_long[nerve_decon_w_long$invasion == 'NonInvaded',], aes(x = ID, y = pct, fill = CellType)) +
  geom_bar(width = 0.5, stat="identity",color = NA) +
  scale_fill_manual(values = color_maping) +
  ggtitle("Non-invaded nerves deconvolution composition")+
  labs(x = "Nerve ID",
       y = "Composition") +
  facet_grid(~ patient_id, scales = "free_x", space = "free_x", labeller = as_labeller(c(
               "70865" = "Patient 3",
               "6249" = "Patient 2",
               "54703" = "Patient 1"))) +
  theme_minimal()+
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
ggsave("../figures/Noninvaded_nerves_composition.pdf", p2, width = 14, height = 10)
p2
```

# Visualization of composition
Now that we have the weights, we want to visualize the composition of the important categories that we care about. 
```{r}
inv_nerve_barcodes <- spe[, !is.na(colData(spe)$Invaded) & colData(spe)$Invaded == "Invaded"]$barcode
non_inv_nerve_barcodes <- spe[, !is.na(colData(spe)$Invaded) & colData(spe)$Invaded == "NonInvaded"]$barcode
surr_inv_nerve_barcodes <- spe[, !is.na(colData(spe)$invasion) & colData(spe)$invasion == "Surr_Invaded"]$barcode
surr_non_inv_nerve_barcodes <- spe[, !is.na(colData(spe)$invasion) & colData(spe)$invasion == "Surr_NonInvaded"]$barcode


Avrg_inv <- W_df |> 
  filter(barcode %in% inv_nerve_barcodes) |> 
  select(-barcode) |> 
  summarise(across(everything(), mean, na.rm = TRUE)) |> 
  as.data.frame() |> 
  tibble::as_tibble() |> 
  pivot_longer(everything(), names_to = "cell_type", values_to = "mean") |> 
  filter(mean > 0) |> 
  arrange(desc(mean)) |> 
  mutate(pct = mean / sum(mean),
         Categ = "Invaded", location = "Inside")

Avrg_non_inv <- W_df |> 
  filter(barcode %in% non_inv_nerve_barcodes) |> 
  select(-barcode) |> 
  summarise(across(everything(), mean, na.rm = TRUE)) |> 
  as.data.frame() |> 
  tibble::as_tibble() |> 
  pivot_longer(everything(), names_to = "cell_type", values_to = "mean") |> 
  filter(mean > 0) |> 
  arrange(desc(mean)) |> 
  mutate(pct = mean / sum(mean),
         Categ = "NonInvaded", location = "Inside")

Avrg_surr_inv <- W_df |> 
  filter(barcode %in% surr_inv_nerve_barcodes) |> 
  select(-barcode) |> 
  summarise(across(everything(), mean, na.rm = TRUE)) |> 
  as.data.frame() |> 
  tibble::as_tibble() |> 
  pivot_longer(everything(), names_to = "cell_type", values_to = "mean") |> 
  filter(mean > 0) |> 
  arrange(desc(mean)) |> 
  mutate(pct = mean / sum(mean),
         Categ = "Invaded", location = "Proximal")

Avrg_surr_non_inv <- W_df |> 
  filter(barcode %in% surr_non_inv_nerve_barcodes) |> 
  select(-barcode) |> 
  summarise(across(everything(), mean, na.rm = TRUE)) |> 
  as.data.frame() |> 
  tibble::as_tibble() |> 
  pivot_longer(everything(), names_to = "cell_type", values_to = "mean") |> 
  filter(mean > 0) |> 
  arrange(desc(mean)) |> 
  mutate(pct = mean / sum(mean),
         Categ = "NonInvaded", location = "Proximal")

compos <- rbind(Avrg_inv,Avrg_non_inv,Avrg_surr_inv,Avrg_surr_non_inv)
```


```{r}
p3 <- ggplot(compos, aes(y = location, x = pct, fill = cell_type)) +
  geom_bar(stat = "identity", width = 0.7, color = NA) +
  scale_x_continuous(
    labels = function(x) x * 100,
    expand = c(0, 0),
    limits = c(0, 1)
  ) +
  scale_y_discrete(position = "right") +   
  scale_fill_manual("Cell type", values = color_maping) +
  facet_grid(Categ ~ ., switch = "y") +    
  theme_minimal(base_size = 10) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(size = 10),
    axis.text.y.right = element_text(size = 10),  
    axis.text.y.left  = element_text(size = 10),        
    axis.ticks.y.left = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
    strip.text.y = element_text(angle = 0),
    strip.placement = "outside",
    legend.position = "bottom",
    plot.margin = margin(10, 10, 20, 10)
  )
ggsave("../figures/F2_C.pdf", p3, width = 8, height = 4)
p3
```

# Heatmap with z-scores
Now that we have the cell type information, we want to see the contribution of the different cell types in our categories.
```{r}
sub_spe <- spe[, colData(spe)$invasion != "other" ] #to have more clear results 

rownames(W_df) <- W_df$barcode
W_df$barcode <-NULL

sub_w <- W_df[rownames(W_df) %in% sub_spe$barcode, , drop = TRUE]

categories <- data.frame(
  barcodes = as.character(colData(sub_spe)[, "barcode"]),
  category = as.character(colData(sub_spe)[, "invasion"]))
cats <- unique(categories$category)
cell_types <- colnames(W_df)

# ---- Function for enrichment z-score ----
enrichment_z <- function(vec, group, target_cat) {
  bcs_in  <- group$barcodes[group$category == target_cat]
  x_in  <- vec[bcs_in]
  mu_in <- mean(x_in,  na.rm = TRUE)
  mu_all<- mean(vec, na.rm = TRUE)
  (mu_in - mu_all) / sd(vec)
}

```

```{r}
z_mat <- sapply(cell_types, function(ct) {
 v <- setNames(sub_w[, ct], rownames(sub_w))
  sapply(cats, function(ca) enrichment_z(v, categories, ca))
})
rownames(z_mat) <- cats
rownames(z_mat) <- dplyr::recode(rownames(z_mat),
                                 "Surr_Invaded" = "Invaded nerve proximal",
                                 "Nerve" = "Non-invaded nerve",
                                 "Surr_NonInvaded" = "Non-invaded nerve proximal",
                                 "InvadedNerve" = "Invaded nerve",
                                 "TumorDistal" = "Tumor Distal")
colnames(z_mat) <- gsub("_", " ", colnames(z_mat))

p4 <- pheatmap(z_mat,
         cluster_rows = TRUE, cluster_cols = TRUE,
         scale = "none",
         color = colorRampPalette(c(c25[1], "white", c25[2]))(100),
         fontsize_number = 10)

ggsave("../figures/heatmap_deconvolution.pdf", p4, width = 12, height = 6)
p4
```
Saving the SPE object for python analysis.
```{r}
writeH5AD(spe,"../data/SPE/clean_adata.h5ad",X_name="logcounts",verbose=TRUE)
```

