
#This script investigates DRAXIN expression in pancreatic tumour-associated nerves using spatial transcriptomic data. It first visualises DRAXIN across all tissue sections, then focuses on nerve-associated spots to compare DRAXIN levels across three nerve states: non-invaded adjacent nerves, non-invaded diseased nerves, and invaded nerves. The script quantifies DRAXIN both at the nerve level and the spot level, and then examines whether DRAXIN-high tumour-associated nerves are associated with specific Schwann cell transcriptional programs.

#Output Figures:

#Spatial DRAXIN expression across tissues – Faceted spatial plot showing DRAXIN signal across all samples.
#Spatial DRAXIN with nerve outlines – Spatial DRAXIN map with nerve borders coloured by invasion status.
#Saved as: ../figures/Draxin_signal.png
#Spatial Adipocyte signal with nerve outlines by invasion status- Figure S4A vs 2026_March
#Nerve-level DRAXIN expression – Violin/boxplot showing mean DRAXIN per nerve across adjacent, diseased, and invaded nerves.
#Spot-level DRAXIN expression – Violin/boxplot showing DRAXIN expression per spot across nerve categories.
#Saved as: DRAXIN_logcounts_violing_spot.pdf
#Schwann programs vs DRAXIN levels – Heatmap of Schwann signature enrichment across DRAXIN-low, normal, and high tumour-associated nerves.
#Saved as: ../figures/heatmap_Schwann_by_DRAXIN_tumour_associated_nerves.pdf
#Schwann programs across nerve categories – Heatmap of Schwann signature enrichment across adjacent, diseased, and invaded nerves.
#Saved as: ../figures/heatmap_Schwann_cat3vs2.pdf


```{r, warning=FALSE}}
#=================================
#1. Load libraries and input data
#=================================
# This section loads required packages, helper functions, colour mappings,
# and the integrated spatial object with accompanying annotation data.
# Output: no figure; workspace and input objects for downstream DRAXIN analyses.

library(patchwork)
library(SpatialExperiment)
library(scuttle)
library(batchelor)
library(scran)
library(scater)
library(ggspavis)
library(dplyr)
library(ggrepel)
library(reshape2)
library(ggplot2)
library(tidyr)
library(readxl)
library(pheatmap)
library(UCell)

theme_set(theme_minimal(base_size = 14))
source("../scripts/general_functions.R")

minmaxscale <- function(x, ...) {
  (x - min(x, ...)) / (max(x, ...) - min(x, ...))
}

get_gene_vec <- function(se, gene, assay_priority = c("logcounts", "counts")) {
  rn <- rownames(se)
  hit <- which(rn == gene)
  if (length(hit) == 0) hit <- which(toupper(rn) == toupper(gene))
  if (length(hit) == 0) stop("Gene not found in rownames(se): ", gene)

  for (assay_name in assay_priority) {
    if (assay_name %in% assayNames(se)) {
      return(as.numeric(assay(se, assay_name)[hit[1], ]))
    }
  }
  stop("None of the requested assays found: ", paste(assay_priority, collapse = ", "))
}

score_signature_mat <- function(genes, mat) {
  idx <- match(toupper(unique(genes)), toupper(rownames(mat)))
  idx <- idx[!is.na(idx)]
  if (length(idx) == 0) return(rep(NA_real_, ncol(mat)))
  colMeans(mat[idx, , drop = FALSE], na.rm = TRUE)
}

c25 <- c(
  "dodgerblue2", "#E31A1C", "green4", "#6A3D9A", "#FF7F00",
  "black", "gold1", "skyblue2", "#FB9A99", "palegreen2",
  "#CAB2D6", "#FDBF6F", "gray70", "khaki2", "maroon",
  "orchid1", "deeppink1", "blue1", "steelblue4", "darkturquoise",
  "green1", "yellow4", "yellow3", "darkorange4", "brown"
)

color_maping <- c(
  "Acinar" = "gold1",
  "B_cells" = "maroon",
  "Lymphocyte(CD4)" = "green4",
  "Lymphocytes" = "green4",
  "Myeloid(Macrophages)" = "#6A3D9A",
  "Endothelial" = "#FF7F00",
  "Endocrine" = "yellow4",
  "Stellate" = "khaki2",
  "CAF" = "skyblue2",
  "Adipocytes" = "steelblue4",
  "Lymphocyte(NK)" = "green2",
  "Lymphocyte(CD8)" = "palegreen2",
  "Myeloid" = "#CAB2D6",
  "Schwann" = "#FDBF6F",
  "Myeloid(C1q_macrophages)" = "#CAB2D6",
  "Nerve" = "#FDBF6F",
  "Muscle" = "orchid1",
  "Ductal_malignant" = "gray70",
  "Ductal_normal" = "black",
  "B_cells_(Plasma cells)" = "brown1"
)

spe <- readRDS("../data/SPE/Integrated_NMF.rds")
colData(spe) <- DataFrame(read.csv("../data/coldata/DE_SPE_annotaion.csv"))
colnames(spe) <- spe$barcode

bg <- data.frame(colData(spe)[, c("array_col", "array_row", "tumor", "sample_id")])
W_df <- readRDS("../data/Deconvolution/RCTD_weights.Rds")

# Extract NMF component matrix for selected spatial signal visualisations
comp <- as.matrix(t(reducedDims(spe)[["compk30"]]))
spe$adipocytes <- comp["Adipocytes", ]

```

```{r}
#=================================
#2. Visualise DRAXIN expression across all tissue sections
#=================================
# This section plots the spatial distribution of DRAXIN across all samples.
# Output: faceted spatial DRAXIN expression plot across all tissue blocks.

ggcells(spe, aes(x = array_col, y = -array_row, color = DRAXIN)) +
  geom_point(size = 0.7, pch = 19) +
  facet_wrap(~sample_label) +
  scale_color_viridis_c(option = "A") +
  theme_void() +
  labs(color = "DRAXIN")

```


```{r}
#=================================
#3. Define nerve-only objects and metadata
#=================================
# This section subsets the spatial object to nerve spots and builds a unified
# metadata table used in downstream DRAXIN quantification and category-based analyses.
# Output: no figure; nerve-only objects and metadata tables.

spe_nerve <- spe[, colData(spe)$is.nerve == TRUE]

nerve_meta_df <- data.frame(
  barcode    = as.character(spe_nerve$barcode),
  patient_id = as.factor(colData(spe_nerve)$patient_id),
  NerveID    = as.character(colData(spe_nerve)$NerveID),
  Invaded    = as.character(colData(spe_nerve)$Invaded),
  type_raw   = as.character(colData(spe_nerve)$type),
  stringsAsFactors = FALSE
)

nerve_meta_df$types_grouped <- ifelse(nerve_meta_df$type_raw == "noPNI", "Adjacent", "Diseased")
nerve_meta_df$nerve_category <- ifelse(
  nerve_meta_df$Invaded == "NonInvaded",
  paste0("NonInvaded_", nerve_meta_df$types_grouped),
  nerve_meta_df$Invaded
)

nerve_meta_df <- nerve_meta_df[!is.na(nerve_meta_df$nerve_category), ]
nerve_meta_df$nerve_category <- factor(
  nerve_meta_df$nerve_category,
  levels = c("NonInvaded_Adjacent", "NonInvaded_Diseased", "Invaded")
)

```

```{r}
#=================================
#4. Plot DRAXIN across tissue blocks with nerve outlines
#=================================
# This section visualises spatial DRAXIN expression across all samples and
# overlays nerve hulls coloured by invasion status.
# Output: spatial DRAXIN plot with invaded and non-invaded nerve outlines.

sample_order <- c(
  "54703bl38_noPNI",
  "6249bl23_noPNI",
  "70865bl5_noPNI",
  "54703bl28_highPNI",
  "6249bl18_noPNIandhighPNI",
  "70865bl15_highPNI"
)

label_map <- c(
  "54703bl38_noPNI" = "P1-Adjacent",
  "6249bl23_noPNI" = "P2-Adjacent",
  "70865bl5_noPNI" = "P3-Adjacent",
  "54703bl28_highPNI" = "P1-Diseased",
  "6249bl18_noPNIandhighPNI" = "P2-Diseased",
  "70865bl15_highPNI" = "P3-Diseased"
)

colData(spe)$sample_id <- factor(as.character(colData(spe)$sample_id), levels = sample_order)
colData(spe_nerve)$sample_id <- factor(as.character(colData(spe_nerve)$sample_id), levels = sample_order)

p_draxin_spatial <- ggcells(
  spe,
  aes(x = array_col, y = -array_row, color = DRAXIN)
) +
  geom_point(size = 0.7, pch = 19) +
  facet_wrap(~sample_id, labeller = as_labeller(label_map)) +
  scale_color_viridis_c(option = "A") +
  theme_void() +
  labs(color = "DRAXIN") +
  theme(
    legend.position = "right",
    strip.text = element_text(size = 10, face = "bold"),
    strip.background = element_blank()
  )

make_nerve_hulls <- function(dat) {
  dat %>%
    group_by(sample_id, NerveID, Invaded) %>%
    filter(n() >= 3) %>%
    slice(chull(x, y)) %>%
    ungroup() %>%
    mutate(group_id = interaction(sample_id, NerveID, drop = TRUE))
}

nerve_all <- as.data.frame(colData(spe_nerve)) %>%
  transmute(
    x = as.numeric(array_col),
    y = -as.numeric(array_row),
    sample_id = factor(as.character(sample_id), levels = sample_order),
    NerveID = as.character(NerveID),
    Invaded = as.character(Invaded)
  ) %>%
  filter(is.finite(x), is.finite(y), !is.na(sample_id), !is.na(NerveID), !is.na(Invaded))

hulls_all <- make_nerve_hulls(nerve_all)
hulls_inv <- hulls_all %>% filter(Invaded == "Invaded")
hulls_non <- hulls_all %>% filter(Invaded == "NonInvaded")

p_draxin_spatial_hulls <- p_draxin_spatial +
  geom_polygon(
    data = hulls_non,
    inherit.aes = FALSE,
    aes(x = x, y = y, group = group_id),
    fill = NA, linewidth = 0, color = "white"
  ) +
  geom_polygon(
    data = hulls_inv,
    inherit.aes = FALSE,
    aes(x = x, y = y, group = group_id),
    fill = NA, linewidth = 0, color = "white"
  ) +
  geom_polygon(
    data = hulls_non,
    inherit.aes = FALSE,
    aes(x = x, y = y, group = group_id, linetype = "Non-invaded"),
    fill = NA, linewidth = 0.4, color = "skyblue2"
  ) +
  geom_polygon(
    data = hulls_inv,
    inherit.aes = FALSE,
    aes(x = x, y = y, group = group_id, linetype = "Invaded"),
    fill = NA, linewidth = 0.4, color = "khaki2"
  ) +
  scale_linetype_manual(
    name = "Nerve border",
    values = c("Invaded" = "solid", "Non-invaded" = "solid")
  ) +
  guides(linetype = guide_legend(override.aes = list(linewidth = 1.2)))

p_draxin_spatial_hulls
```
```{r}
#=================================
#5. Save DRAXIN spatial plot with nerve outlines
#=================================
# This section exports the spatial DRAXIN plot with nerve hull overlays.
# Output: ../figures/Draxin_signal.png

ggsave("../figures/Draxin_signal.png", p_draxin_spatial_hulls,
       width = 12, height = 7, dpi = 600, bg = "white")
```


```{r}
#=================================
#4b. Plot DRAXIN across selected tissue blocks with nerve outlines
#=================================
# This section visualises spatial DRAXIN expression across the selected samples
# and overlays nerve hulls coloured by invasion status.
# Output: spatial DRAXIN plot with invaded and non-invaded nerve outlines.

sample_order <- c(
  "54703bl38_noPNI",
  "54703bl28_highPNI"
)

label_map <- c(
  "54703bl38_noPNI" = "P1-Adjacent",
  "54703bl28_highPNI" = "P1-Diseased"
)

# Keep only the two samples of interest
spe_sub <- spe[, as.character(colData(spe)$sample_id) %in% sample_order]
spe_nerve_sub <- spe_nerve[, as.character(colData(spe_nerve)$sample_id) %in% sample_order]

colData(spe_sub)$sample_id <- factor(
  as.character(colData(spe_sub)$sample_id),
  levels = sample_order
)

colData(spe_nerve_sub)$sample_id <- factor(
  as.character(colData(spe_nerve_sub)$sample_id),
  levels = sample_order
)

p_draxin_spatial <- ggcells(
  spe_sub,
  aes(x = array_col, y = -array_row, color = DRAXIN)
) +
  geom_point(size = 0.9, pch = 19) +
  facet_wrap(~sample_id, labeller = as_labeller(label_map)) +
  scale_color_viridis_c(option = "A") +
  theme_void() +
  labs(color = "DRAXIN") +
  theme(
    legend.position = "right",
    strip.text = element_text(size = 10, face = "bold"),
    strip.background = element_blank()
  )

make_nerve_hulls <- function(dat) {
  dat %>%
    group_by(sample_id, NerveID, Invaded) %>%
    filter(n() >= 3) %>%
    slice(chull(x, y)) %>%
    ungroup() %>%
    mutate(group_id = interaction(sample_id, NerveID, drop = TRUE))
}

nerve_all <- as.data.frame(colData(spe_nerve_sub)) %>%
  transmute(
    x = as.numeric(array_col),
    y = -as.numeric(array_row),
    sample_id = factor(as.character(sample_id), levels = sample_order),
    NerveID = as.character(NerveID),
    Invaded = as.character(Invaded)
  ) %>%
  filter(is.finite(x), is.finite(y), !is.na(sample_id), !is.na(NerveID), !is.na(Invaded))

hulls_all <- make_nerve_hulls(nerve_all)
hulls_inv <- hulls_all %>% filter(Invaded == "Invaded")
hulls_non <- hulls_all %>% filter(Invaded == "NonInvaded")

p_draxin_spatial_hulls2 <- p_draxin_spatial +
  geom_polygon(
    data = hulls_non,
    inherit.aes = FALSE,
    aes(x = x, y = y, group = group_id),
    fill = NA, linewidth = 0, color = "white"
  ) +
  geom_polygon(
    data = hulls_inv,
    inherit.aes = FALSE,
    aes(x = x, y = y, group = group_id),
    fill = NA, linewidth = 0, color = "white"
  ) +
  geom_polygon(
    data = hulls_non,
    inherit.aes = FALSE,
    aes(x = x, y = y, group = group_id, linetype = "Non-invaded"),
    fill = NA, linewidth = 0.6, color = "skyblue2"
  ) +
  geom_polygon(
    data = hulls_inv,
    inherit.aes = FALSE,
    aes(x = x, y = y, group = group_id, linetype = "Invaded"),
    fill = NA, linewidth = 0.6, color = "khaki2"
  ) +
  scale_linetype_manual(
    name = "Nerve border",
    values = c("Invaded" = "solid", "Non-invaded" = "solid")
  ) +
  guides(linetype = guide_legend(override.aes = list(linewidth = 1.2)))

p_draxin_spatial_hulls2
```

```{r}
#=================================
#5b. Save DRAXIN spatial plot with nerve outlines
#=================================
# This section exports the spatial DRAXIN plot with nerve hull overlays.
# Output: ../figures/Draxin_signal.png

ggsave("../figures/Draxin_signal_patient1.png", p_draxin_spatial_hulls2,
       width = 6, height = 4, dpi = 600, bg = "white")
```

```{r}
#=================================
#6. Plot adipocyte-associated signal across tissue blocks with nerve outlines
#=================================
# This section visualises the adipocyte-associated NMF signal across all samples
# and overlays nerve hulls coloured by invasion status. The signal is clipped
# for display to improve contrast while preserving the overall spatial pattern.
# Output: spatial adipocyte signal plot with invaded and non-invaded nerve outlines.

adipo_df <- as.data.frame(colData(spe)) %>%
  transmute(
    x = as.numeric(array_col),
    y = -as.numeric(array_row),
    adipocytes = as.numeric(adipocytes),
    sample_id = factor(as.character(sample_id), levels = sample_order)
  ) %>%
  filter(is.finite(x), is.finite(y), is.finite(adipocytes), !is.na(sample_id))

clip_q <- c(0.01, 0.99)
q_adipo <- quantile(adipo_df$adipocytes, probs = clip_q, na.rm = TRUE)

adipo_df <- adipo_df %>%
  mutate(adipo_plot = pmin(pmax(adipocytes, q_adipo[1]), q_adipo[2]))

p_adipo_spatial_hulls <- ggplot() +
  geom_point(
    data = adipo_df,
    aes(x = x, y = y, fill = adipo_plot),
    pch = 21,
    size = 0.9,
    color = "transparent",
    stroke = 0,
    alpha = 1
  ) +
  geom_polygon(
    data = hulls_non,
    aes(x = x, y = y, group = group_id),
    fill = NA, linewidth = 0, color = "white"
  ) +
  geom_polygon(
    data = hulls_inv,
    aes(x = x, y = y, group = group_id),
    fill = NA, linewidth = 0, color = "white"
  ) +
  geom_polygon(
    data = hulls_non,
    aes(x = x, y = y, group = group_id, linetype = "Non-invaded"),
    fill = NA, linewidth = 0.2, color = "skyblue2"
  ) +
  geom_polygon(
    data = hulls_inv,
    aes(x = x, y = y, group = group_id, linetype = "Invaded"),
    fill = NA, linewidth = 0.2, color = "khaki2"
  ) +
  facet_wrap(~sample_id, labeller = as_labeller(label_map)) +
  scale_fill_viridis_c(
    option = "A",
    name = "Adipocyte signal",
    limits = c(q_adipo[1], q_adipo[2]),
    oob = scales::squish
  ) +
  scale_linetype_manual(
    name = "Nerve border",
    values = c("Invaded" = "solid", "Non-invaded" = "solid")
  ) +
  guides(linetype = guide_legend(override.aes = list(linewidth = 1.2))) +
  theme_void() +
  theme(
    legend.position = "right",
    strip.text = element_text(size = 10, face = "bold"),
    strip.background = element_blank()
  )

p_adipo_spatial_hulls
```
```{r}
#=================================
#7. Save adipocyte-associated spatial plot with nerve outlines
#=================================
# This section exports the adipocyte-associated spatial plot with nerve hull overlays.
# Output: ../figures/adipocyte_signal.png

ggsave("../figures/adipocyte_signal.png", p_adipo_spatial_hulls,
       width = 12, height = 7, dpi = 600, bg = "white")
```

```{r}
#=================================
#8. Quantify DRAXIN at the nerve level
#=================================
# This section computes mean DRAXIN expression per nerve and compares it
# across adjacent non-invaded, diseased non-invaded, and invaded nerves.
# Output: nerve-level DRAXIN violin/boxplot with patient-coloured points and NerveID labels.

draxin_all <- SummarizedExperiment::assay(spe, "logcounts")["DRAXIN", ]

draxin_df <- data.frame(
  barcode = as.character(spe$barcode),
  DRAXIN = draxin_all,
  stringsAsFactors = FALSE
)

spot_df <- nerve_meta_df %>%
  left_join(draxin_df, by = "barcode") %>%
  filter(!is.na(DRAXIN), !is.na(NerveID), NerveID != "")

nerve_level_df <- spot_df %>%
  group_by(patient_id, NerveID, nerve_category) %>%
  summarise(
    DRAXIN = mean(DRAXIN, na.rm = TRUE),
    n_spots = n(),
    .groups = "drop"
  )

label_df <- nerve_level_df %>%
  select(NerveID, nerve_category) %>%
  distinct()

pos_nerve <- position_jitter(width = 0.18, height = 0, seed = 1)

y_min <- min(nerve_level_df$DRAXIN, na.rm = TRUE)
y_max <- max(nerve_level_df$DRAXIN, na.rm = TRUE)
label_y <- y_min - 0.10 * (y_max - y_min)

p_draxin_violin_nerve <- ggplot(nerve_level_df, aes(x = nerve_category, y = DRAXIN)) +
  geom_violin(
    fill = "grey80", color = NA,
    trim = TRUE, scale = "width", alpha = 0.9
  ) +
  geom_boxplot(
    width = 0.18, outlier.shape = NA,
    fill = "grey80", color = "grey30", alpha = 0.8
  ) +
  geom_point(
    aes(color = patient_id),
    position = pos_nerve,
    size = 0.9, alpha = 0.75
  ) +
  geom_text(
    data = label_df,
    aes(x = nerve_category, y = label_y, label = NerveID),
    position = position_jitter(width = 0.18, height = 0, seed = 1),
    inherit.aes = FALSE,
    size = 2.4,
    color = "black"
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    plot.margin = margin(10, 10, 35, 10),
    legend.position = "right"
  ) +
  labs(
    x = NULL,
    y = "DRAXIN (nerve-mean expression)",
    color = "Patient"
  )

p_draxin_violin_nerve
```
```{r}
#=================================
#8. Quantify DRAXIN at the nerve level
#=================================
# This section computes mean DRAXIN expression per nerve and compares it
# across adjacent non-invaded, diseased non-invaded, and invaded nerves.
# Output: nerve-level DRAXIN violin/boxplot with patient-coloured points and NerveID labels.

base_size <- 11
panel_aspect <- 0.5



draxin_all <- SummarizedExperiment::assay(spe, "logcounts")["DRAXIN", ]

draxin_df <- data.frame(
  barcode = as.character(spe$barcode),
  DRAXIN = draxin_all,
  stringsAsFactors = FALSE
)

spot_df <- nerve_meta_df %>%
  left_join(draxin_df, by = "barcode") %>%
  filter(!is.na(DRAXIN), !is.na(NerveID), NerveID != "")

nerve_level_df <- spot_df %>%
  group_by(patient_id, NerveID, nerve_category) %>%
  summarise(
    DRAXIN = mean(DRAXIN, na.rm = TRUE),
    n_spots = n(),
    .groups = "drop"
  ) %>%
  mutate(
    patient_id = factor(patient_id, levels = c("54703", "6249", "70865")),
    nerve_category = factor(
      nerve_category,
      levels = c("NonInvaded_Adjacent", "NonInvaded_Diseased", "Invaded")
    )
  )

# For label placement: put each NerveID just below its corresponding point

p_draxin_nerve <- ggplot(
  nerve_level_df,
  aes(x = nerve_category, y = DRAXIN)
) +
  geom_violin(
    fill = "grey85",
    color = NA,
    alpha = 0.6,
    trim = FALSE
  ) +
  geom_boxplot(
    width = 0.25,
    outlier.shape = NA,
    fill = "white",
    color = "grey30",
    linewidth = 0.4
  ) +
  geom_jitter(
    aes(color = patient_id),
    width = 0.10,
    size = 0.9,
    alpha = 0.95
  ) +
  theme_classic(base_size = base_size) +
  theme(
    aspect.ratio = panel_aspect,
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4)
  ) +
  ylab("Mean DRAXIN expression per nerve") +
  xlab("Nerve category") +
  labs(color = "Patient ID")

p_draxin_nerve

```

```{r}
#=================================
#10. Save nerve-level DRAXIN violin plot
#=================================
# This section exports the nerve-level DRAXIN plot.

cat("Saving figure...\n")
ggsave(
  filename = "../figures/DRAXIN_logcounts_violing_nerve_level.pdf",
  plot = p_draxin_nerve,
  width = 4,
  height = 4,
  units = "in",
  dpi = 600
)
cat("Saving figure...\n")
```



```{r}
#=================================
#9. Quantify DRAXIN at the spot level
#=================================
# This section compares spot-level DRAXIN expression across the three nerve
# categories to visualise within-nerve and within-patient variability.
# Output: spot-level DRAXIN violin/boxplot.

spot_df$patient_id <- factor(spot_df$patient_id, levels = c("54703", "6249", "70865"))

p_draxin_violin_spot <- ggplot(spot_df, aes(x = nerve_category, y = DRAXIN)) +
  geom_violin(
    fill = "grey80",
    color = NA,
    trim = TRUE,
    scale = "width",
    alpha = 0.9
  ) +
  geom_boxplot(
    width = 0.18,
    outlier.shape = NA,
    fill = "grey80",
    color = "grey30",
    alpha = 0.8
  ) +
  geom_point(
    aes(color = patient_id),
    position = position_jitter(width = 0.18, height = 0, seed = 1),
    size = 0.7,
    alpha = 0.6
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right"
  ) +
  labs(
    x = NULL,
    y = "DRAXIN expression (logcounts; spot-level)",
    color = "Patient"
  )

p_draxin_violin_spot
```


```{r}
#=================================
#10. Save spot-level DRAXIN violin plot
#=================================
# This section exports the spot-level DRAXIN plot.
# Output: DRAXIN_logcounts_violing_spot.pdf

cat("Saving figure...\n")
ggsave(
  filename = "DRAXIN_logcounts_violing_spot.pdf",
  plot = p_draxin_violin_spot,
  width = 4,
  height = 3,
  units = "in",
  dpi = 600
)
cat("Saving figure...\n")
```


```{r}
#=================================
#11. Load major cell type and Schwann cell signatures
#=================================
# This section imports differential-expression-derived cell state signatures
# used for downstream Schwann enrichment analyses.
# Output: no figure; signature lists for later scoring.

SupTab_major <- read_xlsx("../data/coldata/SupTable2.xlsx", sheet = "S2A. major", skip = 1) %>%
  filter(p_val_adj <= 0.05, avg_log2FC >= 1)

Major_signatures <- split(SupTab_major$gene, f = SupTab_major$cluster)
Major_signatures <- Major_signatures[!names(Major_signatures) %in% "Platelet"]

remove <- c("Schwann", "Ductal")
Major_signatures2 <- Major_signatures[!names(Major_signatures) %in% remove]

SupTab_Schwann <- read_xlsx("../data/coldata/SupTable2.xlsx", sheet = "S2H. Schwann", skip = 1) %>%
  filter(p_val_adj <= 0.05, avg_log2FC >= 1)

Schwann_signatures <- split(SupTab_Schwann$gene, f = SupTab_Schwann$cluster)
```


```{r}
#=================================
#12. Define tumour-associated nerve subset
#=================================
# This section classifies nerves by category and retains tumour-associated
# nerves only, defined here as diseased non-invaded and invaded nerves.
# Output: no figure; tumour-associated nerve object for pseudobulk analysis.

spe_nerve$types <- ifelse(spe_nerve$type == "noPNI", "Adjacent", "Diseased")

colData(spe_nerve)$nerve_types <- ifelse(
  spe_nerve$Invaded == "NonInvaded",
  paste0("NonInvaded_", spe_nerve$types),
  as.character(spe_nerve$Invaded)
)

spe_nerve <- spe_nerve[, !is.na(colData(spe_nerve)$nerve_types)]

ta_nerve_spe <- spe_nerve[, colData(spe_nerve)$nerve_types %in% c("Invaded", "NonInvaded_Diseased")]

colData(ta_nerve_spe)$nerve_types <- factor(
  colData(ta_nerve_spe)$nerve_types,
  levels = c("NonInvaded_Diseased", "Invaded")
)

table(colData(ta_nerve_spe)$nerve_types)
```

```{r}
#=================================
#13. Build pseudobulk profiles for tumour-associated nerves
#=================================
# This section aggregates spot-level counts into one pseudobulk column per nerve
# and attaches matching nerve-level metadata.
# Output: pseudobulk tumour-associated nerve object.

ids <- DataFrame(id = ta_nerve_spe$NerveID)
sb_ta <- aggregateAcrossCells(ta_nerve_spe, ids)

colnames(sb_ta) <- colData(sb_ta)$id
sb_ta <- sb_ta[, sb_ta$ncells > 1]

sb_ta <- computeLibraryFactors(sb_ta)
sb_ta <- logNormCounts(sb_ta)

ta_meta <- as.data.frame(colData(ta_nerve_spe)) %>%
  dplyr::select(NerveID, sample_id, patient_id, nerve_types) %>%
  distinct()

ta_meta <- ta_meta[!duplicated(ta_meta$NerveID), ]
rownames(ta_meta) <- ta_meta$NerveID
ta_meta <- ta_meta[colnames(sb_ta), , drop = FALSE]

colData(sb_ta)$sample_id <- ta_meta$sample_id
colData(sb_ta)$patient_id <- ta_meta$patient_id
colData(sb_ta)$nerve_types <- ta_meta$nerve_types
```

```{r}
#=================================
#14. Stratify tumour-associated nerves by DRAXIN expression
#=================================
# This section groups tumour-associated nerves into Low, Normal, and High DRAXIN
# classes based on pseudobulk DRAXIN expression.
# Output: no figure; DRAXIN group labels stored in pseudobulk metadata.

draxin_expression <- logcounts(sb_ta)["DRAXIN", ]

mu_drx <- mean(draxin_expression, na.rm = TRUE)
sd_drx <- sd(draxin_expression, na.rm = TRUE)

drx_group <- ifelse(
  draxin_expression < (mu_drx - sd_drx), "Low",
  ifelse(draxin_expression > (mu_drx + sd_drx), "High", "Normal")
)

drx_group <- factor(drx_group, levels = c("Low", "Normal", "High"))
colData(sb_ta)$DRAXIN_lev <- drx_group

table(colData(sb_ta)$DRAXIN_lev)
table(colData(sb_ta)$DRAXIN_lev, colData(sb_ta)$nerve_types)
```


```{r}
#=================================
#15. Score Schwann signatures in tumour-associated nerves
#=================================
# This section calculates one score per Schwann signature and pseudobulk nerve.
# Output: signature score matrix for tumour-associated nerves.

X_ta <- logcounts(sb_ta)

S_ta <- sapply(Schwann_signatures, function(gs) {
  score_signature_mat(gs, X_ta)
})

S_ta <- as.matrix(S_ta)
rownames(S_ta) <- colnames(sb_ta)
colnames(S_ta) <- names(Schwann_signatures)
```

```{r}
#=================================
#16. Compute Schwann signature enrichment across DRAXIN groups
#=================================
# This section converts Schwann signature scores into z-enrichment values across
# Low, Normal, and High DRAXIN tumour-associated nerve groups.
# Output: enrichment z-score matrix for heatmap plotting.

enrichZ_group <- function(v, grp) {
  mu_grp <- tapply(v, grp, mean, na.rm = TRUE)
  mu_all <- mean(v, na.rm = TRUE)
  sd_all <- sd(v, na.rm = TRUE)
  (mu_grp - mu_all) / sd_all
}

z_mat_ta <- sapply(colnames(S_ta), function(sig) {
  enrichZ_group(S_ta[, sig], colData(sb_ta)$DRAXIN_lev)
})

z_mat_ta <- t(z_mat_ta)
z_mat_ta <- z_mat_ta[, c("Low", "Normal", "High"), drop = FALSE]
```

```{r}
#=================================
#17. Plot heatmap of Schwann programs across DRAXIN groups
#=================================
# This section visualises Schwann signature enrichment across tumour-associated
# nerves stratified by DRAXIN level.
# Output: ../figures/heatmap_Schwann_by_DRAXIN_tumour_associated_nerves.pdf

SC_DRAXIN <- pheatmap(
  z_mat_ta,
  color = colorRampPalette(c(c25[1], "white", c25[2]))(100),
  border_color = NA,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  angle_col = 0,
  filename = "../figures/heatmap_Schwann_by_DRAXIN_tumour_associated_nerves.pdf",
  width = 4,
  height = 8
)

SC_DRAXIN
```


```{r}
#=================================
#18. Score Schwann signatures across all nerve categories
#=================================
# This section scores Schwann signatures across all nerve spots and computes
# category-level enrichment for adjacent non-invaded, diseased non-invaded,
# and invaded nerves.
# Output: enrichment matrix across the three nerve categories.

X_all <- if ("logcounts" %in% assayNames(spe_nerve)) {
  assay(spe_nerve, "logcounts")
} else {
  assay(spe_nerve, "counts")
}

bcs <- as.character(colData(spe_nerve)$barcode)
keep <- intersect(bcs, nerve_meta_df$barcode)

X_all <- X_all[, match(keep, bcs), drop = FALSE]
meta_all <- nerve_meta_df[match(keep, nerve_meta_df$barcode), ]

S_all <- sapply(Schwann_signatures, function(gs) {
  score_signature_mat(gs, X_all)
})

rownames(S_all) <- meta_all$barcode

enrichZ_category <- function(v) {
  mu_in <- tapply(v, meta_all$nerve_category, mean, na.rm = TRUE)
  mu_all <- mean(v, na.rm = TRUE)
  sd_all <- sd(v, na.rm = TRUE)
  (mu_in - mu_all) / sd_all
}

z_mat_all <- sapply(colnames(S_all), function(sig) enrichZ_category(S_all[, sig]))
z_mat_all <- t(z_mat_all) |> t()

rownames(z_mat_all) <- dplyr::recode(
  rownames(z_mat_all),
  "NonInvaded_Adjacent" = "NonInvaded\nAdjacent",
  "NonInvaded_Diseased" = "NonInvaded\nDiseased"
)
```

```{r}
#=================================
#19. Plot heatmap of Schwann program enrichment across nerve categories
#=================================
# This section visualises Schwann signature enrichment across the three major
# nerve categories.
# Output: ../figures/heatmap_Schwann_cat3vs2.pdf

pheatmap(
  z_mat_all,
  color = colorRampPalette(c(c25[1], "white", c25[2]))(100),
  border_color = NA,
  angle_col = 45,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  filename = "../figures/heatmap_Schwann_cat3vs2.pdf",
  width = 10,
  height = 4
)
```
