#========================================
# Gene Expression Profiles Across Nerve Categories
#========================================
#
# This script analyses gene expression across three nerve states:
# NonInvaded_Adjacent, NonInvaded_Diseased, and Invaded nerves
# in a spatial transcriptomic dataset.
#
# Gene expression is averaged per patient and nerve category,
# then aggregated across patients to obtain robust category-level
# profiles. Values are z-scored per gene to highlight relative
# differences between nerve states.
#
# Three gene panels are evaluated:
# 1. Curated immune and stromal-response genes
# 2. Invasion-associated genes from differential analysis
# 3. Tumour/NMF-correlated genes
#
# The script generates both heatmaps (category-level trends)
# and violin plots (spot-level distributions).
#
# Output:
# - Figure2E_heatmap.pdf
# - Violin_gene_plots_figure2.pdf
# - Heatmap_genes_3_categories_figure2.pdf
# - Figure3_invasion_gene_heatmap.pdf
# - Figure3_tumorNMF_correlated_gene_heatmap.pdf

Here, we want to see the average gene expression across the different categories.
```{r}
#========================================
# 1. Load libraries and input data
#========================================
#
# This section loads all required R packages, defines the color palette,
# imports the integrated SpatialExperiment object, and replaces its column
# metadata with the annotated table used for downstream analysis.
#
library(ggplot2)
library(patchwork)
library(rlang)
library(SpatialExperiment)
library(SummarizedExperiment)
library(scuttle)
library(scran)
library(scater)
library(dplyr)
library(tidyr)
library(ggsc)

c25 <- c(
  "dodgerblue2", "#E31A1C", # red
"green4",
"#6A3D9A", # purple
"#FF7F00", # orange
"black", "gold1",
"skyblue2", "#FB9A99", # lt pink
"palegreen2",
"#CAB2D6", # lt purple
"#FDBF6F", # lt orange
"gray70", "khaki2",
"maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
"darkturquoise", "green1", "yellow4", "yellow3",
"darkorange4", "brown"
)

spe <- readRDS("../data/SPE/Integrated_NMF.rds")
colData(spe) <- DataFrame(read.csv("../data/coldata/DE_SPE_annotaion.csv"))
colnames(spe) <- spe$barcode

```


```{r}
#========================================
# 2. Subset nerve spots and define nerve categories
#========================================
#
# This section subsets the dataset to nerve spots only and defines the
# comparison groups used throughout the analysis. Non-invaded nerves are
# relabelled as either Adjacent or Diseased depending on tissue context,
# while invaded nerves remain as Invaded.
#

nerve_spe <- spe[, colData(spe)$is.nerve == TRUE]
nerve_spe$types <- nerve_spe$type
colData(nerve_spe)$types[!colData(nerve_spe)$types == "noPNI"] <- "Diseased"
colData(nerve_spe)$types[colData(nerve_spe)$types == "noPNI"] <- "Adjacent"

colData(nerve_spe)$nerve_types <- nerve_spe$Invaded
colData(nerve_spe)$nerve_types <- ifelse(
  colData(nerve_spe)$Invaded == "NonInvaded",
  paste0(colData(nerve_spe)$Invaded, "_", colData(nerve_spe)$types),
  colData(nerve_spe)$nerve_types
)
nerve_spe <- nerve_spe[, !is.na(colData(nerve_spe)$nerve_types)]
```

```{r}
#========================================
# 3. Define gene sets and prepare averaged expression matrix
#========================================
#
# This section defines biologically selected gene sets of interest (selected from GSEA results, genes listed as driving enrichment), keeps # # only genes present in the dataset, extracts the log-normalized expression matrix,
# and computes mean expression first per patient and nerve category, and then
# across patients for each nerve category.
#
# Expression values are subsequently z-scored per gene to highlight relative
# enrichment patterns across categories.

keep_genes <- function(x) intersect(x, rownames(nerve_spe))

gene_sets <- list(
  "Endothelial_activation_transport" = keep_genes(c("VCAM1","SELP","JAM2","ACKR1","PLVAP","CLDN5")),
  "Lymphoid_chemokine_axis" = keep_genes(c("CCL21","CXCL12","CCL19","SELL", "CCL5")),
  "Tcells/NKcells" = keep_genes(c("CD8A","TRAC","ZAP70","IL7R","KLRK1")),
  "Plasma_humoral" = keep_genes(c("XBP1","JCHAIN","BCL6","POU2F2","IKZF1","CD52")),
  "Myeloid_macrophage" = keep_genes(c("CD14","C5AR1","CSF1R","CD163","HMOX1","APOE","C1QA","C1QC","FOLR2","MRC1","STAB1")),
  "Complement" = keep_genes(c("C3","CFD","CR1","C1R","C1S","SERPING1","C7")),
  "IFN_inflammation" = keep_genes(c("IFIT2","IFITM3","DHX58","IFI16","BATF2")),
  "Immunoregulation" = keep_genes(c("THBS1","FGL2","FCGR2B","IL10RA","VSIR","CD200")),
  "IL6_STAT3_axis" = keep_genes(c("IL6","OSMR","SOCS3","NAMPT","IL1R1")),
  "Prostaglandin" = keep_genes(c("PTGER4","PTGDS","PTGER3", "P2RY6"))
 # "Stromal remodeling" = keep_genes(c("ADAMTS9","ADAMTS4","FBN1", "DCN", "TNXB", "EFEMP1", "FSTL1"))
)

gene_map <- stack(gene_sets)
colnames(gene_map) <- c("gene","category")
gene_map$category <- factor(gene_map$category, levels = names(gene_sets))

assay_name <- "logcounts"
expr <- as.matrix(assay(nerve_spe, assay_name))  # genes x spots

meta <- as.data.frame(colData(nerve_spe))[, c("patient_id","nerve_types")]

genes <- unique(gene_map$gene)
expr <- expr[genes, , drop=FALSE]

# ---- force nerve_types order ----
# ---- force nerve_types order (match your constructed labels) ----
nerve_order <- c("NonInvaded_Adjacent", "NonInvaded_Diseased", "Invaded")
meta$nerve_types <- factor(meta$nerve_types, levels = nerve_order)

# (optional) sanity check:
# table(meta$nerve_types, useNA = "ifany")


# mean per patient x nerve_type
grp <- paste(meta$patient_id, meta$nerve_types, sep="___")
levs <- unique(grp)

expr_patient_type <- sapply(levs, function(g) {
  cols <- which(grp == g)
  Matrix::rowMeans(expr[, cols, drop=FALSE])
})
colnames(expr_patient_type) <- levs

# mean across patients (collapse to nerve_types)
# mean across patients (collapse to nerve_types) in forced order
nerve_types <- sub("^.*___", "", colnames(expr_patient_type))

nerve_order_present <- nerve_order[nerve_order %in% nerve_types]

expr_type <- sapply(nerve_order_present, function(nt) {
  Matrix::rowMeans(expr_patient_type[, nerve_types == nt, drop=FALSE])
})
expr_type <- as.matrix(expr_type)  # genes x nerve_types

# z-score per gene (like center+scale)
expr_z <- t(scale(t(expr_type)))
expr_z[is.na(expr_z)] <- 0

df <- as.data.frame(expr_z) %>%
  tibble::rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to="nerve_type", values_to="z") %>%
  left_join(gene_map, by="gene") %>%
  filter(!is.na(category))

# order genes within category (optional): keep your gene_sets order
gene_order <- unlist(gene_sets, use.names = FALSE)
gene_order <- gene_order[gene_order %in% unique(df$gene)]
df$gene <- factor(df$gene, levels = rev(gene_order))  # rev so top-to-bottom reads nicely
df$nerve_type <- factor(df$nerve_type, levels = nerve_order_present)
```

```{r}
#========================================
# 4. Plot scaled heatmap of selected genes
#========================================
#
# This section generates a heatmap of scaled average gene expression across
# nerve categories. Genes are grouped by biological category, and expression
# is shown as z-scores to emphasize relative up- and downregulation.

p <- ggplot(df, aes(x = nerve_type, y = gene, fill = z)) +
  geom_tile() +
  facet_grid(rows = vars(category), scales = "free_y", space = "free_y") +
  scale_fill_distiller(
  palette = "RdBu",
  direction = -1
) +
theme_bw() +
  theme(
    panel.spacing.y = unit(0.6, "lines"),
    strip.text.y = element_text(angle = 0),
    axis.text.y = element_text(size = 7)
  ) +
  labs(x = NULL, y = NULL)

p
```

```{r}
#========================================
# 5. Save category-level heatmap
#========================================
#
# This section saves the scaled heatmap of selected genes across nerve
# categories as a publication-quality PDF.
#

 ggsave(
  filename = "Figure2E_heatmap.pdf",
  plot = p,
  width = 5,
  height = 10,
  units = "in",
  dpi = 600
)
```

```{r}
#========================================
# 6. Define gene groups and generate violin plots
#========================================
#
# This section defines all curated gene groups of interest, generates violin
# plots for each group using a reusable helper function, combines all plots
# into a single figure, and reuses the same gene list for the grouped heatmap.


# Define all gene groups in one place
gene_groups <- list(
  "C3 Fibroblasts" = c("C3", "CFH", "SPARCL1", "CXCL12", "APOD", "APOE"),
  "Lymphocyte Recruitment" = c("VCAM1", "SELL", "JAM2", "CCL21"),
  "Agprocessing" = c("FCGR2B", "THBS1", "FGL2"),
  "B cell" = c("BCL6", "XBP1", "CR1"),
  "Complement" = c("C1QA", "CFD", "C1QC", "RGCC"),
  "PGE2" = c("SFRP1", "PTGER4", "P2RY6"),
  "IFNgamma" = c("STAT4", "IFIT2", "IFITM3", "SOCS3"),
  "IL6" = c("IL6", "OSMR", "HMOX1", "CD36"),
  "Inflammatory" = c("CX3CL1", "C5AR1", "IL1R1", "IL10RA", "IL18R1", "CCL2", "CD14"),
  "ERK1_ERK2" = c("CSF1R", "BRAF", "DNAJC27", "NRP1")
)

# Set grouping labels for plotting
colLabels(nerve_spe) <- colData(nerve_spe)$nerve_types

# Helper function to generate one violin plot panel
make_violin_plot <- function(spe_obj, genes, plot_title, slot_name = "logcounts", ncol = 3) {
  
  # keep only genes present in the object
  genes <- intersect(genes, rownames(spe_obj))
  
  p <- sc_violin(
    spe_obj,
    features = genes,
    slot = slot_name,
    ncol = ncol
  ) +
    ggplot2::theme_bw() +
    labs(title = plot_title) +
    xlab("Category") +
    ylab("Gene expression")
  
  facet_var <- as_name(p$facet$params$facets[[1]])
  p + facet_wrap(vars(!!sym(facet_var)), scales = "free_y")
}

# Generate all violin plots automatically
violin_plots <- lapply(names(gene_groups), function(group_name) {
  make_violin_plot(
    spe_obj = nerve_spe,
    genes = gene_groups[[group_name]],
    plot_title = group_name,
    ncol = if (group_name == "C3 Fibroblasts") 2 else 3
  )
})

names(violin_plots) <- names(gene_groups)

# Combine plots into one figure
combined_plots <- wrap_plots(violin_plots, ncol = 1)

# Save combined violin plot figure
ggsave(
  filename = "../figures/Violin_gene_plots_figure2.pdf",
  plot = combined_plots,
  width = 16,
  height = 30,
  units = "in"
)

combined_plots
```

```{r}
#========================================
# 7. Reuse the same gene groups for grouped heatmap
#========================================
#
# This section combines all genes from the curated gene groups and visualizes
# them in a grouped heatmap across nerve categories, with patient-level blocking.
#

Genes_of_interest <- unique(unlist(gene_groups))

Genes_of_interest <- Genes_of_interest[Genes_of_interest %in% rownames(nerve_spe)]

p11 <- plotGroupedHeatmap(
  nerve_spe,
  group = "nerve_types",
  features = Genes_of_interest,
  block = "patient_id",
  center = TRUE,
  scale = TRUE,
  fontsize = 14
)
```

```{r}
#========================================
# 8. Save category-level heatmap
#========================================
#
# This section saves the scaled heatmap of selected genes across nerve
# categories as a publication-quality PDF.
#

ggsave(
  filename = "../figures/Heatmap_genes_3_categories_figure2.pdf",
  plot = p11,
  width = 8,
  height = 16
)

p11
```

```{r}
#========================================
# 9. Plot selected invasion-associated genes across nerve categories
#========================================
#
# This section visualises a second curated gene panel derived from the
# 342 genes identified across the two invasion-related contrasts.
# These genes were selected to represent invasion-associated programs such as
# nerve injury, synaptic-like signalling, extracellular matrix remodeling,
# vasculature association, axon guidance, myelin debris clearance, and
# Schwann cell niche signalling.
#
# As above, expression is averaged first within each patient and nerve
# category, then across patients for each nerve category, and finally
# z-scored per gene to highlight relative enrichment patterns.

keep_genes <- function(x) intersect(x, rownames(nerve_spe))

gene_sets_invasion <- list(
  "Nerve_injury" = keep_genes(c("GAP43", "GLI3", "GFRA2", "NRCAM", "NES", "SOX2")),

  "Synaptic/Glutamatergic" = keep_genes(c("SYT11", "DLGAP1", "SYNGAP1", "SYNPO", "CNIH3")),

  "Extracellular_Matrix" = keep_genes(c("ITGAV", "ITGA1", "DAG1", "MMP17", "MMP28", "PLOD3")),

  "Vasculature_associated" = keep_genes(c("ESM1", "EDIL3", "VEZF1", "ENTPD1", "THSD7B")),

  "Axon_guidance" = keep_genes(c("NAV2", "EPHB2", "UNC5C", "PLXNC1", "DRAXIN", "SEMA3C")),

  "Myelin_debris_clearance" = keep_genes(c("GAL3ST1", "SCD", "LPIN1", "AXL", "FABP5", "CLCN7")),

  "SC_niche_signaling" = keep_genes(c("CLCF1", "CRLF1", "IL11", "CCN3", "FGF5", "BTC"))
)
```

```{r}

#========================================
# 10. Prepare averaged expression matrix for invasion-associated genes
#========================================
#
# This section reshapes the selected invasion-associated genes into a grouped
# expression matrix, computes mean expression per patient and nerve category,
# then collapses these values across patients to obtain category-level averages.
# Expression values are then z-scored per gene for heatmap visualisation.

gene_map_invasion <- stack(gene_sets_invasion)
colnames(gene_map_invasion) <- c("gene","category")
gene_map_invasion$category <- factor(gene_map_invasion$category, levels = names(gene_sets_invasion))

assay_name <- "logcounts"
expr <- as.matrix(assay(nerve_spe, assay_name))

meta <- as.data.frame(colData(nerve_spe))[, c("patient_id","nerve_types")]

genes <- unique(gene_map_invasion$gene)
expr <- expr[genes, , drop = FALSE]

grp <- paste(meta$patient_id, meta$nerve_types, sep = "___")
levs <- unique(grp)

expr_patient_type <- sapply(levs, function(g) {
  cols <- which(grp == g)
  Matrix::rowMeans(expr[, cols, drop = FALSE])
})
colnames(expr_patient_type) <- levs

nerve_types <- sub("^.*___", "", colnames(expr_patient_type))
nerve_order <- c("NonInvaded_Adjacent", "NonInvaded_Diseased", "Invaded")
nerve_order_present <- nerve_order[nerve_order %in% nerve_types]

expr_type <- sapply(nerve_order_present, function(nt) {
  Matrix::rowMeans(expr_patient_type[, nerve_types == nt, drop = FALSE])
})
expr_type <- as.matrix(expr_type)

expr_z <- t(scale(t(expr_type)))
expr_z[is.na(expr_z)] <- 0

df_invasion <- as.data.frame(expr_z) %>%
  tibble::rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to = "nerve_type", values_to = "z") %>%
  left_join(gene_map_invasion, by = "gene") %>%
  filter(!is.na(category))

gene_order_invasion <- unlist(gene_sets_invasion, use.names = FALSE)
gene_order_invasion <- gene_order_invasion[gene_order_invasion %in% unique(df_invasion$gene)]

df_invasion$gene <- factor(df_invasion$gene, levels = rev(gene_order_invasion))
df_invasion$nerve_type <- factor(df_invasion$nerve_type, levels = nerve_order_present)
```

```{r}
#========================================
# 11. Plot heatmap of invasion-associated genes
#========================================
#
# This section generates a heatmap of scaled average expression for the
# selected invasion-associated genes across the three nerve categories.
# Genes are grouped by biological program, and values are displayed as
# z-scores to emphasise relative enrichment between categories.

p_invasion <- ggplot(df_invasion, aes(x = nerve_type, y = gene, fill = z)) +
  geom_tile() +
  facet_grid(rows = vars(category), scales = "free_y", space = "free_y") +
  scale_fill_distiller(
    palette = "RdBu",
    direction = -1
  ) +
  theme_bw() +
  theme(
    panel.spacing.y = unit(0.6, "lines"),
    strip.text.y = element_text(angle = 0),
    axis.text.y = element_text(size = 7)
  ) +
  labs(x = NULL, y = NULL)

p_invasion
```

```{r}
#========================================
# 12. Save invasion-associated gene heatmap
#========================================
#
# This section saves the heatmap of scaled average expression for the selected
# invasion-associated genes as a publication-quality PDF.


ggsave(
  filename = "../figures/Figure3_invasion_gene_heatmap.pdf",
  plot = p_invasion,
  width = 5,
  height = 10,
  units = "in",
  dpi = 600
)
```

```{r}
#========================================
# 13. Define tumour/NMF-correlated gene set
#========================================
#
# This section defines a focused set of genes associated with
# tumour-derived NMF programs.
#

keep_genes2 <- function(x) intersect(x, rownames(nerve_spe))

gene_sets2 <- list(
  "Tumor_NMF_correlated" = keep_genes2(c(
    "PDGFA","PDGFB","AXIN2","SEMA3C","EPHB2","EXOC8","WHAMM","CDC42BPA","SQSTM1","WIPI2","CHST11"
  ))
)

```

```{r}
#========================================
# 14. Prepare averaged expression matrix (NMF genes)
#========================================
#
# This section computes mean expression per patient and nerve
# category, then aggregates across patients and z-scores per gene.
#

gene_map2 <- stack(gene_sets2)
colnames(gene_map2) <- c("gene","category")
gene_map2$category <- factor(gene_map2$category, levels = names(gene_sets2))

assay_name2 <- "logcounts"
expr2 <- as.matrix(assay(nerve_spe, assay_name2))  # genes x spots

meta2 <- as.data.frame(colData(nerve_spe))[, c("patient_id","nerve_types")]
stopifnot(all(c("patient_id","nerve_types") %in% colnames(meta2)))

## subset to the genes you want
genes2 <- unique(gene_map2$gene)
expr2 <- expr2[genes2, , drop = FALSE]

## group spots by patient x nerve_type
grp2  <- paste(meta2$patient_id, meta2$nerve_types, sep="___")
levs2 <- unique(grp2)

## mean per (patient x nerve_type)
expr_patient_type2 <- sapply(levs2, function(g) {
  cols <- which(grp2 == g)
  Matrix::rowMeans(expr2[, cols, drop = FALSE])
})
expr_patient_type2 <- as.matrix(expr_patient_type2)
colnames(expr_patient_type2) <- levs2

## collapse to nerve_types by averaging across patients
nerve_types2 <- sub("^.*___", "", colnames(expr_patient_type2))
nt_levels2   <- unique(meta2$nerve_types)  # preserves your original ordering in metadata

expr_type2 <- sapply(nt_levels2, function(nt) {
  cols <- which(nerve_types2 == nt)
  if (length(cols) == 0) return(rep(NA_real_, nrow(expr_patient_type2)))
  Matrix::rowMeans(expr_patient_type2[, cols, drop = FALSE])
})
expr_type2 <- as.matrix(expr_type2)
rownames(expr_type2) <- rownames(expr_patient_type2)
colnames(expr_type2) <- nt_levels2

## z-score per gene
expr_z2 <- t(scale(t(expr_type2)))
expr_z2[is.na(expr_z2)] <- 0

## build plotting dataframe (IMPORTANT: join with gene_map2, not gene_map)
df2 <- as.data.frame(expr_z2) |>
  tibble::rownames_to_column("gene") |>
  tidyr::pivot_longer(-gene, names_to="nerve_type", values_to="z") |>
  dplyr::left_join(gene_map2, by="gene") |>
  dplyr::filter(!is.na(category))

## order genes within category following gene_sets2
gene_order2 <- unlist(gene_sets2, use.names = FALSE)
gene_order2 <- gene_order2[gene_order2 %in% unique(df2$gene)]
df2$gene <- factor(df2$gene, levels = rev(gene_order2))
df2$nerve_type <- factor(df2$nerve_type, levels = colnames(expr_z2))
df2$category <- factor(df2$category, levels = names(gene_sets2))

```

```{r}
#========================================
# 15. Plot heatmap of tumour/NMF-correlated genes
#========================================
#
# This section visualises scaled expression of NMF-correlated genes
# across nerve categories.
#

class(nerve_spe)
p2 <- ggplot(df2, aes(x = nerve_type, y = gene, fill = z)) +
  geom_tile() +
  facet_grid(rows = vars(category), scales = "free_y", space = "free_y") +
  scale_fill_distiller(palette = "RdBu", direction = -1) +
  theme_bw() +
  theme(
    panel.spacing.y = grid::unit(0.6, "lines"),
    strip.text.y = element_text(angle = 0),
    axis.text.y = element_text(size = 7)
  ) +
  labs(x = NULL, y = NULL)

p2
```

```{r}
#========================================
# 16. Save tumour/NMF-correlated gene heatmap
#========================================
#
# This section exports the heatmap as a publication-quality PDF.
#


ggsave(
  filename = "../figures/Figure3_tumorNMF_correlated_gene_heatmap.pdf",
  plot = p2,
  width = 5,
  height = 10,
  units = "in",
  dpi = 600
)
```
