#========================================
# Gene expression profiles across nerve categories
#========================================
#
# This script analyses DRAXIN expression and associated transcriptional programs
# across nerve categories in the spatial transcriptomic dataset. It builds a
# unified nerve-level metadata table, aggregates spot-level measurements per nerve,
# and generates correlation plots relating DRAXIN to NMF-derived programs and a
# curated TGFβ-response signature.
#
# Output figures generated by this script:
# - Figure4J_DRAXIN_vs_NMF_nerve_tumor_perNerve.png
#   Correlation of mean DRAXIN expression per nerve with mean nerve and tumour
#   NMF program scores across all nerve categories.
#
# - Figure4J_invaded_DRAXIN_vs_NMFnerve_mnTumor_perNerve.pdf
#   Correlation of mean DRAXIN expression per invaded nerve with mean nerve and
#   tumour NMF program scores, coloured by patient.
#
# - Figure4J_DRAXIN_vs_TGFB_perTumourAssociatedNerve.pdf
#   Correlation of mean DRAXIN expression per tumour-associated nerve with a
#   mean TGFβ-response score derived from curated genes.


```{r}
#========================================
# 1. Load libraries and input data
#========================================
# Goal: Load all required packages and import the SpatialExperiment object and
# associated metadata only once to avoid duplicated setup.

library(ggplot2)
library(patchwork)
library(rlang)
library(SpatialExperiment)
library(SummarizedExperiment)
library(scuttle)
library(scran)
library(scater)
library(dplyr)
library(tidyr)
library(ggsc)
library(Matrix)
library(purrr)
library(lme4)
library(stringr)
library(msigdbr)

c25 <- c(
  "dodgerblue2", "#E31A1C", "green4", "#6A3D9A", "#FF7F00",
  "black", "gold1", "skyblue2", "#FB9A99", "palegreen2",
  "#CAB2D6", "#FDBF6F", "gray70", "khaki2", "maroon",
  "orchid1", "deeppink1", "blue1", "steelblue4",
  "darkturquoise", "green1", "yellow4", "yellow3",
  "darkorange4", "brown"
)

spe <- readRDS("../data/SPE/Integrated_NMF.rds")
colData(spe) <- DataFrame(read.csv("../data/coldata/DE_SPE_annotaion.csv"))
colnames(spe) <- spe$barcode

bg <- data.frame(
  colData(spe)[, c("array_col", "array_row", "tumor", "sample_id",
                   "BorderInvaded", "BorderID", "tumorProximalCells")]
)

W_df <- readRDS("../data/Deconvolution/RCTD_weights.Rds")
```

```{r}
#========================================
# 2. Define helper functions
#========================================
# Goal: Centralise repeated logic for column detection, nerve annotation, and
# safe gene filtering so the same objects are not rebuilt multiple times.

pick_col <- function(df, candidates) {
  hit <- candidates[candidates %in% colnames(df)]
  if (length(hit) > 0) hit[1] else NA_character_
}

keep_genes <- function(x, expr_mat) {
  intersect(x, rownames(expr_mat))
}

build_nerve_spe <- function(spe_object) {
  nerve_spe <- spe_object[, colData(spe_object)$is.nerve == TRUE]

  nerve_spe$types <- nerve_spe$type
  colData(nerve_spe)$types[colData(nerve_spe)$types == "noPNI"] <- "Adjacent"
  colData(nerve_spe)$types[colData(nerve_spe)$types != "Adjacent"] <- "Diseased"

  colData(nerve_spe)$nerve_types <- ifelse(
    colData(nerve_spe)$Invaded == "NonInvaded",
    paste0(colData(nerve_spe)$Invaded, "_", colData(nerve_spe)$types),
    as.character(colData(nerve_spe)$Invaded)
  )

  nerve_spe[, !is.na(colData(nerve_spe)$nerve_types)]
}

build_nerve_meta <- function(nerve_spe, assay_name = "logcounts") {
  expr <- as.matrix(assay(nerve_spe, assay_name))
  meta <- as.data.frame(colData(nerve_spe))

  stopifnot("DRAXIN" %in% rownames(expr))
  meta$DRAXIN_expr <- as.numeric(expr["DRAXIN", ])

  list(expr = expr, meta = meta)
}
```

```{r}
#========================================
# 3. Build nerve-level input objects
#========================================
# Goal: Create the nerve-only SpatialExperiment and extract expression and
# metadata once so all downstream analyses use the same starting objects.

nerve_spe <- build_nerve_spe(spe)
nerve_data <- build_nerve_meta(nerve_spe, assay_name = "logcounts")

expr <- nerve_data$expr
meta <- nerve_data$meta

nmf_nerve_col <- "nerve"
nmf_tumor_col <- "mnTumor"
```

```{r}
#========================================
# 4. Aggregate all nerves for DRAXIN and NMF program analysis
#========================================
# Goal: Compute one nerve-level profile per patient/sample/nerve by averaging
# DRAXIN and NMF scores across spots belonging to the same nerve.

nerve_df <- meta %>%
  group_by(patient_id, sample_id, NerveID, nerve_types) %>%
  summarise(
    DRAXIN_mean    = mean(DRAXIN_expr, na.rm = TRUE),
    NMF_nerve_mean = mean(.data[[nmf_nerve_col]], na.rm = TRUE),
    NMF_tumor_mean = mean(.data[[nmf_tumor_col]], na.rm = TRUE),
    n_spots        = dplyr::n(),
    .groups = "drop"
  ) %>%
  filter(n_spots >= 3) %>%
  mutate(
    nerve_types = factor(
      nerve_types,
      levels = c("Invaded", "NonInvaded_Diseased", "NonInvaded_Adjacent")
    )
  )

df_long_all <- nerve_df %>%
  pivot_longer(
    cols = c(NMF_nerve_mean, NMF_tumor_mean),
    names_to = "program",
    values_to = "program_score"
  ) %>%
  mutate(
    program = recode(
      program,
      NMF_nerve_mean = "NMF nerve program",
      NMF_tumor_mean = "NMF tumour program"
    )
  )

cor_stats_all <- df_long_all %>%
  group_by(program) %>%
  summarise(
    rho = suppressWarnings(
      cor(program_score, DRAXIN_mean, method = "spearman", use = "pairwise.complete.obs")
    ),
    p = suppressWarnings(
      cor.test(program_score, DRAXIN_mean, method = "spearman", exact = FALSE)$p.value
    ),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("Spearman \u03C1 = ", sprintf("%.2f", rho), "\n", "p = ", signif(p, 2)))
```

```{r}
#========================================
# 5. Plot DRAXIN versus NMF programs across all nerves
#========================================
# Goal: Visualise the relationship between mean DRAXIN expression and mean NMF
# program scores per nerve across all nerve categories.

p4J_all <- ggplot(df_long_all, aes(x = program_score, y = DRAXIN_mean, color = nerve_types)) +
  geom_point(alpha = 0.85, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.7) +
  facet_wrap(~ program, scales = "free_x") +
  theme_bw() +
  theme(
    legend.position = "right",
    strip.text = element_text(face = "bold"),
    plot.margin = margin(10, 20, 10, 10)
  ) +
  labs(
    x = "Mean NMF program score per nerve",
    y = "Mean DRAXIN expression (logcounts) per nerve",
    color = "Nerve type"
  ) +
  geom_text(
    data = cor_stats_all,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = -0.05,
    vjust = 1.1,
    size = 3.3
  )

p4J_all
```

```{r}
ggsave(
  filename = "../figures/Figure4J_DRAXIN_vs_NMF_nerve_tumor_perNerve.png",
  plot = p4J_all,
  width = 7,
  height = 4.2,
  units = "in",
  dpi = 600
)
```

```{r}
#========================================
# 6. Restrict to invaded nerves only
#========================================
# Goal: Subset the nerve metadata to invaded nerves only and compute one mean
# DRAXIN/NMF profile per invaded nerve for patient-stratified plotting.

inv_idx <- which(meta$InvadedNerve %in% c(TRUE, 1, "TRUE"))
meta_inv <- meta[inv_idx, , drop = FALSE]

inv_nerve <- meta_inv %>%
  group_by(patient_id, sample_id, NerveID) %>%
  summarise(
    nerve_mean   = mean(.data[[nmf_nerve_col]], na.rm = TRUE),
    mnTumor_mean = mean(.data[[nmf_tumor_col]], na.rm = TRUE),
    DRAXIN_mean  = mean(DRAXIN_expr, na.rm = TRUE),
    n_spots      = dplyr::n(),
    .groups = "drop"
  ) %>%
  filter(n_spots >= 3) %>%
  mutate(patient_id = factor(patient_id, levels = c("54703", "6249", "70865")))

df_long_inv <- inv_nerve %>%
  pivot_longer(
    cols = c(nerve_mean, mnTumor_mean),
    names_to = "program",
    values_to = "program_score"
  ) %>%
  mutate(
    program = recode(
      program,
      nerve_mean = "NMF nerve program",
      mnTumor_mean = "NMF tumour program"
    )
  )

cor_stats_inv <- df_long_inv %>%
  group_by(program) %>%
  summarise(
    rho = suppressWarnings(
      cor(program_score, DRAXIN_mean, method = "spearman", use = "pairwise.complete.obs")
    ),
    p = suppressWarnings(
      cor.test(program_score, DRAXIN_mean, method = "spearman", exact = FALSE)$p.value
    ),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("Spearman \u03C1 = ", sprintf("%.2f", rho), "\n", "p = ", signif(p, 2)))
```

```{r}
#========================================
# 7. Plot DRAXIN versus NMF programs in invaded nerves only
#========================================
# Goal: Plot the DRAXIN-program relationship specifically in invaded nerves and
# colour points by patient to visualise inter-patient consistency.

p4J_inv <- ggplot(df_long_inv, aes(x = program_score, y = DRAXIN_mean, color = patient_id)) +
  geom_point(alpha = 0.85, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.7) +
  facet_wrap(~ program, scales = "free_x") +
  theme_bw() +
  theme(
    legend.position = "right",
    strip.text = element_text(face = "bold"),
    plot.margin = margin(10, 20, 10, 10)
  ) +
  labs(
    x = "Mean program score per invaded nerve",
    y = "Mean DRAXIN expression (logcounts) per invaded nerve",
    color = "Patient"
  ) +
  geom_text(
    data = cor_stats_inv,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = -0.05,
    vjust = 1.1,
    size = 3.3
  )

p4J_inv
```

```{r}
ggsave(
  "../figures/Figure4J_invaded_DRAXIN_vs_NMFnerve_mnTumor_perNerve.pdf",
  p4J_inv,
  width = 5.99,
  height = 4.2,
  units = "in",
  dpi = 600
)
```

```{r}
#========================================
# 8. Define tumour-associated nerves and MSigDB Hallmark TGFβ gene set
#========================================
# Goal: Restrict analysis to tumour-associated nerves and define an independent
# MSigDB Hallmark TGFβ-response signature for score calculation.

plot_cfg <- list(
  z_scope = "all_nerve_spots",
  point_size = 2,
  point_alpha = 0.85,
  add_lm = TRUE,
  color_map = c(
    "Invaded" = "#E31A1C",
    "NonInvaded_Diseased" = "dodgerblue2"
  ),
  cor_method = "spearman",
  cor_exact = FALSE,
  caption_wrap_width = 110,
  bottom_margin_pt = 28
)

ta_idx <- which(meta$nerve_types %in% c("Invaded", "NonInvaded_Diseased"))
meta_ta <- meta[ta_idx, , drop = FALSE]

if (plot_cfg$z_scope == "all_nerve_spots") {
  z <- t(scale(t(expr)))
  z[is.na(z)] <- 0
  z_ta <- z[, ta_idx, drop = FALSE]
} else {
  expr_ta <- expr[, ta_idx, drop = FALSE]
  z_ta <- t(scale(t(expr_ta)))
  z_ta[is.na(z_ta)] <- 0
}

# Retrieve MSigDB Hallmark TGFβ signalling gene set.
# This uses the independently curated HALLMARK_TGF_BETA_SIGNALING signature.
tgfb_hallmark_df <- msigdbr(
  species = "Homo sapiens",
  collection = "H"
)

tgfb_genes_raw <- tgfb_hallmark_df %>%
  filter(gs_name == "HALLMARK_TGF_BETA_SIGNALING") %>%
  pull(gene_symbol) %>%
  unique()

cat("\nNumber of genes in MSigDB Hallmark TGFβ signalling set:\n")
print(length(tgfb_genes_raw))

# Keep only genes present in expr.
# keep_genes() is assumed to be defined earlier in your script.
tgfb_genes <- keep_genes(tgfb_genes_raw, expr)

# Exclude DRAXIN if present, to avoid circularity in the DRAXIN correlation.
tgfb_genes <- setdiff(tgfb_genes, "DRAXIN")

cat("\nNumber of Hallmark TGFβ genes detected in expr and used for scoring:\n")
print(length(tgfb_genes))

cat("\nHallmark TGFβ genes used for scoring:\n")
print(tgfb_genes)

# Require a minimum number of detected genes for a meaningful score.
stopifnot(length(tgfb_genes) >= 5)
```

```{r}
#========================================
# 9. Compute MSigDB Hallmark TGFβ score per tumour-associated nerve
#========================================
# Goal: Calculate a spot-level mean Hallmark TGFβ z-score and then aggregate it
# per tumour-associated nerve together with DRAXIN expression.

meta_ta$TGFB_HALLMARK_score <- Matrix::colMeans(
  z_ta[tgfb_genes, , drop = FALSE],
  na.rm = TRUE
)

ta_nerve <- meta_ta %>%
  group_by(patient_id, sample_id, NerveID, nerve_types) %>%
  summarise(
    DRAXIN_mean        = mean(DRAXIN_expr, na.rm = TRUE),
    TGFB_HALLMARK_mean = mean(TGFB_HALLMARK_score, na.rm = TRUE),
    mnTumor_mean       = mean(.data[[nmf_tumor_col]], na.rm = TRUE),
    n_spots            = dplyr::n(),
    .groups = "drop"
  ) %>%
  filter(n_spots >= 3) %>%
  mutate(
    patient_id = factor(patient_id),
    nerve_types = factor(nerve_types, levels = names(plot_cfg$color_map))
  )

ct <- cor.test(
  ta_nerve$TGFB_HALLMARK_mean,
  ta_nerve$DRAXIN_mean,
  method = plot_cfg$cor_method,
  exact = plot_cfg$cor_exact
)

lab <- paste0(
  "Spearman \u03C1 = ", sprintf("%.2f", unname(ct$estimate)),
  "\n", "p = ", signif(ct$p.value, 2)
)

gene_caption <- paste0(
  "MSigDB Hallmark TGF\u03B2 signalling genes used for scoring, DRAXIN excluded if present: ",
  paste(tgfb_genes, collapse = ", ")
)

gene_caption <- stringr::str_wrap(
  gene_caption,
  width = plot_cfg$caption_wrap_width
)
```

```{r}
#========================================
# 10. Plot DRAXIN versus Hallmark TGFβ response in tumour-associated nerves
#========================================
# Goal: Plot the association between the MSigDB Hallmark TGFβ-response signature
# and mean DRAXIN expression across tumour-associated nerves.

p_tgfb <- ggplot(
  ta_nerve,
  aes(
    x = TGFB_HALLMARK_mean,
    y = DRAXIN_mean,
    color = nerve_types
  )
) +
  geom_point(
    alpha = plot_cfg$point_alpha,
    size = plot_cfg$point_size
  ) +
  {
    if (isTRUE(plot_cfg$add_lm)) {
      geom_smooth(
        method = "lm",
        se = FALSE,
        color = "black",
        linewidth = 0.7
      )
    }
  } +
  scale_color_manual(
    values = plot_cfg$color_map,
    drop = FALSE
  ) +
  theme_bw() +
  labs(
    x = "Mean MSigDB Hallmark TGF\u03B2 signalling score per tumour-associated nerve (z)",
    y = "Mean DRAXIN expression (logcounts) per nerve",
    color = "Nerve type",
    caption = gene_caption
  ) +
  annotate(
    "text",
    x = -Inf,
    y = Inf,
    label = lab,
    hjust = -0.05,
    vjust = 1.1,
    size = 3.3
  ) +
  theme(
    plot.margin = margin(6, 10, plot_cfg$bottom_margin_pt, 10),
    plot.caption = element_text(size = 8.5, hjust = 0),
    legend.position = "right"
  )

p_tgfb
```

```{r}
ggsave(
  "../figures/FigureS9D_DRAXIN_vs_MSigDB_Hallmark_TGFB_perTumourAssociatedNerve.pdf",
  p_tgfb,
  width = 6.2,
  height = 4.6,
  units = "in",
  dpi = 600
)
```
